---
title: "v_plants — Paired Resurvey Analysis"
author: "Valtýr"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: kable
    code_folding: hide
    fig_width: 9
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo        = TRUE,
  warning     = FALSE,
  message     = FALSE,
  cache       = TRUE,         # long-running chunks (NMDS, IndVal) cached to disk
  cache.path  = "cache/v_plants_",
  fig.align   = "center"
)

# Helper: print a tibble fully if ≤ 400 rows, truncated otherwise
print_full <- function(x, caption = NULL) {
  n <- nrow(x)
  if (n <= 400) {
    knitr::kable(x, digits = 4, caption = caption)
  } else {
    knitr::kable(
      head(x, 400),
      digits  = 4,
      caption = paste0(caption, sprintf("  [showing first 400 of %d rows]", n))
    )
  }
}
```

```{r libraries}
library(tidyverse)
library(vegan)
library(broom)
library(ggrepel)
library(indicspecies)
library(knitr)

set.seed(123)
```

```{r bootstrap-required-objects}
# Knit runs in a clean session. Use shared prep script as single source of truth.
prep_script <- "v_plants_prep.R"
if (!file.exists(prep_script)) {
  stop("Required script not found: ", prep_script)
}
source(prep_script, local = .GlobalEnv)
```

---

> **Purpose.**  This document picks up directly after `v_plants_variable_screening.html`.
> It assumes the following objects are already in your R environment — source
> the screening script first if needed:
>
> | Object | Description |
> |---|---|
> | `community_long_paired` | Long-format cover table (plot × sample × taxon) |
> | `community_wide_strict` | Wide community matrix with `plot_number` + `sample` columns |
> | `meta_strict` | Metadata aligned row-for-row with `community_wide_strict` |
> | `mat_strict` | Numeric species matrix (318 rows × 747 cols) |
> | `nmds_strict` | `metaMDS` result on `mat_strict` (all plots, pre-outlier removal) |
> | `fit_strict` | `envfit` result on `nmds_strict` |
> | `vars_driver_strict_final`, `driver_final` | Variable vectors from screening |

---

# Part 1 — Preliminary NMDS (all plots)

## 1.1 Ordination plot

This first pass runs on the full 167-plot dataset before any outlier removal.
It is included here as a diagnostic to show why the three coastal/dune plots
need special treatment.

```{r nmds-all-plot, fig.height=7, fig.cap="NMDS of all 167 revisited plots. Note the three outlier sites near NMDS1 = 17 that stretch the axis and compress the main community cloud."}

# Extract site scores
nmds_scores <- as.data.frame(scores(nmds_strict, display = "sites")) %>%
  bind_cols(meta_strict %>% select(plot_number, sample))

# --- envfit arrows (continuous vectors) ---
sig_thresh <- 0.05

vec_df <- tibble(variable = character(), NMDS1 = numeric(), NMDS2 = numeric(),
                 r2 = numeric(), p_value = numeric(), kind = character())

if (!is.null(fit_strict$vectors) && !is.null(fit_strict$vectors$arrows)) {
  vec_raw  <- as.data.frame(fit_strict$vectors$arrows) %>%
    rownames_to_column("variable")
  vec_axes <- setdiff(names(vec_raw), "variable")
  if (length(vec_axes) >= 2) {
    vec_df <- vec_raw %>%
      rename(NMDS1 = all_of(vec_axes[1]), NMDS2 = all_of(vec_axes[2])) %>%
      select(variable, NMDS1, NMDS2) %>%
      mutate(
        r2      = as.numeric(fit_strict$vectors$r[variable]),
        p_value = as.numeric(fit_strict$vectors$pvals[variable]),
        kind    = "numeric"
      )
  }
}

# --- envfit factor centroids ---
fac_df <- tibble(variable = character(), NMDS1 = numeric(), NMDS2 = numeric(),
                 r2 = numeric(), p_value = numeric(), kind = character())

if (!is.null(fit_strict$factors) && !is.null(fit_strict$factors$centroids)) {
  fac_raw  <- as.data.frame(fit_strict$factors$centroids) %>%
    rownames_to_column("variable")
  fac_axes <- setdiff(names(fac_raw), "variable")
  if (length(fac_axes) >= 2) {
    fac_df <- fac_raw %>%
      rename(NMDS1 = all_of(fac_axes[1]), NMDS2 = all_of(fac_axes[2])) %>%
      select(variable, NMDS1, NMDS2) %>%
      mutate(r2 = NA_real_, p_value = NA_real_, kind = "factor")
  }
}

arrow_df_all <- bind_rows(vec_df, fac_df) %>%
  filter(is.na(p_value) | p_value <= sig_thresh)

ordi_radius_all <- max(abs(c(nmds_scores$NMDS1, nmds_scores$NMDS2))) * 0.7

arrow_df_all <- arrow_df_all %>%
  mutate(xend = NMDS1 * ordi_radius_all,
         yend = NMDS2 * ordi_radius_all)

ggplot(nmds_scores, aes(x = NMDS1, y = NMDS2)) +
  geom_line(aes(group = plot_number),
            colour = "grey70", linewidth = 0.3, alpha = 0.6) +
  geom_point(aes(colour = sample, shape = sample),
             size = 2.0, alpha = 0.85) +
  geom_segment(
    data = arrow_df_all %>% filter(kind == "numeric"),
    aes(x = 0, y = 0, xend = xend, yend = yend),
    arrow = arrow(length = unit(0.2, "cm")),
    colour = "#c0392b", linewidth = 0.7, inherit.aes = FALSE
  ) +
  geom_label_repel(
    data = arrow_df_all %>% filter(kind == "numeric"),
    aes(x = xend * 1.1, y = yend * 1.1, label = variable),
    size = 2.8, colour = "#c0392b",
    fill = alpha("white", 0.7), label.size = 0,
    max.overlaps = 20, seed = 42, inherit.aes = FALSE
  ) +
  scale_colour_manual(values = c(first = "#2980b9", last = "#e67e22"),
                      labels = c(first = "First survey", last = "Last survey")) +
  scale_shape_manual(values  = c(first = 16, last = 17),
                     labels  = c(first = "First survey", last = "Last survey")) +
  labs(
    title    = "NMDS of plant community composition (Bray–Curtis) — all plots",
    subtitle = sprintf("167 plots · stress = %.3f · paired PERMANOVA R² = 2.7%%, p = 0.001",
                       nmds_strict$stress),
    x = "NMDS1", y = "NMDS2", colour = NULL, shape = NULL,
    caption = "Arrows: envfit vectors (p ≤ 0.05). Lines connect first and last survey of the same plot."
  ) +
  theme_classic(base_size = 11) +
  theme(legend.position = "bottom")
```

## 1.2 Outlier identification

The three plots sitting far to the right are coastal/sandy communities.
Their species composition is so distinct from the rest of the Icelandic
vegetation dataset that NMDS places them ~17 units away from the main cloud,
stretching the axis and making the central cloud unreadable.

```{r outlier-table}
# Identify the six rows (first + last) for the three outlier plots
outlier_plots <- c("RN-02", "RN-03", "RN-65-01")

meta_outliers <- meta_strict %>%
  filter(plot_number %in% outlier_plots) %>%
  select(plot_number, sample, háplöntuþekja, soil_depth, total_cover,
         vegetation_height_1, soil_type_name, moisture_name, topography_name)

print_full(meta_outliers,
           caption = "Metadata for the three outlier plots (both surveys).")
```

```{r taxa-shift-outliers}
# Which taxa are responsible for the outlier separation?
# Compare mean relative abundance in outlier vs non-outlier rows.
outlier_rows     <- meta_strict$plot_number %in% outlier_plots
mat_rel          <- mat_strict / rowSums(mat_strict)

taxa_shift <- tibble(
  taxon               = colnames(mat_strict),
  mean_rel_outlier    = colMeans(mat_rel[outlier_rows,  , drop = FALSE]),
  mean_rel_non_outlier = colMeans(mat_rel[!outlier_rows, , drop = FALSE])
) %>%
  mutate(delta_rel = mean_rel_outlier - mean_rel_non_outlier) %>%
  arrange(desc(abs(delta_rel)))

print_full(taxa_shift %>% slice_head(n = 30),
           caption = "Top 30 taxa driving the outlier separation (|Δ relative abundance|).")
```

---

# Part 2 — Core NMDS (outliers excluded)

```{r build-core-matrices, cache=TRUE}
keep_rows <- !meta_strict$plot_number %in% outlier_plots
meta_core <- meta_strict[keep_rows, ]
mat_core  <- mat_strict[keep_rows, ]

# Drop taxa that only occurred in the removed plots
taxa_present <- colSums(mat_core) > 0
mat_core     <- mat_core[, taxa_present]
```

After removing **`r length(outlier_plots)` plots** (`r paste(outlier_plots, collapse = ", ")`),
the core matrix has **`r nrow(mat_core)` rows** and **`r ncol(mat_core)` taxa**.
The three excluded plots are analysed separately in Part 5.

## 2.1 Re-run NMDS

```{r nmds-core, cache=TRUE}
nmds_core <- metaMDS(
  mat_core,
  distance = "bray",
  k        = 2,
  trymax   = 200,
  trace    = FALSE
)
```

**NMDS stress (core, k = 2): `r round(nmds_core$stress, 4)`**

```{r stress-interpretation, echo=FALSE, results='asis'}
s <- nmds_core$stress
interp <- dplyr::case_when(
  s < 0.10 ~ "**Excellent** — ordination is a reliable low-dimensional representation.",
  s < 0.15 ~ "**Good** — ordination is suitable for publication.",
  s < 0.20 ~ "**Acceptable** — ordination is usable; mention stress in figure caption.",
  TRUE     ~ "**Poor** — consider k = 3 or interpret with caution."
)
cat(paste0("> Stress interpretation: ", interp, "\n"))

if (s > 0.20) {
  cat("\n> ⚠️  **Stress > 0.20.** Consider re-running with `k = 3`.",
      "A 3D ordination can be visualised as a pairs plot (`pairs(scores(nmds_core))`).\n")
}
```

### Shepard plot

```{r shepard, fig.height=5, fig.width=5, fig.cap="Shepard diagram: ordination distances vs. dissimilarities. Points should follow the step function tightly."}
stressplot(nmds_core,
           main = sprintf("Shepard plot — stress %.4f", nmds_core$stress))
```

## 2.2 Re-fit environmental vectors

```{r envfit-core, cache=TRUE}
env_core <- meta_core %>%
  select(all_of(vars_driver_strict_final)) %>%
  mutate(
    across(where(is.character), as.factor),
    across(where(is.logical),   as.factor)
  )

fit_core <- envfit(nmds_core, env_core, permutations = 999)
```

```{r envfit-table}
fit_tbl_core <- bind_rows(
  tibble(
    variable = names(fit_core$vectors$r),
    r2       = as.numeric(fit_core$vectors$r),
    p_value  = as.numeric(fit_core$vectors$pvals),
    kind     = "numeric"
  ),
  if (!is.null(fit_core$factors)) tibble(
    variable = names(fit_core$factors$r),
    r2       = as.numeric(fit_core$factors$r),
    p_value  = as.numeric(fit_core$factors$pvals),
    kind     = "factor"
  )
) %>%
  arrange(p_value) %>%
  mutate(sig = case_when(p_value < 0.001 ~ "***",
                         p_value < 0.01  ~ "**",
                         p_value < 0.05  ~ "*",
                         TRUE            ~ "ns"))

print_full(fit_tbl_core,
           caption = "envfit results on core dataset (outliers removed). r² = goodness-of-fit.")
```

## 2.3 NMDS biplot

```{r nmds-core-plot, fig.height=7.5, fig.cap="NMDS biplot of the core 164-plot dataset. Red arrows are significant envfit vectors (p ≤ 0.05); length is proportional to r². Grey lines connect first and last survey of the same plot."}

# --- Build arrow data frame ---
sig_thresh <- 0.05

arrow_df_core <- as.data.frame(fit_core$vectors$arrows) %>%
  rownames_to_column("variable") %>%
  rename(ax1 = NMDS1, ax2 = NMDS2) %>%
  mutate(
    r2      = fit_core$vectors$r,
    p_value = fit_core$vectors$pvals
  ) %>%
  filter(p_value <= sig_thresh)

# Factor centroids
factor_df_core <- tryCatch({
  as.data.frame(scores(fit_core, display = "factors")) %>%
    rownames_to_column("label") %>%
    rename(fax1 = NMDS1, fax2 = NMDS2) %>%
    mutate(
      p_value = fit_core$factors$pvals[
        sapply(label, function(nm) {
          matched <- names(fit_core$factors$pvals)[startsWith(nm, names(fit_core$factors$pvals))]
          if (length(matched)) matched[1] else NA_character_
        })
      ]
    ) %>%
    filter(!is.na(p_value), p_value <= sig_thresh)
}, error = function(e) tibble())

# --- Site scores ---
site_scores <- as.data.frame(scores(nmds_core, display = "sites")) %>%
  bind_cols(meta_core %>% select(plot_number, sample))

ordi_radius_core <- max(abs(c(site_scores$NMDS1, site_scores$NMDS2))) * 0.70

arrow_df_core <- arrow_df_core %>%
  mutate(xend = ax1 * ordi_radius_core,
         yend = ax2 * ordi_radius_core)

# --- Plot ---
p_nmds_core <- ggplot(site_scores, aes(x = NMDS1, y = NMDS2)) +
  geom_line(aes(group = plot_number),
            colour = "grey60", linewidth = 0.30, alpha = 0.55) +
  geom_point(aes(colour = sample, shape = sample),
             size = 2.0, alpha = 0.85) +
  geom_segment(
    data = arrow_df_core,
    aes(x = 0, y = 0, xend = xend, yend = yend),
    arrow = arrow(length = unit(0.22, "cm"), type = "closed"),
    colour = "#c0392b", linewidth = 0.75, inherit.aes = FALSE
  ) +
  geom_label_repel(
    data        = arrow_df_core,
    aes(x = xend * 1.12, y = yend * 1.12, label = variable),
    size        = 3.0, colour = "#c0392b",
    fill        = alpha("white", 0.80), label.size = 0,
    max.overlaps = 20, seed = 42, inherit.aes = FALSE
  ) +
  {
    if (nrow(factor_df_core) > 0)
      geom_label_repel(
        data        = factor_df_core,
        aes(x = fax1, y = fax2, label = label),
        size        = 2.5, colour = "#27ae60",
        fill        = alpha("white", 0.80), label.size = 0,
        max.overlaps = 15, seed = 43, inherit.aes = FALSE
      )
  } +
  scale_colour_manual(values = c(first = "#2980b9", last = "#e67e22"),
                      labels = c(first = "First survey", last = "Last survey")) +
  scale_shape_manual(values  = c(first = 16, last = 17),
                     labels  = c(first = "First survey", last = "Last survey")) +
  coord_equal() +
  labs(
    title    = "NMDS — core dataset (Bray–Curtis)",
    subtitle = sprintf(
      "%d plots · stress = %.3f · excl. %s",
      n_distinct(site_scores$plot_number),
      nmds_core$stress,
      paste(outlier_plots, collapse = ", ")
    ),
    x = "NMDS1", y = "NMDS2", colour = NULL, shape = NULL,
    caption = "Red arrows: envfit vectors (p ≤ 0.05); length ∝ r².\nGrey lines connect first and last survey of the same plot."
  ) +
  theme_classic(base_size = 11) +
  theme(
    legend.position = "bottom",
    plot.subtitle   = element_text(size = 8, colour = "grey40"),
    plot.caption    = element_text(size = 7.5, colour = "grey50", hjust = 0)
  )

p_nmds_core
```

---

# Part 3 — Per-plot compositional turnover

## 3.1 Bray–Curtis distance per plot (first → last)

```{r bc-per-plot}
bc_matrix <- as.matrix(vegdist(mat_strict, method = "bray"))

plot_bc <- meta_strict %>%
  mutate(row_idx = row_number()) %>%
  group_by(plot_number) %>%
  summarise(
    idx_first = row_idx[sample == "first"],
    idx_last  = row_idx[sample == "last"],
    .groups   = "drop"
  ) %>%
  mutate(bc_dist = map2_dbl(idx_first, idx_last, ~ bc_matrix[.x, .y]))

env_first <- meta_strict %>%
  filter(sample == "first") %>%
  select(plot_number, soil_type_name, moisture_name,
         háplöntuþekja, soil_depth, vegetation_height_1, total_cover)

plot_bc_env <- plot_bc %>%
  left_join(env_first, by = "plot_number")
```

```{r bc-summary-table}
bc_summary <- plot_bc_env %>%
  group_by(soil_type_name) %>%
  summarise(
    n         = n(),
    median_bc = round(median(bc_dist, na.rm = TRUE), 3),
    mean_bc   = round(mean(bc_dist,   na.rm = TRUE), 3),
    sd_bc     = round(sd(bc_dist,     na.rm = TRUE), 3),
    .groups   = "drop"
  ) %>%
  arrange(desc(median_bc))

print_full(bc_summary,
           caption = "Per-plot Bray–Curtis distance (first → last) summarised by soil type.")
```

## 3.2 Kruskal–Wallis tests

```{r kruskal-wallis}
kw_soil  <- kruskal.test(bc_dist ~ soil_type_name, data = plot_bc_env)
kw_moist <- kruskal.test(bc_dist ~ moisture_name,  data = plot_bc_env)

kw_tbl <- tibble(
  test      = c("Soil type", "Moisture class"),
  chi_sq    = c(kw_soil$statistic,  kw_moist$statistic),
  df        = c(kw_soil$parameter,  kw_moist$parameter),
  p_value   = c(kw_soil$p.value,    kw_moist$p.value),
  sig       = case_when(p_value < 0.001 ~ "***",
                        p_value < 0.01  ~ "**",
                        p_value < 0.05  ~ "*",
                        TRUE            ~ "ns")
)

print_full(kw_tbl,
           caption = "Kruskal–Wallis: does soil type / moisture predict magnitude of compositional change?")
```

```{r kw-followup, results='asis'}
if (kw_soil$p.value < 0.05) {
  cat("> **Soil type is significant (p < 0.05).** Consider a post-hoc Dunn test:\n")
  cat("> `dunn.test::dunn.test(plot_bc_env$bc_dist, plot_bc_env$soil_type_name, method = 'bh')`\n\n")
}
if (kw_moist$p.value < 0.05) {
  cat("> **Moisture class is significant (p < 0.05).** Consider a post-hoc Dunn test:\n")
  cat("> `dunn.test::dunn.test(plot_bc_env$bc_dist, plot_bc_env$moisture_name, method = 'bh')`\n\n")
}
```

## 3.3 Boxplot by soil type

```{r bc-boxplot, fig.cap="Per-plot Bray–Curtis distance (first → last survey) by soil type. Points are individual plots."}
ggplot(
  plot_bc_env %>% filter(!is.na(soil_type_name)),
  aes(x = reorder(soil_type_name, bc_dist, FUN = median),
      y = bc_dist, fill = soil_type_name)
) +
  geom_boxplot(alpha = 0.7, outlier.size = 1.2) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1) +
  labs(
    title    = "Compositional turnover by soil type",
    subtitle = "Per-plot Bray–Curtis distance (first → last survey)",
    x = "Soil type", y = "Bray–Curtis distance", fill = NULL
  ) +
  theme_classic(base_size = 11) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1),
        legend.position = "none")
```

---

# Part 4 — Indicator species analysis (IndVal)

Using `indicspecies::multipatt()` to identify taxa that are significantly
associated with either the first or last survey, using IndVal.g with 999
permutations. P-values are Benjamini–Hochberg FDR-corrected.

```{r indval, cache=TRUE}
group_vec <- ifelse(meta_strict$sample == "first", 1L, 2L)

indval_result <- multipatt(
  mat_strict,
  cluster = group_vec,
  control = how(nperm = 999),
  func    = "IndVal.g",
  duleg   = TRUE
)

indval_tbl <- indval_result$sign %>%
  rownames_to_column("taxon") %>%
  as_tibble() %>%
  rename(group = index) %>%
  mutate(
    survey = if_else(group == 1L, "first", "last"),
    p_adj  = p.adjust(p.value, method = "BH")
  ) %>%
  filter(p_adj <= 0.05) %>%
  arrange(p_adj)
```

**`r nrow(indval_tbl)` taxa** pass FDR ≤ 5%:
`r sum(indval_tbl$survey == "first")` associated with the **first** survey,
`r sum(indval_tbl$survey == "last")` with the **last** survey.

## 4.1 Full indicator taxa table

```{r indval-table}
print_full(
  indval_tbl %>%
    select(taxon, survey, stat, p.value, p_adj) %>%
    mutate(sig = case_when(p_adj < 0.001 ~ "***",
                           p_adj < 0.01  ~ "**",
                           p_adj < 0.05  ~ "*",
                           TRUE          ~ "ns")),
  caption = "Indicator taxa (IndVal.g, FDR ≤ 5%). 'first' = associated with first survey; 'last' = last survey."
)
```

## 4.2 Top 15 per survey — bar chart

```{r indval-plot, fig.height=7, fig.cap="Top 15 indicator taxa per survey by IndVal.g statistic (FDR ≤ 5%)."}
top_indval <- indval_tbl %>%
  group_by(survey) %>%
  slice_min(order_by = p_adj, n = 15, with_ties = FALSE) %>%
  ungroup()

ggplot(top_indval,
       aes(x = stat, y = reorder(taxon, stat), fill = survey)) +
  geom_col(alpha = 0.85) +
  facet_wrap(~survey, scales = "free_y",
             labeller = as_labeller(c(first = "First survey", last = "Last survey"))) +
  scale_fill_manual(values = c(first = "#2980b9", last = "#e67e22")) +
  labs(
    title    = "Indicator taxa — first vs last survey",
    subtitle = "IndVal.g statistic; top 15 per group (FDR ≤ 5%)",
    x = "IndVal statistic", y = NULL, fill = NULL
  ) +
  theme_classic(base_size = 10) +
  theme(legend.position = "none",
        strip.background = element_rect(fill = "grey92", colour = NA))
```

---

# Part 5 — Univariate change in community descriptors

Paired Wilcoxon signed-rank test for each continuous driver variable.
Each plot contributes one matched pair (first survey value, last survey value).

```{r univar-wilcox}
univar_vars <- c("háplöntuþekja", "vegetation_height_1", "total_cover", "soil_depth")

env_paired_wide <- meta_strict %>%
  select(plot_number, sample, all_of(univar_vars)) %>%
  pivot_wider(
    names_from  = sample,
    values_from = all_of(univar_vars),
    names_sep   = "_"
  )

wilcox_results <- map_dfr(univar_vars, function(v) {
  first_col    <- paste0(v, "_first")
  last_col     <- paste0(v, "_last")
  x            <- env_paired_wide[[first_col]]
  y            <- env_paired_wide[[last_col]]
  complete_idx <- complete.cases(x, y)
  n_pairs      <- sum(complete_idx)
  test <- wilcox.test(y[complete_idx], x[complete_idx],
                      paired = TRUE, conf.int = TRUE)
  tibble(
    variable      = v,
    n_pairs       = n_pairs,
    median_first  = median(x[complete_idx], na.rm = TRUE),
    median_last   = median(y[complete_idx], na.rm = TRUE),
    median_change = median(y[complete_idx] - x[complete_idx], na.rm = TRUE),
    ci_low        = test$conf.int[1],
    ci_high       = test$conf.int[2],
    W             = test$statistic,
    p_value       = test$p.value
  )
}) %>%
  mutate(
    p_adj = p.adjust(p_value, method = "BH"),
    sig   = case_when(p_adj < 0.001 ~ "***",
                      p_adj < 0.01  ~ "**",
                      p_adj < 0.05  ~ "*",
                      TRUE          ~ "ns")
  )
```

## 5.1 Results table

```{r univar-table}
print_full(wilcox_results,
           caption = "Paired Wilcoxon signed-rank test results. median_change = last − first. BH-adjusted p-values.")
```

## 5.2 Change distributions

```{r univar-plot, fig.cap="Distribution of per-plot change (last − first) for each community descriptor. Significance stars are BH-adjusted."}
env_paired_long_change <- map_dfr(univar_vars, function(v) {
  first_col <- paste0(v, "_first")
  last_col  <- paste0(v, "_last")
  env_paired_wide %>%
    transmute(
      plot_number = plot_number,
      variable    = v,
      change      = .data[[last_col]] - .data[[first_col]]
    )
}) %>%
  filter(!is.na(change)) %>%
  left_join(wilcox_results %>% select(variable, p_adj, sig), by = "variable")

y_top <- max(env_paired_long_change$change, na.rm = TRUE) * 1.07

ggplot(env_paired_long_change, aes(x = variable, y = change)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_violin(fill = "#bdc3c7", alpha = 0.5, trim = TRUE) +
  geom_boxplot(width = 0.18, fill = "white", outlier.size = 0.8) +
  geom_text(
    data = env_paired_long_change %>% distinct(variable, sig),
    aes(y = y_top, label = sig),
    size = 5, colour = "#c0392b"
  ) +
  labs(
    title    = "Change in continuous community descriptors (last − first)",
    subtitle = "Paired Wilcoxon; * p<0.05, ** p<0.01, *** p<0.001 (BH-adjusted)",
    x = NULL, y = "Change (last − first)"
  ) +
  theme_classic(base_size = 11)
```

---

# Part 6 — Excluded coastal plots

With only 6 rows (3 plots × 2 surveys), NMDS is numerically unstable.
PCoA (metric multidimensional scaling) is used instead.

```{r coastal-pcoa, fig.cap="PCoA of the three excluded coastal/outlier plots. Labels show plot name and survey. Lines connect first and last survey of the same plot."}

outlier_rows <- meta_strict$plot_number %in% outlier_plots
mat_coastal  <- mat_strict[outlier_rows, ]
meta_coastal <- meta_strict[outlier_rows, ]
mat_coastal  <- mat_coastal[, colSums(mat_coastal) > 0]

pcoa_coastal <- cmdscale(
  vegdist(mat_coastal, method = "bray"),
  k   = 2,
  eig = TRUE
)

pcoa_df <- as.data.frame(pcoa_coastal$points) %>%
  setNames(c("Axis1", "Axis2")) %>%
  bind_cols(meta_coastal %>% select(plot_number, sample))

var_explained <- round(
  100 * pcoa_coastal$eig[1:2] / sum(abs(pcoa_coastal$eig)), 1
)

ggplot(pcoa_df, aes(x = Axis1, y = Axis2)) +
  geom_line(aes(group = plot_number), colour = "grey60", linewidth = 0.4) +
  geom_point(aes(colour = sample, shape = sample), size = 3) +
  geom_label_repel(
    aes(label = paste(plot_number, sample)),
    size = 3, max.overlaps = 20, seed = 1
  ) +
  scale_colour_manual(values = c(first = "#2980b9", last = "#e67e22"),
                      labels = c(first = "First survey", last = "Last survey")) +
  scale_shape_manual(values  = c(first = 16, last = 17),
                     labels  = c(first = "First survey", last = "Last survey")) +
  coord_equal() +
  labs(
    title    = "PCoA — excluded coastal/outlier plots",
    subtitle = sprintf("Bray–Curtis · Axis 1 = %.1f%% · Axis 2 = %.1f%%",
                       var_explained[1], var_explained[2]),
    x = sprintf("PCoA Axis 1 (%.1f%%)", var_explained[1]),
    y = sprintf("PCoA Axis 2 (%.1f%%)", var_explained[2]),
    colour = NULL, shape = NULL
  ) +
  theme_classic(base_size = 11) +
  theme(legend.position = "bottom")
```

```{r coastal-species}
# Top species in each coastal plot (last survey only, for brevity)
coastal_last <- meta_coastal %>%
  filter(sample == "last") %>%
  pull(plot_number)

coastal_species <- community_long_paired %>%
  filter(plot_number %in% coastal_last, sample == "last") %>%
  arrange(plot_number, desc(cover)) %>%
  group_by(plot_number) %>%
  slice_head(n = 10) %>%
  ungroup()

print_full(coastal_species,
           caption = "Top 10 species by cover in each coastal plot (last survey).")
```

---

# Part 7 — Suggested next steps

Based on the results above, the following analyses are the logical
continuation — choose based on what the figures showed.

```{r next-steps, echo=FALSE, results='asis'}
cat(paste(
  "**A — dbRDA (distance-based redundancy analysis)**",
  "If háplöntuþekja or vegetation height changed significantly in Part 5, use",
  "dbRDA to formally partition variance between the temporal effect (*sample*)",
  "and the environmental gradient, while conditioning on plot identity:",
  "",
  "```r",
  "vegan::dbrda(",
  "  mat_strict ~ sample + háplöntuþekja + Condition(plot_number),",
  "  data   = meta_strict,",
  "  method = \"bray\"",
  ")",
  "```",
  "",
  "**B — Stratum-specific PERMANOVAs**",
  "If soil type or moisture was significant in the Kruskal–Wallis (Part 3),",
  "split the dataset by those classes and re-run paired PERMANOVA within each",
  "stratum — this tests whether the temporal signal is consistent across",
  "habitat types or concentrated in one.",
  "",
  "**C — Taxon-level ecology**",
  "Review the indicator taxa from Part 4 against Icelandic vegetation",
  "literature. Key questions:",
  "- Are the 'last survey' gainers typically associated with warming, reduced",
  "  grazing, or nitrogen deposition in Icelandic heathlands?",
  "- Do *Racomitrium* or *Dryas* appear as consistent losers?",
  "- Are wetland sedges (*Carex lyngbyei*, *C. nigra*) showing directional change?",
  "",
  "**D — Spatial autocorrelation of turnover**",
  "Test whether compositional change (the per-plot Bray–Curtis from Part 3) is",
  "spatially structured using a Mantel test:",
  "",
  "```r",
  "geo_dist <- dist(cbind(env_first$longitude_1, env_first$latitude_1))",
  "vegan::mantel(bc_dist_matrix, geo_dist, method = \"pearson\", permutations = 999)",
  "```",
  sep = "\n"
))
```

---

# Session info

```{r session-info}
sessionInfo()
```
