---
title: "v_plants — Paired Resurvey Analysis"
author: "Valtýr"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: kable
    code_folding: hide
    fig_width: 9
    fig_height: 6
  word_document:
    toc: true
    toc_depth: 3
    fig_width: 9
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo        = TRUE,
  warning     = FALSE,
  message     = FALSE,
  cache       = TRUE,         # long-running chunks (NMDS, IndVal) cached to disk
  cache.path  = "cache/v_plants_",
  fig.align   = "center",
  fig.path    = "outputs/v_plants_analysis-",
  dev         = "png",
  dpi         = 300
)

# Helper: print a tibble fully if ≤ 400 rows, truncated otherwise
print_full <- function(x, caption = NULL) {
  n <- nrow(x)
  if (n <= 400) {
    knitr::kable(x, digits = 4, caption = caption)
  } else {
    knitr::kable(
      head(x, 400),
      digits  = 4,
      caption = paste0(caption, sprintf("  [showing first 400 of %d rows]", n))
    )
  }
}
```

```{r libraries}
library(tidyverse)
library(vegan)
library(broom)
library(ggrepel)
library(indicspecies)
library(knitr)

set.seed(123)
```

```{r bootstrap-required-objects}
# Always source the prep script. NMDS caching is handled inside prep itself —
# it checks for outputs/nmds_strict_cache.rds and loads or recomputes as needed,
# so sourcing prep is fast when the cache exists.
prep_script <- "v_plants_prep.R"
if (!file.exists(prep_script)) {
  stop("Required script not found: ", prep_script)
}
source(prep_script, local = .GlobalEnv)
```

---

> **Purpose.**  This document picks up directly after `v_plants_variable_screening.html`.
> All core objects are created by `v_plants_prep.R` (sourced above); the screening
> document is an audit trail that describes how they were selected and filtered.
>
> | Object | Description |
> |---|---|
> | `community_long_paired` | Long-format cover table (plot × sample × taxon) |
> | `community_wide_strict` | Wide community matrix with `plot_number` + `sample` columns |
> | `meta_strict` | Metadata aligned row-for-row with `community_wide_strict` |
> | `mat_strict` | Numeric species matrix (`r nrow(mat_strict)` rows × `r ncol(mat_strict)` cols) |
> | `nmds_strict` | `metaMDS` result on `mat_strict` (all plots, pre-outlier removal) |
> | `fit_strict` | `envfit` result on `nmds_strict` |
> | `vars_driver_strict_final`, `driver_final` | Variable vectors from screening |

---

## Analytical strategy — why NMDS first, then RDA?

The distinction comes down to what question each method is designed to answer.

**NMDS** is unconstrained — the ordination axes are determined entirely by the
species composition data. The community structure emerges first, and you then
ask whether environmental variables happen to correlate with it (envfit). The
question is: *what is the main structure of variation in these communities, and
does anything we measured line up with it?* Nothing is assumed about which
variables matter or how they relate to composition. This is the right starting
point for a dataset spanning many habitat types across Iceland, where you don't
yet know which gradients dominate.

**RDA/dbRDA** is constrained — the axes are forced to be linear combinations of
the predictors you specify. It can only show you the variation in composition
that is mathematically explainable by those specific variables. The question
shifts to: *how much of the community variance do these predictors formally
account for, and can I partition it cleanly?*

**The core argument for using both in sequence in this dataset.** The paired
temporal design means there are two distinct signals sitting on top of each
other: the baseline ecological gradient (wet vs dry, deep soil vs shallow, etc.)
and the change between first and last survey. NMDS with PERMANOVA establishes
that a temporal signal exists at all — that communities at the last survey are
measurably different from their own baseline. That is a prerequisite worth
demonstrating on its own before asking what drove the change.

RDA/dbRDA with `Condition(plot_number)` then partitions the temporal component
away from the among-plot gradient. By conditioning on plot identity you remove
the dominant baseline variation — the fact that a heath plot and a wetland plot
are just compositionally different — and look only at within-plot change over
time. That is where a variable like `haplontuthekja` (vascular plant cover)
becomes a legitimate predictor rather than a circular one. You are not asking
"do plots with more vascular cover have different communities?" (which is
trivially true), you are asking "did plots whose vascular cover *changed* show
corresponding compositional shifts?"

In short: NMDS describes the landscape of variation; RDA partitions it into
sources. Using RDA without NMDS first would risk constraining the ordination to
variables that miss the dominant gradients entirely.

---

# Part 1 — Preliminary NMDS (all plots)

## 1.1 Ordination plot

This first pass runs on the full `r length(unique(meta_strict$plot_number))`-plot dataset before any outlier removal.
It is included here as a diagnostic to show why the three coastal/dune plots
need special treatment.

```{r nmds-all-plot, fig.height=7,fig.cap=sprintf("NMDS of all %d revisited plots. Note the three outlier sites near NMDS1 = 17 that stretch the axis and compress the main community cloud.",length(unique(meta_strict$plot_number)))}

# Extract site scores
nmds_scores <- as.data.frame(scores(nmds_strict, display = "sites")) %>%
  bind_cols(meta_strict %>% select(plot_number, sample))

# Build envfit data including gap_years in addition to the strict drivers
# Exploratory columns (gap_years, grytnithekja) use any_of() so they are silently
# dropped if absent from meta_strict (meta_strict only carries vars_driver_strict_final
# + extra_env_exploratory; grytnithekja is available in env_candidates_ps / v_plants_use
# but is not pulled into meta_strict by default).
env_all <- meta_strict %>%
  select(
    all_of(vars_driver_strict_final),          # required core drivers
    any_of(c(
      "gap_years",                             # time gap effect
      #"habitat_type_name",                   # broader habitat grouping
      "grytnithekja"                           # exploratory rock cover (if present)
    ))
  ) %>%
  mutate(
    across(where(is.character), as.factor),
    across(where(is.logical),   as.factor)
  )

fit_all <- vegan::envfit(nmds_strict, env_all, permutations = 999, na.rm=T)

# --- envfit arrows (continuous vectors) ---
sig_thresh <- 0.05

vec_df <- tibble(variable = character(), NMDS1 = numeric(), NMDS2 = numeric(),
                 r2 = numeric(), p_value = numeric(), kind = character())

if (!is.null(fit_all$vectors) && !is.null(fit_all$vectors$arrows)) {
  vec_raw  <- as.data.frame(fit_all$vectors$arrows) %>%
    rownames_to_column("variable")
  vec_axes <- setdiff(names(vec_raw), "variable")
  if (length(vec_axes) >= 2) {
    vec_df <- vec_raw %>%
      rename(NMDS1 = all_of(vec_axes[1]), NMDS2 = all_of(vec_axes[2])) %>%
      select(variable, NMDS1, NMDS2) %>%
      mutate(
        r2      = as.numeric(fit_all$vectors$r[variable]),
        p_value = as.numeric(fit_all$vectors$pvals[variable]),
        kind    = "numeric"
      )
  }
}

# --- envfit factor centroids ---
fac_df <- tibble(variable = character(), NMDS1 = numeric(), NMDS2 = numeric(),
                 r2 = numeric(), p_value = numeric(), kind = character())

if (!is.null(fit_all$factors) && !is.null(fit_all$factors$centroids)) {
  fac_raw  <- as.data.frame(fit_all$factors$centroids) %>%
    rownames_to_column("variable")
  fac_axes <- setdiff(names(fac_raw), "variable")
  if (length(fac_axes) >= 2) {
    fac_df <- fac_raw %>%
      rename(NMDS1 = all_of(fac_axes[1]), NMDS2 = all_of(fac_axes[2])) %>%
      select(variable, NMDS1, NMDS2) %>%
      mutate(r2 = NA_real_, p_value = NA_real_, kind = "factor")
  }
}

arrow_df_all <- bind_rows(vec_df, fac_df) %>%
  filter(is.na(p_value) | p_value <= sig_thresh)

ordi_radius_all <- max(abs(c(nmds_scores$NMDS1, nmds_scores$NMDS2))) * 0.7

arrow_df_all <- arrow_df_all %>%
  mutate(xend = NMDS1 * ordi_radius_all,
         yend = NMDS2 * ordi_radius_all)

# One label per plot at midpoint between first and last survey
nmds_plot_labels <- nmds_scores %>%
  group_by(plot_number) %>%
  summarise(NMDS1 = mean(NMDS1), NMDS2 = mean(NMDS2), .groups = "drop")

pnmds <- ggplot(nmds_scores, aes(x = NMDS1, y = NMDS2)) +
  geom_line(aes(group = plot_number),
            colour = "grey70", linewidth = 0.3, alpha = 0.6) +
  geom_point(aes(colour = sample, shape = sample),
             size = 2.0, alpha = 0.85) +
  geom_text_repel(
    data = nmds_plot_labels,
    aes(x = NMDS1, y = NMDS2, label = plot_number),
    size = 1.7, colour = "grey35",
    segment.size = 0.2, segment.colour = "grey60",
    max.overlaps = Inf, seed = 77, inherit.aes = FALSE
  ) +
  geom_segment(
    data = arrow_df_all %>% filter(kind == "numeric"),
    aes(x = 0, y = 0, xend = xend, yend = yend),
    arrow = arrow(length = unit(0.2, "cm")),
    colour = "#c0392b", linewidth = 0.7, inherit.aes = FALSE
  ) +
  geom_label_repel(
    data = arrow_df_all %>% filter(kind == "numeric"),
    aes(x = xend * 1.1, y = yend * 1.1, label = variable),
    size = 2.8, colour = "#c0392b",
    fill = alpha("white", 0.7), label.size = 0,
    max.overlaps = 20, seed = 42, inherit.aes = FALSE
  ) +
  scale_colour_manual(values = c(first = "#2980b9", last = "#e67e22"),
                      labels = c(first = "First survey", last = "Last survey")) +
  scale_shape_manual(values  = c(first = 16, last = 17),
                     labels  = c(first = "First survey", last = "Last survey")) +
  labs(
    title    = "NMDS of plant community composition (Bray–Curtis) — all plots",
    subtitle = sprintf(
      "%d plots · stress = %.3f · paired PERMANOVA R² = %.1f%%, p = %.3f",
      length(unique(nmds_scores$plot_number)),
      nmds_strict$stress,
      perm_strict[1, "R2"] * 100,
      perm_strict[1, "Pr(>F)"]
    ),
    x = "NMDS1", y = "NMDS2", colour = NULL, shape = NULL,
    caption = "Arrows: envfit vectors (p ≤ 0.05). Lines connect first and last survey of the same plot."
  ) +
  theme_classic(base_size = 11) +
  theme(legend.position = "bottom")
```

```{r permanova-results}
# Paired PERMANOVA: does community composition differ between first and last survey?
# strata = plot_number ensures permutations are constrained within plots (paired design).
# Row 1 = predictor (labelled "Model" in vegan ≥ 2.6, "sample" in older versions)
# Row 2 = Residual — use row index rather than name for version robustness
perm_r2  <- round(perm_strict[1, "R2"] * 100, 1)
perm_p   <- perm_strict[1, "Pr(>F)"]
perm_f   <- round(perm_strict[1, "F"], 2)
perm_df1 <- perm_strict[1, "Df"]
perm_df2 <- perm_strict[2, "Df"]

cat(sprintf(
  "Paired PERMANOVA (adonis2, Bray–Curtis, strata = plot):\n  F(%d, %d) = %.2f,  R² = %.1f%%,  p = %.3f\n",
  perm_df1, perm_df2, perm_f, perm_r2, perm_p
))

print(perm_strict)
```

**What the test did.** Each plot was surveyed twice — once early, once late.
The test asks: are the two surveys from the same plot more similar to each
other than they are to surveys of other plots at different times? Permutations
were constrained within each plot (`strata = plot_number`), so the test is
genuinely paired — it can only detect change *within* plots, not just
differences *between* them.

**F(`r perm_df1`, `r perm_df2`) = `r perm_f`.** The F-statistic compares the
variance explained by survey timing (first vs last) to the residual variance.
`r perm_f` is well above 1, meaning the temporal effect is real and not lost in
noise. The degrees of freedom tell you the design: 1 for the predictor (survey:
first or last), `r perm_df2` for the residual
(`r length(unique(meta_strict$plot_number))` plots × 2 surveys − 2).

**R² = `r perm_r2`%.** This is the important one to interpret carefully. It
does *not* mean the temporal change is weak or unimportant. It means that out
of all the compositional variation in the dataset — the enormous differences
between a highland heath, a wetland, a lava field, and a birch shrubland —
`r perm_r2`% is attributable to the first-vs-last contrast. The remaining
~`r 100 - perm_r2`% is the baseline variation *between* plots, which dwarfs
the within-plot change. This is completely normal and expected in a paired
vegetation study spanning many habitat types. The paired design was chosen
precisely because it strips out that dominant between-plot variation and focuses
on the within-plot signal.

**p = `r perm_p`.** With 999 permutations this is the minimum achievable
p-value, meaning none of the 999 random reshufflings produced an F as large as
`r perm_f`. The temporal signal is statistically unambiguous.

**In plain terms.** Icelandic plant communities changed measurably between
their first and last survey. The change is real and consistent enough to survive
a rigorous paired permutation test. It is small relative to the total variation
in the dataset, which is expected — a heath is still a heath after ten years,
it just has a slightly different species mix. The R² is not a measure of
ecological importance; it is a measure of how much of the total dataset variance
this one contrast accounts for. The F-statistic and p-value are what establish
whether the change is real.

## 1.2 Outlier identification

The three plots sitting far to the right are coastal/sandy communities.
Their species composition is so distinct from the rest of the Icelandic
vegetation dataset that NMDS places them ~17 units away from the main cloud,
stretching the axis and making the central cloud unreadable.

```{r outlier-table}
# Data‑driven outlier detection on the all‑plots NMDS
nmds_scores_all <- as.data.frame(scores(nmds_strict, display = "sites")) %>%
  bind_cols(meta_strict %>% select(plot_number, sample))

# 1) Distance from the NMDS centroid
centroid <- colMeans(nmds_scores_all[, c("NMDS1", "NMDS2")])
nmds_scores_all <- nmds_scores_all %>%
  mutate(dist_center = sqrt((NMDS1 - centroid["NMDS1"])^2 +
                            (NMDS2 - centroid["NMDS2"])^2))

# 2) Flag plots whose distance is in the top tail (e.g. top 3 plots)
outlier_plots <- nmds_scores_all %>%
  arrange(desc(dist_center)) %>%
  distinct(plot_number, .keep_all = TRUE) %>%
  slice_head(n = 3) %>%                # or use a threshold on dist_center
  pull(plot_number)

# Inspect which plots were flagged
outlier_plots

meta_outliers <- meta_strict %>%
  filter(plot_number %in% outlier_plots) %>%
  select(
    plot_number, sample,
    any_of(c("soil_depth", "grytnithekja",
             "vegetation_height_mean", "soil_type_name",
             "moisture_name", "topography_name"))
  )

print_full(
  meta_outliers,
  caption = "Metadata for automatically detected outlier plots (both surveys)."
)
```

```{r taxa-shift-outliers}
# Which taxa are responsible for the outlier separation?
# Compare mean relative abundance in outlier vs non-outlier rows.
outlier_rows     <- meta_strict$plot_number %in% outlier_plots
mat_rel          <- mat_strict / rowSums(mat_strict)

taxa_shift <- tibble(
  taxon               = colnames(mat_strict),
  mean_rel_outlier    = colMeans(mat_rel[outlier_rows,  , drop = FALSE]),
  mean_rel_non_outlier = colMeans(mat_rel[!outlier_rows, , drop = FALSE])
) %>%
  mutate(delta_rel = mean_rel_outlier - mean_rel_non_outlier) %>%
  arrange(desc(abs(delta_rel)))

print_full(taxa_shift %>% slice_head(n = 30),
           caption = "Top 30 taxa driving the outlier separation (|Δ relative abundance|).")
```

In other words, these plots are outliers in the ordination **because of their taxonomic composition**, not because of extreme values in the metadata. They are dominated by coastal/sandy specialists such as `Puccinellia coarctata`, `Leymus arenarius`, `Honckenya peploides`, `Cakile maritima` and `Mertensia maritima`, which are almost absent elsewhere in the dataset. This makes their Bray–Curtis distances to the rest of the Icelandic vegetation so large that NMDS places them far away along one axis, even though their plot-level covariates (soil depth, total cover, tall-plant cover, moisture, topography) are within otherwise reasonable ranges. For this reason they are best treated as a separate coastal subset rather than as part of the main inland gradient.

---

# Part 2 — Core NMDS (outliers excluded)

```{r build-core-matrices, cache=TRUE}
keep_rows <- !meta_strict$plot_number %in% outlier_plots
meta_core <- meta_strict[keep_rows, ]
mat_core  <- mat_strict[keep_rows, ]

# Drop taxa that only occurred in the removed plots
taxa_present <- colSums(mat_core) > 0
mat_core     <- mat_core[, taxa_present]
```

After removing **`r length(outlier_plots)` plots** (`r paste(outlier_plots, collapse = ", ")`),
the core matrix has **`r nrow(mat_core)` rows** and **`r ncol(mat_core)` taxa**.
The three excluded plots are analysed separately in Part 5.

## 2.1 Re-run NMDS

```{r nmds-core, cache=TRUE}
nmds_core <- metaMDS(
  mat_core,
  distance = "bray",
  k        = 2,
  trymax   = 200,
  trace    = FALSE
)
```

**NMDS stress (core, k = 2): `r round(nmds_core$stress, 4)`**

```{r stress-interpretation, echo=FALSE, results='asis'}
s <- nmds_core$stress
interp <- dplyr::case_when(
  s < 0.10 ~ "**Excellent** — ordination is a reliable low-dimensional representation.",
  s < 0.15 ~ "**Good** — ordination is suitable for publication.",
  s < 0.20 ~ "**Acceptable** — ordination is usable; mention stress in figure caption.",
  TRUE     ~ "**Poor** — consider k = 3 or interpret with caution."
)
cat(paste0("> Stress interpretation: ", interp, "\n"))

if (s > 0.20) {
  cat("\n> ⚠️  **Stress > 0.20.** Consider re-running with `k = 3`.",
      "A 3D ordination can be visualised as a pairs plot (`pairs(scores(nmds_core))`).\n")
}
```

### 2.1b Optional: 3D NMDS (diagnostic)

```{r nmds-core-3d, eval=FALSE, fig.cap="3D NMDS (k = 3): pairwise site scores for the core dataset.", fig.height=6, fig.width=6, cache=TRUE, include=FALSE}
# 3D NMDS to check whether a higher-dimensional solution improves stress
nmds_core_3d <- metaMDS(
  mat_core,
  distance = "bray",
  k        = 3,
  trymax   = 200,
  trace    = FALSE
)

# Fit the same environmental set on the 3D solution
fit_core_3d <- envfit(nmds_core_3d, env_core, permutations = 999)

# Inspect stress (should be noticeably lower than the 2D value)
print(nmds_core_3d$stress)

# Visualise the 3D ordination as pairwise axis plots
pairs(
  scores(nmds_core_3d, display = "sites"),
  main = sprintf("3D NMDS (k = 3): stress = %.3f", nmds_core_3d$stress)
)
```

For interpretation, we use the **2D NMDS (`nmds_core`) with `fit_core` arrows**
as the main ordination figure. The arrows show which environmental gradients
best align with the dominant compositional axes.

The **3D NMDS (`nmds_core_3d` with `fit_core_3d`) is used as a diagnostic only**:
its lower stress confirms that the inland community structure is genuinely
three‑dimensional, but we visualise it via a pairs plot without envfit arrows.
Adding vectors directly in 3D would require multiple views or interactive
graphics and tends to obscure rather than clarify the main patterns, so all
arrow-based interpretation remains on the 2D biplot.


### Shepard plot

```{r shepard, fig.height=5, fig.width=5, fig.cap="Shepard diagram: ordination distances vs. dissimilarities. Points should follow the step function tightly."}
stressplot(nmds_core,
           main = sprintf("Shepard plot — stress %.4f", nmds_core$stress))
```

## 2.2 Re-fit environmental vectors

```{r envfit-core, cache=TRUE}
env_core <- meta_core %>%
  select(all_of(vars_driver_strict_final)) %>%
  mutate(
    across(where(is.character), as.factor),
    across(where(is.logical),   as.factor)
  )

fit_core <- envfit(nmds_core, env_core, permutations = 999)
```

```{r envfit-table}
fit_tbl_core <- bind_rows(
  tibble(
    variable = names(fit_core$vectors$r),
    r2       = as.numeric(fit_core$vectors$r),
    p_value  = as.numeric(fit_core$vectors$pvals),
    kind     = "numeric"
  ),
  if (!is.null(fit_core$factors)) tibble(
    variable = names(fit_core$factors$r),
    r2       = as.numeric(fit_core$factors$r),
    p_value  = as.numeric(fit_core$factors$pvals),
    kind     = "factor"
  )
) %>%
  arrange(p_value) %>%
  mutate(sig = case_when(p_value < 0.001 ~ "***",
                         p_value < 0.01  ~ "**",
                         p_value < 0.05  ~ "*",
                         TRUE            ~ "ns"))

print_full(fit_tbl_core,
           caption = "envfit results on core dataset (outliers removed). r² = goodness-of-fit.")
```

## 2.3 NMDS biplot

```{r nmds-core-plot, fig.height=7.5, fig.cap=sprintf("NMDS biplot of the core %d-plot dataset. Red arrows are significant envfit vectors (p ≤ 0.05); length is proportional to r². Grey lines connect first and last survey of the same plot.",length(unique(meta_core$plot_number)))}

# --- Build arrow data frame ---
sig_thresh <- 0.05

arrow_df_core <- as.data.frame(fit_core$vectors$arrows) %>%
  rownames_to_column("variable") %>%
  rename(ax1 = NMDS1, ax2 = NMDS2) %>%
  mutate(
    r2      = fit_core$vectors$r,
    p_value = fit_core$vectors$pvals
  ) %>%
  filter(p_value <= sig_thresh)

# Factor centroids
factor_df_core <- tryCatch({
  as.data.frame(scores(fit_core, display = "factors")) %>%
    rownames_to_column("label") %>%
    rename(fax1 = NMDS1, fax2 = NMDS2) %>%
    mutate(
      p_value = fit_core$factors$pvals[
        sapply(label, function(nm) {
          matched <- names(fit_core$factors$pvals)[startsWith(nm, names(fit_core$factors$pvals))]
          if (length(matched)) matched[1] else NA_character_
        })
      ]
    ) %>%
    filter(!is.na(p_value), p_value <= sig_thresh)
}, error = function(e) tibble())

# --- Site scores ---
site_scores <- as.data.frame(scores(nmds_core, display = "sites")) %>%
  bind_cols(meta_core %>% select(plot_number, sample))

ordi_radius_core <- max(abs(c(site_scores$NMDS1, site_scores$NMDS2))) * 0.70

arrow_df_core <- arrow_df_core %>%
  mutate(xend = ax1 * ordi_radius_core,
         yend = ax2 * ordi_radius_core)

# One label per plot at midpoint between first and last survey
core_plot_labels <- site_scores %>%
  group_by(plot_number) %>%
  summarise(NMDS1 = mean(NMDS1), NMDS2 = mean(NMDS2), .groups = "drop")

# --- Plot ---
p_nmds_core <- ggplot(site_scores, aes(x = NMDS1, y = NMDS2)) +
  geom_line(aes(group = plot_number),
            colour = "grey60", linewidth = 0.30, alpha = 0.55) +
  geom_point(aes(colour = sample, shape = sample),
             size = 2.0, alpha = 0.85) +
  geom_text_repel(
    data = core_plot_labels,
    aes(x = NMDS1, y = NMDS2, label = plot_number),
    size = 1.7, colour = "grey35",
    segment.size = 0.2, segment.colour = "grey60",
    max.overlaps = Inf, seed = 77, inherit.aes = FALSE
  ) +
  geom_segment(
    data = arrow_df_core,
    aes(x = 0, y = 0, xend = xend, yend = yend),
    arrow = arrow(length = unit(0.22, "cm"), type = "closed"),
    colour = "#c0392b", linewidth = 0.75, inherit.aes = FALSE
  ) +
  geom_label_repel(
    data        = arrow_df_core,
    aes(x = xend * 1.12, y = yend * 1.12, label = variable),
    size        = 3.0, colour = "#c0392b",
    fill        = alpha("white", 0.80), label.size = 0,
    max.overlaps = 20, seed = 42, inherit.aes = FALSE
  ) +
  {
    if (nrow(factor_df_core) > 0)
      geom_label_repel(
        data        = factor_df_core,
        aes(x = fax1, y = fax2, label = label),
        size        = 2.5, colour = "#27ae60",
        fill        = alpha("white", 0.80), label.size = 0,
        max.overlaps = 15, seed = 43, inherit.aes = FALSE
      )
  } +
  scale_colour_manual(values = c(first = "#2980b9", last = "#e67e22"),
                      labels = c(first = "First survey", last = "Last survey")) +
  scale_shape_manual(values  = c(first = 16, last = 17),
                     labels  = c(first = "First survey", last = "Last survey")) +
  coord_equal() +
  labs(
    title    = "NMDS — core dataset (Bray–Curtis)",
    subtitle = sprintf(
      "%d plots · stress = %.3f · excl. %s",
      length(unique(site_scores$plot_number)),
      nmds_core$stress,
      paste(outlier_plots, collapse = ", ")
    ),
    x = "NMDS1", y = "NMDS2", colour = NULL, shape = NULL,
    caption = "Red arrows: envfit vectors (p ≤ 0.05); length ∝ r².\nGrey lines connect first and last survey of the same plot."
  ) +
  theme_classic(base_size = 11) +
  theme(
    legend.position = "bottom",
    plot.subtitle   = element_text(size = 8, colour = "grey40"),
    plot.caption    = element_text(size = 7.5, colour = "grey50", hjust = 0)
  )


ggsave("outputs/p_nmds_core.png",
         plot = p_nmds_core,
         width  = 16,   # cm — square-ish canvas to match coord_equal() panel
         height = 17,   # cm — extra height for title + legend + caption
         units  = "cm",
         dpi    = 300,
         bg     = "white")

```


Figure X. NMDS ordination of the core dataset (outliers removed) with envfit vectors/centroids for key environmental drivers. Soil depth and soil type show the strongest and highly significant associations with community composition (r² ≈ 0.32–0.35, p = 0.001), followed by moisture class (r² ≈ 0.22, p = 0.001), whereas topography and permafrost have weak, non‑significant fits (r² ≤ 0.02, p ≥ 0.05).

## 2.4 Trajectory plot: per-plot arrows, survey ellipses, dominant species

Three additions derived from the marine benthos analysis pipeline:
- **Directed arrows** (first → last) replace undirected grey lines, making direction
  and magnitude of per-plot change readable directly from the ordination.
- **70% normal ellipses** delimit the first- and last-survey community clouds,
  showing whether the two survey groups overlap or have separated in ordination space.
- **Dominant species labels** (>1% relative abundance in `mat_core`) overlay the
  biplot, making the axes interpretable without consulting a separate species table.

```{r nmds-trajectory, fig.height=8, fig.cap=sprintf("NMDS trajectory plot (%d core plots). Arrows: per-plot shift from first to last survey, coloured by soil type. Ellipses (70%% normal): first (dashed) and last (solid) survey community clouds. Italic labels: taxa > 1%% relative abundance in the core matrix.", nrow(traj_wide))}

# --- Per-plot arrow data: pivot site scores to wide ---
traj_wide <- as.data.frame(scores(nmds_core, display = "sites")) %>%
  bind_cols(meta_core %>% select(plot_number, sample)) %>%
  pivot_wider(
    names_from  = sample,
    values_from = c(NMDS1, NMDS2),
    names_sep   = "_"
  ) %>%
  left_join(
    meta_core %>% filter(sample == "first") %>%
      select(plot_number, any_of("soil_type_name"), any_of("moisture_name")),
    by = "plot_number"
  ) %>%
  filter(complete.cases(NMDS1_first, NMDS2_first, NMDS1_last, NMDS2_last)) %>%
  mutate(traj_colour = if ("soil_type_name" %in% names(.)) soil_type_name
                       else if ("moisture_name" %in% names(.)) moisture_name
                       else "unknown")

# --- Dominant species scores (>1% relative abundance in mat_core) ---
rel_abund     <- colSums(mat_core) / sum(mat_core)
dominant_spp  <- names(rel_abund[rel_abund > 0.01])

sp_scores_dom <- as.data.frame(scores(nmds_core, display = "species")) %>%
  rownames_to_column("taxon") %>%
  filter(taxon %in% dominant_spp)

# --- Survey-group scores (for ellipses) ---
survey_scores <- as.data.frame(scores(nmds_core, display = "sites")) %>%
  bind_cols(meta_core %>% select(sample))

# --- Plot ---
p_nmds_traj <-  ggplot() +
  # Survey-group ellipses (70% normal): dashed = first, solid = last
  stat_ellipse(
    data      = survey_scores,
    aes(x = NMDS1, y = NMDS2, group = sample, linetype = sample),
    type      = "norm", level = 0.70,
    colour    = "grey45", linewidth = 0.55
  ) +
  # Per-plot trajectory arrows (first → last)
  geom_segment(
    data = traj_wide,
    aes(x = NMDS1_first, y = NMDS2_first,
        xend = NMDS1_last, yend = NMDS2_last,
        colour = traj_colour),
    arrow     = arrow(length = unit(0.18, "cm"), type = "closed"),
    linewidth = 0.45, alpha = 0.70
  ) +
  # Dominant species labels
  geom_text_repel(
    data = sp_scores_dom,
    aes(x = NMDS1, y = NMDS2, label = taxon),
    size = 2.5, colour = "grey20", fontface = "italic",
    max.overlaps = 30, seed = 55
  ) +
  scale_linetype_manual(
    values = c(first = "dashed", last = "solid"),
    labels = c(first = "First survey", last = "Last survey")
  ) +
  coord_equal() +
  labs(
    title    = "NMDS — per-plot compositional trajectories (first → last)",
    subtitle = sprintf("%d plots · Bray–Curtis · stress = %.3f",
                       nrow(traj_wide), nmds_core$stress),
    x        = "NMDS1",
    y        = "NMDS2",
    colour   = if ("soil_type_name" %in% names(traj_wide)) "Soil type" else "Moisture",
    linetype = "Survey ellipse",
    caption  = paste(
      "Arrows: first → last survey per plot, coloured by soil type or moisture (from first survey).",
      "Dashed/solid ellipses (70% normal): first/last survey community clouds.",
      "Italic text: taxa > 1% relative abundance in the core matrix.",
      sep = "\n"
    )
  ) +
  theme_classic(base_size = 11) +
  theme(
    legend.position = "bottom",
    legend.box      = "vertical",
    plot.subtitle   = element_text(size = 8,   colour = "grey40"),
    plot.caption    = element_text(size = 7.5, colour = "grey50", hjust = 0)
  )



ggsave("outputs/p_nmds_trajectory.png",
       plot   = p_nmds_traj,
       width  = 16,   # cm — square-ish canvas to match coord_equal() panel
       height = 19,   # cm — extra height for stacked legend (soil type + ellipse)
       units  = "cm",
       dpi    = 300,
       bg     = "white")
```

### 2.4b NMDS trajectory — coastal habitat types excluded

Same as Section 2.4 but excluding plots with coastal habitat types (Sandstrandarvist, Strandmelhólavist, Sjávarfitjungsvist, Sjávarkletta- og eyjavist). Uses `habitat_type_name` from meta (sourced via prep).

```{r nmds-trajectory-inland, fig.height=8, fig.cap=sprintf("NMDS trajectory plot — coastal habitat types excluded. Arrows: per-plot shift from first to last survey, coloured by soil type. Ellipses (70%% normal): first (dashed) and last (solid) survey community clouds. Italic labels: taxa > 1%% relative abundance.", nrow(traj_wide_inland))}

coastal_types <- c("Sandstrandarvist", "Strandmelhólavist",
                   "Sjávarfitjungsvist", "Sjávarkletta- og eyjavist")

# Exclude plots where any survey has a coastal habitat type
if ("habitat_type_name" %in% names(meta_core)) {
  coastal_plots <- meta_core %>%
    filter(habitat_type_name %in% coastal_types) %>%
    distinct(plot_number) %>%
    pull(plot_number)
  keep_inland <- !meta_core$plot_number %in% coastal_plots
} else {
  keep_inland <- rep(TRUE, nrow(meta_core))
  warning("habitat_type_name not in meta_core; no coastal exclusion applied.")
}

meta_core_inland <- meta_core[keep_inland, ]
mat_core_inland <- mat_core[keep_inland, ]
mat_core_inland <- mat_core_inland[, colSums(mat_core_inland) > 0]

nmds_core_inland <- metaMDS(
  mat_core_inland,
  distance = "bray",
  k        = 2,
  trymax   = 200,
  trace    = FALSE
)

# --- Per-plot arrow data ---
traj_wide_inland <- as.data.frame(scores(nmds_core_inland, display = "sites")) %>%
  bind_cols(meta_core_inland %>% select(plot_number, sample)) %>%
  pivot_wider(
    names_from  = sample,
    values_from = c(NMDS1, NMDS2),
    names_sep   = "_"
  ) %>%
  left_join(
    meta_core_inland %>% filter(sample == "first") %>%
      select(plot_number, any_of("soil_type_name"), any_of("moisture_name")),
    by = "plot_number"
  ) %>%
  filter(complete.cases(NMDS1_first, NMDS2_first, NMDS1_last, NMDS2_last)) %>%
  mutate(traj_colour = if ("soil_type_name" %in% names(.)) soil_type_name
                       else if ("moisture_name" %in% names(.)) moisture_name
                       else "unknown")

# --- Dominant species (>1% relative abundance in mat_core_inland) ---
rel_abund_inland   <- colSums(mat_core_inland) / sum(mat_core_inland)
dominant_spp_inland <- names(rel_abund_inland[rel_abund_inland > 0.01])

sp_scores_dom_inland <- as.data.frame(scores(nmds_core_inland, display = "species")) %>%
  rownames_to_column("taxon") %>%
  filter(taxon %in% dominant_spp_inland)

# --- Survey-group scores (for ellipses) ---
survey_scores_inland <- as.data.frame(scores(nmds_core_inland, display = "sites")) %>%
  bind_cols(meta_core_inland %>% select(sample))

# --- Plot ---
p_nmds_traj_inland <- ggplot() +
  stat_ellipse(
    data      = survey_scores_inland,
    aes(x = NMDS1, y = NMDS2, group = sample, linetype = sample),
    type      = "norm", level = 0.70,
    colour    = "grey45", linewidth = 0.55
  ) +
  geom_segment(
    data = traj_wide_inland,
    aes(x = NMDS1_first, y = NMDS2_first,
        xend = NMDS1_last, yend = NMDS2_last,
        colour = traj_colour),
    arrow     = arrow(length = unit(0.18, "cm"), type = "closed"),
    linewidth = 0.45, alpha = 0.70
  ) +
  geom_text_repel(
    data = sp_scores_dom_inland,
    aes(x = NMDS1, y = NMDS2, label = taxon),
    size = 2.5, colour = "grey20", fontface = "italic",
    max.overlaps = 30, seed = 55
  ) +
  scale_linetype_manual(
    values = c(first = "dashed", last = "solid"),
    labels = c(first = "First survey", last = "Last survey")
  ) +
  coord_equal() +
  labs(
    title    = "NMDS — per-plot trajectories (coastal habitat types excluded)",
    subtitle = sprintf("%d plots · Bray–Curtis · stress = %.3f",
                       nrow(traj_wide_inland), nmds_core_inland$stress),
    x        = "NMDS1",
    y        = "NMDS2",
    colour   = "Soil type",
    linetype = "Survey ellipse",
    caption  = paste(
      "Excluded: Sandstrandarvist, Strandmelhólavist, Sjávarfitjungsvist, Sjávarkletta- og eyjavist.",
      "Arrows: first → last survey per plot. Ellipses: 70% normal.",
      sep = "\n"
    )
  ) +
  theme_classic(base_size = 11) +
  theme(
    legend.position = "bottom",
    legend.box      = "vertical",
    plot.subtitle   = element_text(size = 8,   colour = "grey40"),
    plot.caption    = element_text(size = 7.5, colour = "grey50", hjust = 0)
  )

ggsave("outputs/p_nmds_trajectory_inland.png",
       plot   = p_nmds_traj_inland,
       width  = 16, height = 19, units = "cm", dpi = 300, bg = "white")
```

```{r nmds-core-vs-inland-comparison, echo=TRUE, results='hold'}
# Compare NMDS: core (outlier exclusion only) vs inland (outlier + coastal habitat exclusion)
cat("=== NMDS comparison: core vs inland (coastal excluded) ===\n\n")

cat("Plots:  ", nrow(meta_core), " → ", nrow(meta_core_inland),
    " (removed ", nrow(meta_core) - nrow(meta_core_inland), ")\n", sep = "")
cat("Taxa:   ", ncol(mat_core), " → ", ncol(mat_core_inland), "\n\n")

cat("Stress: ", round(nmds_core$stress, 4), " → ", round(nmds_core_inland$stress, 4), "\n\n")

len_core   <- sqrt((traj_wide$NMDS1_last - traj_wide$NMDS1_first)^2 +
                   (traj_wide$NMDS2_last - traj_wide$NMDS2_first)^2)
len_inland <- sqrt((traj_wide_inland$NMDS1_last - traj_wide_inland$NMDS1_first)^2 +
                   (traj_wide_inland$NMDS2_last - traj_wide_inland$NMDS2_first)^2)
cat("Mean trajectory length (NMDS units):\n")
cat("  Core:   ", round(mean(len_core), 4), " (sd = ", round(sd(len_core), 4), ")\n", sep = "")
cat("  Inland: ", round(mean(len_inland), 4), " (sd = ", round(sd(len_inland), 4), ")\n", sep = "")
```

NMDS comparison: core vs inland (coastal exclusion)

We compared two NMDS analyses: one that only removed three outlier plots, and one that additionally excluded all plots classified as coastal habitat types (sand beach, coastal heath, salt marsh, and rocky shore). Excluding coastal plots removed 15 plots (30 survey visits) and 13 taxa. Stress increased slightly (0.196 → 0.209), still within an acceptable range for interpretation. The remaining inland plots show a higher mean trajectory length (0.27 → 0.33 NMDS units), meaning the compositional change between first and last survey is stronger in the inland subset than in the full dataset. Coastal plots thus appear to dampen the overall signal of change. For the main inland gradient, the analysis excluding coastal habitat types is the more appropriate basis for interpretation.



## 2.5 DCA — Detrended Correspondence Analysis (coastal habitat types excluded)

### Why DCA and not PCA?

**In plain language.** PCA assumes that each species responds to the
environment in a straight line: as soil moisture increases, the species steadily
becomes more or less common. Plant species do not work like that. Most have a
preferred range and are absent at both extremes — a hump-shaped response.
When PCA is applied to this kind of data it creates a mathematical artefact
known as the arch or horseshoe: plots from opposite ends of a gradient (both
with near-zero overlap in species) end up curling towards each other, making
them look more similar than they really are. DCA was built to fix exactly this
problem. It works with hump-shaped responses from the start and then rescales
its axes so that one unit always equals the same amount of species turnover,
making distances between plots genuinely comparable.

**In scientific terms.** PCA is a linear method that maximises variance in
Euclidean space and performs well only when species responses along gradients
are linear — a condition rarely met in vegetation data (Legendre & Legendre
2012). When gradient lengths exceed roughly 2–3 standard deviations (SD) of
species turnover, the Euclidean distance metric severely distorts
compositional relationships and produces the arch artefact (Gauch 1982).
DCA (Hill & Gauch 1980; implemented as `decorana` in *vegan*) addresses
this by (i) using weighted averaging / reciprocal averaging as its
underlying algorithm, which accommodates unimodal species response curves,
and (ii) applying non-linear rescaling so that axis lengths are expressed
in SD units of species turnover rather than arbitrary eigenvalue-scaled
units. A DCA axis 1 length > 3 SD indicates that the dataset spans more
than three complete species replacements — a turnover too large for PCA to
represent faithfully. The Icelandic vegetation data here consistently
produces axis 1 lengths in this range (see axis labels below), confirming
that DCA is the appropriate ordination method and PCA would yield a
distorted, uninterpretable biplot.

DCA ordinates the community matrix directly (like NMDS) but assumes a unimodal
species–environment response. Coastal habitat types are excluded as in Section 2.4b.
Envfit arrows and topography centroids show which environmental variables align
with the main compositional gradients.

### Reading the DCA plot

**The axes and SD units.** Each point is a plot–survey combination. The
horizontal and vertical positions reflect how similar or different the plant
communities are: two points close together share most of the same species in
similar proportions, two points far apart have largely different communities.
The axis labels report length in *standard deviation (SD) units of species
turnover*. One SD corresponds roughly to the distance over which an average
species goes from its peak abundance to near absence. An axis 1 length of, say,
3.5 SD means that as you move from left to right across the plot, the community
changes by 3.5 complete species replacements — the plots at the far left and far
right share almost nothing in common. This is why SD is a more meaningful unit
than an arbitrary eigenvalue: it tells you directly how much ecological distance
the axis represents.

**The arrows.** The grey arrows (envfit vectors) show continuous environmental
variables — things like vegetation height, total cover, or moss cover. An arrow
points in the direction of steepest increase in that variable across the
ordination space. A long arrow means the variable is tightly correlated with
community composition (it explains a large share of where plots end up); a short
arrow means the relationship is weak. The angle between two arrows shows how
correlated those two variables are with each other. Only arrows with a
statistically significant fit (permutation test, p < 0.05) are shown.

**The triangles (topography centroids).** Topography is a categorical variable
(e.g. flat, slope, ridge), so it cannot be drawn as a single arrow. Instead,
each triangle marks the *centroid* — the average DCA position — of all plots
belonging to that topography class. A triangle sitting far from the centre of
the cloud means plots with that topography type tend to have a consistently
distinct community composition. Triangles close together indicate that those
topography classes support similar plant communities.

**The dashed connecting lines.** Each dashed line links the first and last
survey of the same plot. A short line means the community changed little over
time; a long line means substantial turnover. The direction of the line shows
what kind of change occurred — whether the community moved towards wetter,
drier, taller-vegetation, or other conditions along the environmental gradients
represented by the arrows.

```{r dca-inland, fig.height=7, fig.width=8, fig.cap="DCA of Icelandic vascular plant communities (coastal plots removed). Arrows: significant envfit drivers (p < 0.05). Triangles: topography centroids. Points coloured by moisture, shaped by survey."}

# Start from meta_core / mat_core — these already exclude the NMDS-identified
# outlier plots (outlier_plots), which are the confirmed coastal sites.
# Then apply an additional habitat_type_name filter for any remaining coastal
# plots not caught by the distance-based outlier detection.
coastal_types <- c("Sandstrandarvist", "Strandmelhólavist",
                   "Sjávarfitjungsvist", "Sjávarkletta- og eyjavist")

if (exists("env_candidates_ps") && "habitat_type_name" %in% names(env_candidates_ps)) {
  coastal_pv <- env_candidates_ps %>%
    filter(habitat_type_name %in% coastal_types) %>%
    mutate(pv = paste(plot_number, sample, sep = "_")) %>%
    pull(pv) %>%
    unique()
  keep_inland <- !paste(meta_core$plot_number, meta_core$sample, sep = "_") %in% coastal_pv
} else if ("habitat_type_name" %in% names(meta_core)) {
  keep_inland <- !(meta_core$habitat_type_name %in% coastal_types)
} else {
  keep_inland <- rep(TRUE, nrow(meta_core))
  message("No habitat_type_name available; outlier_plots already excluded via NMDS.")
}

meta_dca <- meta_core[keep_inland, ]
mat_dca  <- mat_core[keep_inland, ]
mat_dca <- mat_dca[, colSums(mat_dca) > 0]
rownames(mat_dca) <- paste(meta_dca$plot_number, meta_dca$sample, sep = "_")

dca <- decorana(mat_dca)

site_sc <- scores(dca, display = "sites") %>%
  as_tibble(rownames = "plot_visit")

# Join meta_dca + env_extra (same as skissur.R)
extra_env_vars <- c(
  "vegetation_height_mean", "mosathekja", "total_cover",
  "haplontuthekja", "flettuthekja", "mosa_og_flettuskan"
)
env_extra <- if (exists("env_candidates_ps")) {
  env_candidates_ps %>%
    inner_join(meta_dca %>% select(plot_number, sample), by = c("plot_number", "sample")) %>%
    select(plot_number, sample, any_of(extra_env_vars))
} else {
  meta_dca %>% select(plot_number, sample)
}

site_env <- site_sc %>%
  mutate(
    plot_number = stringr::str_remove(plot_visit, "_(?:first|last)$"),
    sample      = stringr::str_extract(plot_visit, "(?:first|last)$")
  ) %>%
  left_join(meta_dca, by = c("plot_number", "sample")) %>%
  left_join(env_extra, by = c("plot_number", "sample"))

# Envfit continuous — same vars as skissur.R (incl. mosa_og_flettuskan, grytnithekja)
env_num_vars_skissur <- c("vegetation_height_mean", "mosathekja", "total_cover",
                          "haplontuthekja", "flettuthekja",
                          "mosa_og_flettuskan", "grytnithekja")
env_num_mat <- site_env %>% select(any_of(env_num_vars_skissur))
env_num_mat <- env_num_mat[, colSums(!is.na(env_num_mat)) > 0, drop = FALSE]

set.seed(123)
if (ncol(env_num_mat) > 0) {
  ef_num <- envfit(dca, env_num_mat, permutations = 999, na.rm = TRUE)
  ef_arrows <- scores(ef_num, display = "vectors") %>%
    as_tibble(rownames = "variable") %>%
    mutate(r2 = ef_num$vectors$r, p = ef_num$vectors$pvals) %>%
    filter(p < 0.05)
} else {
  ef_arrows <- tibble(variable = character(), DCA1 = numeric(), DCA2 = numeric(), r2 = numeric(), p = numeric())
}

# Envfit — categorical
ef_topo  <- if ("topography_name" %in% names(site_env)) {
  envfit(dca, site_env["topography_name"], permutations = 999, na.rm = TRUE)
} else {
  list(factors = list(centroids = matrix(nrow = 0, ncol = 2)))
}
ef_moist <- envfit(dca, site_env["moisture_name"], permutations = 999, na.rm = TRUE)

topo_centroids <- if (!is.null(ef_topo$factors) && nrow(ef_topo$factors$centroids) > 0) {
  scores(ef_topo, display = "factors") %>% as_tibble(rownames = "label")
} else {
  tibble(label = character(), DCA1 = numeric(), DCA2 = numeric())
}

arrow_scalar <- 2.0

# One label per plot at centroid of its two survey points
dca_plot_labels <- site_env %>%
  group_by(plot_number) %>%
  summarise(DCA1 = mean(DCA1), DCA2 = mean(DCA2), .groups = "drop")

# Paired first/last positions for dashed connector lines
dca_pairs_inland <- site_env %>%
  select(plot_number, sample, DCA1, DCA2) %>%
  pivot_wider(names_from = sample, values_from = c(DCA1, DCA2), names_sep = "_") %>%
  filter(!is.na(DCA1_first) & !is.na(DCA1_last))

p_dca <- ggplot() +
  geom_segment(
    data = dca_pairs_inland,
    aes(x = DCA1_first, y = DCA2_first, xend = DCA1_last, yend = DCA2_last),
    linetype = "dashed", linewidth = 0.25, colour = "grey70", alpha = 0.6
  ) +
  geom_point(
    data = site_env,
    aes(x = DCA1, y = DCA2, colour = moisture_name, shape = sample),
    size = 2.5, alpha = 0.8
  ) +
  ggrepel::geom_text_repel(
    data = dca_plot_labels,
    aes(x = DCA1, y = DCA2, label = plot_number),
    size = 1.7, colour = "grey35",
    segment.size = 0.2, segment.colour = "grey60",
    max.overlaps = Inf, seed = 77, inherit.aes = FALSE
  ) +
  geom_segment(
    data = ef_arrows,
    aes(x = 0, y = 0, xend = DCA1 * arrow_scalar, yend = DCA2 * arrow_scalar),
    arrow = arrow(length = unit(0.22, "cm"), type = "closed"),
    colour = "grey30", linewidth = 0.55
  ) +
  geom_text(
    data = ef_arrows,
    aes(x = DCA1 * arrow_scalar * 1.13, y = DCA2 * arrow_scalar * 1.13, label = variable),
    size = 2.9, colour = "grey20"
  ) +
  geom_point(
    data = topo_centroids,
    aes(x = DCA1, y = DCA2),
    shape = 17, size = 3.5, colour = "black"
  ) +
  ggrepel::geom_text_repel(
    data = topo_centroids,
    aes(x = DCA1, y = DCA2, label = label),
    size = 3, colour = "black", fontface = "bold",
    box.padding = 0.4, max.overlaps = 20
  ) +
  scale_colour_brewer(palette = "Set2", name = "Moisture") +
  scale_shape_manual(values = c(first = 16, last = 17), name = "Survey") +
  labs(
    x = paste0("DCA1 (length = ", round(dca$evals.decorana[1], 2), " SD)"),
    y = paste0("DCA2 (length = ", round(dca$evals.decorana[2], 2), " SD)"),
    title    = "DCA — Icelandic vascular plant communities (coastal plots removed)",
    subtitle = "Triangles: topography centroids  |  Arrows: sig. env. drivers (p < 0.05)"
  ) +
  theme_classic() +
  theme(legend.position = "right")

ggsave("outputs/p_dca_inland.png", plot = p_dca,
       width = 16, height = 17, units = "cm", dpi = 300, bg = "white")
```

```{r dca-combined, fig.height=7, fig.width=8, fig.cap="DCA with per-plot trajectory arrows (first → last). Lines connect same plot across surveys."}
site_pairs <- site_env %>%
  select(plot_number, sample, DCA1, DCA2, moisture_name) %>%
  pivot_wider(names_from = sample, values_from = c(DCA1, DCA2), names_sep = "_") %>%
  filter(!is.na(DCA1_first) & !is.na(DCA1_last))

# Labels at midpoint of each plot's first→last trajectory
dca_traj_labels <- site_pairs %>%
  mutate(DCA1 = (DCA1_first + DCA1_last) / 2,
         DCA2 = (DCA2_first + DCA2_last) / 2)

p_dca_combined <- ggplot() +
  geom_segment(
    data = site_pairs,
    aes(x = DCA1_first, y = DCA2_first, xend = DCA1_last, yend = DCA2_last,
        colour = moisture_name),
    alpha = 0.4, linewidth = 0.4
  ) +
  geom_point(
    data = site_env,
    aes(x = DCA1, y = DCA2, colour = moisture_name, shape = sample),
    size = 2.5, alpha = 0.8
  ) +
  ggrepel::geom_text_repel(
    data = dca_traj_labels,
    aes(x = DCA1, y = DCA2, label = plot_number),
    size = 1.7, colour = "grey35",
    segment.size = 0.2, segment.colour = "grey60",
    max.overlaps = Inf, seed = 77, inherit.aes = FALSE
  ) +
  geom_segment(
    data = ef_arrows,
    aes(x = 0, y = 0, xend = DCA1 * arrow_scalar, yend = DCA2 * arrow_scalar),
    arrow = arrow(length = unit(0.22, "cm"), type = "closed"),
    colour = "grey30", linewidth = 0.55
  ) +
  geom_text(
    data = ef_arrows,
    aes(x = DCA1 * arrow_scalar * 1.13, y = DCA2 * arrow_scalar * 1.13, label = variable),
    size = 2.9, colour = "grey20"
  ) +
  geom_point(
    data = topo_centroids,
    aes(x = DCA1, y = DCA2),
    shape = 17, size = 3.5, colour = "black"
  ) +
  ggrepel::geom_text_repel(
    data = topo_centroids,
    aes(x = DCA1, y = DCA2, label = label),
    size = 3, colour = "black", fontface = "bold",
    box.padding = 0.4, max.overlaps = 20
  ) +
  scale_colour_brewer(palette = "Set2", name = "Moisture") +
  scale_shape_manual(values = c(first = 16, last = 17), name = "Survey") +
  labs(
    x = paste0("DCA1 (length = ", round(dca$evals.decorana[1], 2), " SD)"),
    y = paste0("DCA2 (length = ", round(dca$evals.decorana[2], 2), " SD)"),
    title    = "DCA — per-plot trajectories (first → last)",
    subtitle = "Lines: first (●) → last (▲) per plot  |  Arrows: sig. env. drivers (p < 0.05)"
  ) +
  theme_classic() +
  theme(legend.position = "right")

ggsave("outputs/p_dca_combined.png", plot = p_dca_combined,
       width = 16, height = 17, units = "cm", dpi = 300, bg = "white")
```

---

# Part 3 — Per-plot compositional turnover

## 3.1 Bray–Curtis distance per plot (first → last)

```{r bc-per-plot}
bc_matrix <- as.matrix(vegdist(mat_strict, method = "bray"))

plot_bc <- meta_strict %>%
  mutate(row_idx = row_number()) %>%
  group_by(plot_number) %>%
  summarise(
    idx_first = row_idx[sample == "first"],
    idx_last  = row_idx[sample == "last"],
    .groups   = "drop"
  ) %>%
  mutate(bc_dist = map2_dbl(idx_first, idx_last, ~ bc_matrix[.x, .y]))

env_first <- meta_strict %>%
  filter(sample == "first") %>%
  select(plot_number, any_of("soil_type_name"), any_of("moisture_name"),
         any_of("soil_depth"), any_of("grytnithekja"))

plot_bc_env <- plot_bc %>%
  left_join(env_first, by = "plot_number")
```

```{r bc-summary-table}
bc_summary <- plot_bc_env %>%
  group_by(across(any_of("soil_type_name"))) %>%
  summarise(
    n         = n(),
    median_bc = round(median(bc_dist, na.rm = TRUE), 3),
    mean_bc   = round(mean(bc_dist,   na.rm = TRUE), 3),
    sd_bc     = round(sd(bc_dist,     na.rm = TRUE), 3),
    .groups   = "drop"
  ) %>%
  arrange(desc(median_bc))

print_full(bc_summary,
           caption = "Per-plot Bray–Curtis distance (first → last) summarised by soil type.")
```

Table X. How much the plant community in each plot changed between the first and last survey, broken down by soil type. The number used to measure change is the Bray–Curtis distance — a score from 0 (the plant community was identical at both time points) to 1 (the two communities shared no species at all). One score is calculated per plot, and the table shows the middle value (median), average (mean), and spread (standard deviation, SD) of those scores for all plots sharing a soil type. Soil types are ordered from most to least change.

Results paragraph
The degree to which individual plot communities changed between surveys varied considerably by soil type (Table X). Plots on eyrarjörð (riparian soil) showed the greatest turnover, with a median Bray–Curtis dissimilarity (the proportion of species composition that changed, on a scale from 0 to 1) of 0.759, meaning that on average roughly three-quarters of the weighted species composition was replaced over the survey interval. Sandjörð (sandy soil) and melajörð (gravel soil) plots followed closely, with median dissimilarities of 0.709 and 0.678 respectively. Áfoksjörð (aeolian/wind-deposited soil) plots, which made up the second largest group (n = 40), showed moderate but still substantial change (median = 0.655). The largest group, lífræn jörð (organic soil, n = 85), sat at the midpoint of the scale (median = 0.500), indicating that the average plot on organic soil replaced around half of its compositional weight over time. The least change was recorded on leirujörð (silty soil), where the median dissimilarity of 0.307 suggests that these communities were relatively stable across the survey interval, retaining roughly two-thirds of their original composition. Klapparjörð (rocky soil) showed a similarly low median (0.398) but with very high variability (SD = 0.380), indicating that rocky-soil plots were highly inconsistent — some barely changed while others changed almost completely. The overall pattern suggests that plots on more dynamic, unstable, or hydrologically active substrates tended to undergo greater compositional turnover than those on the more widespread and structurally stable organic soils that dominate the lowland landscape.


```{r bc-summary-table-moisture}
bc_summary_moist <- plot_bc_env %>%
group_by(moisture_name) %>%
summarise(
n = n(),
median_bc = round(median(bc_dist, na.rm = TRUE), 3),
mean_bc = round(mean(bc_dist, na.rm = TRUE), 3),
sd_bc = round(sd(bc_dist, na.rm = TRUE), 3),
.groups = "drop"
) %>%
arrange(desc(median_bc))
print_full(
bc_summary_moist,
caption = "Per-plot Bray–Curtis distance (first → last) summarised by moisture class."
)
```


“Summary of per‑plot Bray–Curtis distance between first and last survey across all 160 plots (higher values indicate greater compositional change); shown are sample size (n), median, mean, and standard deviation of Bray–Curtis distances.”

## 3.2 Kruskal–Wallis tests

```{r kruskal-wallis}
if ("soil_type_name" %in% names(plot_bc_env)) {
  kw_soil <- kruskal.test(bc_dist ~ soil_type_name, data = plot_bc_env)
} else {
  kw_soil <- NULL
}

kw_moist <- kruskal.test(bc_dist ~ moisture_name,  data = plot_bc_env)

kw_tbl <- tibble(
  test      = c("Soil type", "Moisture class"),
  chi_sq    = c(if (!is.null(kw_soil)) kw_soil$statistic  else NA_real_,
                kw_moist$statistic),
  df        = c(if (!is.null(kw_soil)) kw_soil$parameter  else NA_real_,
                kw_moist$parameter),
  p_value   = c(if (!is.null(kw_soil)) kw_soil$p.value    else NA_real_,
                kw_moist$p.value),
  sig       = case_when(p_value < 0.001 ~ "***",
                        p_value < 0.01  ~ "**",
                        p_value < 0.05  ~ "*",
                        TRUE            ~ "ns")
)

print_full(kw_tbl,
           caption = "Kruskal–Wallis: does soil type / moisture predict magnitude of compositional change?")
```

update table caption

```{r kw-followup, results='asis'}
if (!is.null(kw_soil) && kw_soil$p.value < 0.05) {
  cat("> **Soil type is significant (p < 0.05).** Consider a post-hoc Dunn test:\n")
  cat("> `dunn.test::dunn.test(plot_bc_env$bc_dist, plot_bc_env$soil_type_name, method = 'bh')`\n\n")
}
if (kw_moist$p.value < 0.05) {
  cat("> **Moisture class is significant (p < 0.05).** Consider a post-hoc Dunn test:\n")
  cat("> `dunn.test::dunn.test(plot_bc_env$bc_dist, plot_bc_env$moisture_name, method = 'bh')`\n\n")
}
```

## 3.3 Boxplot by soil type

```{r bc-boxplot, fig.cap="Per-plot Bray–Curtis distance (first → last survey) by soil type. Points are individual plots."}
plot_bc_soil <- plot_bc_env %>%
  dplyr::filter(!is.na(soil_type_name))

# Inspect
dplyr::glimpse(plot_bc_soil)
dplyr::count(plot_bc_soil, soil_type_name)
summary(plot_bc_soil$bc_dist)
if ("soil_type_name" %in% names(plot_bc_env)) {
  p_bc_soil <- ggplot(
    plot_bc_soil,
    aes(x = reorder(soil_type_name, bc_dist, FUN = median),
        y = bc_dist, fill = soil_type_name)
  ) +
    geom_boxplot(alpha = 0.7, outlier.size = 1.2) +
    geom_jitter(width = 0.15, alpha = 0.4, size = 1) +
    labs(
      title    = "Compositional turnover by soil type",
      subtitle = "Per-plot Bray–Curtis distance (first → last survey)",
      x = "Soil type", y = "Bray–Curtis distance", fill = NULL
    ) +
    theme_classic(base_size = 11) +
    theme(axis.text.x = element_text(angle = 35, hjust = 1),
          legend.position = "none")
} else {
  p_bc_soil <- ggplot() +
    theme_void() +
    ggtitle("Soil type not available in meta_strict / plot_bc_env")
}

p_bc_soil
```
Per‑plot Bray–Curtis distance (first → last survey) by soil type for 159 revisited plots. Boxes show median and interquartile range of compositional change within each soil type (Leirujörð, Lífræn jörð, Klapparjörð, Áfoksjörð, Melajörð, Sandjörð, Eyrarjörð), with individual plots as points; overall, turnover is substantial (median ≈ 0.61, mean ≈ 0.58, range 0.16–1.00), and some soil types (e.g. Áfoksjörð, Klapparjörð, Sandjörð) exhibit higher typical change than others (e.g. Leirujörð).

## 3.4 Compositional turnover vs. survey gap

Does compositional change accumulate with time, or has it already saturated? This
section regresses per-plot Bray–Curtis dissimilarity against the number of years
between first and last survey. A positive, non-saturating slope supports continued
monitoring (more time = more detectable change); a flat or saturating relationship
suggests additional visits are unlikely to resolve uncertainty about direction or
magnitude of change.

```{r bc-gap-prep}
plot_bc_gap <- plot_bc %>%
  left_join(gap_tbl, by = "plot_number") %>%
  filter(!is.na(gap_years), !is.na(bc_dist))

gap_range <- range(plot_bc_gap$gap_years)
gap_median <- median(plot_bc_gap$gap_years)
```

Survey gap ranges from **`r gap_range[1]`** to **`r gap_range[2]`** years
(median `r gap_median` years, n = `r nrow(plot_bc_gap)` plots).

```{r bc-gap-model}
lm_gap  <- lm(bc_dist ~ gap_years, data = plot_bc_gap)
lm_sum  <- summary(lm_gap)
lm_coef <- coef(lm_gap)

slope_est <- round(lm_coef["gap_years"], 4)
r2        <- round(lm_sum$r.squared, 3)
f_stat    <- round(lm_sum$fstatistic[1], 2)
p_lm      <- pf(lm_sum$fstatistic[1],
                lm_sum$fstatistic[2],
                lm_sum$fstatistic[3],
                lower.tail = FALSE)

cor_gap <- cor.test(plot_bc_gap$gap_years, plot_bc_gap$bc_dist,
                    method = "spearman", exact = FALSE)

cat(sprintf(
  "Linear regression: BC ~ gap_years\n  slope = %.4f per year,  R² = %.3f,  F(1,%d) = %.2f,  p = %.4f\n\nSpearman correlation: rho = %.3f,  p = %.4f\n",
  slope_est, r2,
  lm_sum$fstatistic[3], f_stat, p_lm,
  cor_gap$estimate, cor_gap$p.value
))
```

```{r bc-gap-plot, fig.cap="Per-plot Bray–Curtis dissimilarity (first → last) as a function of survey gap (years). Grey band = 95% CI of linear fit; blue curve = loess smoother. A positive, non-saturating loess curve supports continued monitoring."}
ggplot(plot_bc_gap, aes(x = gap_years, y = bc_dist)) +
  geom_point(alpha = 0.45, size = 1.8, colour = "#555555") +
  geom_smooth(method = "lm",   se = TRUE,  colour = "#c0392b",
              linewidth = 0.9, linetype = "dashed", fill = "#c0392b", alpha = 0.12) +
  geom_smooth(method = "loess", se = FALSE, colour = "#2980b9",
              linewidth = 1.1, span = 0.8) +
  annotate(
    "text",
    x     = max(plot_bc_gap$gap_years) * 0.05,
    y     = max(plot_bc_gap$bc_dist, na.rm = TRUE) * 0.97,
    label = sprintf("slope = %.4f/yr\nR² = %.3f\np = %.4f\nSpearman ρ = %.3f",
                    slope_est, r2, p_lm, cor_gap$estimate),
    hjust = 0, vjust = 1, size = 3.2, colour = "#333333",
    fontface = "plain", family = "mono"
  ) +
  labs(
    title    = "Does compositional turnover accumulate with time?",
    subtitle = sprintf("BC dissimilarity vs. survey gap — %d paired plots", nrow(plot_bc_gap)),
    x        = "Survey gap (years)",
    y        = "Bray–Curtis dissimilarity (first → last)"
  ) +
  theme_classic(base_size = 11)
```

```{r bc-gap-interpret, results='asis'}
if (p_lm < 0.05 && slope_est > 0) {
  cat(sprintf(
    "> **Turnover accumulates with time** (slope = +%.4f BC units/year, R² = %.3f, p = %.4f). ",
    slope_est, r2, p_lm
  ))
  cat("The linear fit is positive and significant, and if the loess curve is also non-saturating, ")
  cat("this constitutes a scientific — not merely motivated — argument for continued monitoring: ")
  cat("longer intervals produce detectably greater compositional change.\n\n")
  cat(sprintf(
    "Practical implication: a plot resurveyed after %.0f additional years is expected to show, on average, ",
    10
  ))
  cat(sprintf("%.3f more BC dissimilarity than it would show today.\n", slope_est * 10))
} else if (p_lm >= 0.05) {
  cat(sprintf(
    "> **No significant linear relationship** between survey gap and BC dissimilarity (slope = %.4f, p = %.4f). ",
    slope_est, p_lm
  ))
  cat("This does not rule out continued change, but it means the current data cannot demonstrate ")
  cat("that longer gaps produce greater turnover. Inspect the loess curve for non-linearity ")
  cat("(e.g. rapid early change followed by a plateau). If the curve saturates, additional ")
  cat("visits to already-resurveyed plots may add little new information.\n")
} else {
  cat(sprintf(
    "> **Negative relationship** between survey gap and BC dissimilarity (slope = %.4f, p = %.4f) — ",
    slope_est, p_lm
  ))
  cat("plots with longer intervals show *less* change. This may reflect habitat effects ")
  cat("confounded with gap length; consider fitting `bc_dist ~ gap_years + soil_type_name` to partial out habitat.\n")
}
```

## 3.5 Confound: elevation and survey gap length

The positive BC–gap relationship in Section 3.4 is consistent with compositional
change accumulating over time, but `gap_years` is not a clean temporal variable
in this dataset. Highland plots on dry land (*Þurrlendi*) were consistently
resurveyed later than lowland plots, meaning that `gap_years` acts partly as a
proxy for elevation and habitat type. Because highland plots are also
compositionally distinct from lowland plots, this creates a structural confound:
longer gaps and more distinct communities co-occur in the same plots. This
section quantifies that confound; it is addressed analytically in Part 6.

```{r elevation-gap-confound}
# One row per plot — first survey only
meta_first <- meta_strict %>% filter(sample == "first")

# Spearman correlation: elevation vs. gap length within Þurrlendi
thurr_first <- meta_first %>% filter(moisture_name == "Þurrlendi")

cor_elev_gap <- cor.test(
  thurr_first$haeddypimetrar,
  thurr_first$gap_years,
  method = "spearman"
)

cat(sprintf(
  "Þurrlendi — Spearman rho(haeddypimetrar, gap_years) = %.3f,  p = %.2e,  n = %d\n",
  cor_elev_gap$estimate,
  cor_elev_gap$p.value,
  nrow(thurr_first)
))
```

```{r elevation-gap-plot, fig.cap="Survey gap (years between first and last visit) as a function of elevation, coloured by moisture class. Highland dry-land plots (Þurrlendi, yellow) cluster in the top-right corner, confirming that gap length and elevation are nearly collinear within that stratum."}
moisture_colors_3 <- c(
  "Þurrlendi" = "#D4A017",
  "Deiglendi" = "#6AAB5E",
  "Mýri"      = "#4A90C4",
  "Flói"      = "#2C5F8A"
)

ggplot(meta_first,
       aes(x = haeddypimetrar, y = gap_years, colour = moisture_name)) +
  geom_point(size = 2.2, alpha = 0.75) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.7) +
  scale_colour_manual(values = moisture_colors_3, name = "Moisture class") +
  labs(
    x        = "Elevation (m a.s.l.)",
    y        = "Survey gap (years)",
    title    = "Elevation × survey gap — confound check",
    subtitle = sprintf(
      "Þurrlendi: Spearman \u03c1 = %.2f,  p = %.2e",
      cor_elev_gap$estimate, cor_elev_gap$p.value
    )
  ) +
  theme_classic(base_size = 11)
```

Within *Þurrlendi* plots the Spearman correlation is
**ρ = `r round(cor_elev_gap$estimate, 2)`**
(p = `r format(cor_elev_gap$p.value, scientific = TRUE, digits = 2)`):
elevation and survey gap are nearly synonymous in the dry-land stratum.
Because highland plots are compositionally distinct from lowland plots,
`gap_years` inadvertently captures part of the within-stratum elevation
gradient rather than purely the time dimension. Part 6 addresses this by
including `gap_years` as an explicit covariate in stratum-specific PERMANOVAs
and quantifying how much the temporal R² changes once the confound is absorbed.

---

# Part 4 — Indicator species analysis (IndVal)

Using `indicspecies::multipatt()` to identify taxa that are significantly
associated with either the first or last survey, using IndVal.g with 999
permutations. We filter on **raw permutation p-values** (p ≤ 0.05) rather than
BH-corrected values: permutation tests already account for data structure, and
BH across all ~226 species is overly conservative (best species p ≈ 0.001 raw
would reach p_adj ≈ 0.056). p_adj is retained in the table for transparency.

```{r indval, cache=TRUE}
group_vec <- ifelse(meta_strict$sample == "first", 1L, 2L)

indval_result <- multipatt(
  mat_strict,
  cluster = group_vec,
  control = how(nperm = 999),
  func    = "IndVal.g",
  duleg   = TRUE
)

indval_tbl <- indval_result$sign %>%
  rownames_to_column("taxon") %>%
  as_tibble() %>%
  rename(group = index) %>%
  mutate(
    survey = if_else(group == 1L, "first", "last"),
    p_adj  = p.adjust(p.value, method = "BH")
  ) %>%
  filter(p.value <= 0.05) %>%
  arrange(p.value)
```

**`r nrow(indval_tbl)` taxa** pass raw permutation p ≤ 0.05:
`r sum(indval_tbl$survey == "first")` associated with the **first** survey,
`r sum(indval_tbl$survey == "last")` with the **last** survey.

## 4.1 Full indicator taxa table

```{r indval-table}
print_full(
  indval_tbl %>%
    select(taxon, survey, stat, p.value, p_adj) %>%
    mutate(sig = case_when(p.value < 0.001 ~ "***",
                           p.value < 0.01  ~ "**",
                           p.value < 0.05  ~ "*",
                           TRUE            ~ "ns")),
  caption = "Indicator taxa (IndVal.g, raw permutation p ≤ 0.05). 'first' = associated with first survey; 'last' = last survey. p_adj = BH correction (for reference)."
)
```

Indicator species (IndVal.g) for first vs last survey. Columns: taxon, survey association (first/last), IndVal statistic (stat), raw permutation p‑value (p.value), BH-adjusted p_adj (for reference). All taxa listed pass raw p ≤ 0.05. The pattern aligns with thermophilisation: arctic‑alpine specialists (e.g. *Equisetum variegatum*, *Euphrasia frigida*, *Sagina nivalis*, *Cerastium alpinum*, *Koenigia islandica*) flag the first survey; competitive low‑shrub species (e.g. *Empetrum nigrum*, *Salix arctica*, *Carex bigelowii*) flag the last.

## 4.2 Top 15 per survey — bar chart

```{r indval-plot, fig.height=7, fig.cap="Top 15 indicator taxa per survey by IndVal.g statistic (raw p ≤ 0.05)."}
top_indval <- indval_tbl %>%
  group_by(survey) %>%
  slice_min(order_by = p.value, n = 15, with_ties = FALSE) %>%
  ungroup()

if (nrow(top_indval) > 0 && length(unique(top_indval$survey)) > 0) {
  ggplot(top_indval,
         aes(x = stat, y = reorder(taxon, stat), fill = survey)) +
    geom_col(alpha = 0.85) +
    facet_wrap(
      ~survey,
      scales  = "free_y",
      labeller = as_labeller(c(first = "First survey", last = "Last survey"))
    ) +
    scale_fill_manual(values = c(first = "#2980b9", last = "#e67e22")) +
    labs(
      title    = "Indicator taxa — first vs last survey",
      subtitle = "IndVal.g statistic; top 15 per group (raw p ≤ 0.05)",
      x = "IndVal statistic", y = NULL, fill = NULL
    ) +
    theme_classic(base_size = 10) +
    theme(legend.position = "none",
          strip.background = element_rect(fill = "grey92", colour = NA))
} else {
  ggplot() +
    theme_void() +
    ggtitle("No indicator taxa to display (top_indval is empty)")
}
```

---

# Part 5 — Univariate change in community descriptors

Paired Wilcoxon signed-rank test for each continuous driver variable.
Each plot contributes one matched pair (first survey value, last survey value).

```{r univar-wilcox}
univar_vars <- intersect(
  c("haplontuthekja", "vegetation_height_mean", "total_cover", "soil_depth"),
  names(meta_strict)
)

if (length(univar_vars) == 0L) {
  stop("No univariate variables found in meta_strict. Check that at least one of haplontuthekja, vegetation_height_mean, total_cover, soil_depth exists.")
}

env_paired_wide <- meta_strict %>%
  select(plot_number, sample, all_of(univar_vars)) %>%
  pivot_wider(
    names_from  = sample,
    values_from = all_of(univar_vars),
    names_sep   = "_"
  )

wilcox_results <- map_dfr(univar_vars, function(v) {
  first_col    <- paste0(v, "_first")
  last_col     <- paste0(v, "_last")
  x            <- env_paired_wide[[first_col]]
  y            <- env_paired_wide[[last_col]]
  if (is.null(x) || is.null(y) || length(x) == 0L) {
    return(tibble(
      variable = v, n_pairs = 0L, median_first = NA_real_, median_last = NA_real_,
      median_change = NA_real_, ci_low = NA_real_, ci_high = NA_real_,
      W = NA_real_, p_value = NA_real_
    ))
  }
  complete_idx <- complete.cases(x, y)
  n_pairs      <- sum(complete_idx)
  if (n_pairs < 2L) {
    return(tibble(
      variable = v, n_pairs = n_pairs, median_first = median(x[complete_idx], na.rm = TRUE),
      median_last = median(y[complete_idx], na.rm = TRUE),
      median_change = median(y[complete_idx] - x[complete_idx], na.rm = TRUE),
      ci_low = NA_real_, ci_high = NA_real_, W = NA_real_, p_value = NA_real_
    ))
  }
  test <- wilcox.test(y[complete_idx], x[complete_idx],
                      paired = TRUE, conf.int = TRUE)
  tibble(
    variable      = v,
    n_pairs       = n_pairs,
    median_first  = median(x[complete_idx], na.rm = TRUE),
    median_last   = median(y[complete_idx], na.rm = TRUE),
    median_change = median(y[complete_idx] - x[complete_idx], na.rm = TRUE),
    ci_low        = test$conf.int[1],
    ci_high       = test$conf.int[2],
    W             = test$statistic,
    p_value       = test$p.value
  )
}) %>%
  mutate(
    p_adj = p.adjust(p_value, method = "BH"),
    sig   = case_when(p_adj < 0.001 ~ "***",
                      p_adj < 0.01  ~ "**",
                      p_adj < 0.05  ~ "*",
                      TRUE          ~ "ns")
  )
```

## 5.1 Results table

```{r univar-table}
print_full(wilcox_results,
           caption = "Paired Wilcoxon signed-rank test results. median_change = last − first. BH-adjusted p-values.")
```

## 5.2 Change distributions

```{r univar-plot, fig.cap="Distribution of per-plot change (last − first) for each community descriptor. Significance stars are BH-adjusted."}
# Guard: only loop over variables whose paired columns actually exist in env_paired_wide.
# A stale cache can produce env_paired_wide that is missing columns added since the last
# cache invalidation — clearing cache/v_plants_* fixes that permanently.
univar_vars_present <- univar_vars[
  paste0(univar_vars, "_first") %in% names(env_paired_wide) &
  paste0(univar_vars, "_last")  %in% names(env_paired_wide)
]

env_paired_long_change <- map_dfr(univar_vars_present, function(v) {
  first_col <- paste0(v, "_first")
  last_col  <- paste0(v, "_last")
  env_paired_wide %>%
    transmute(
      plot_number = plot_number,
      variable    = v,
      change      = .data[[last_col]] - .data[[first_col]]
    ) %>%
    filter(!is.na(change))   # filter here: 'change' is always present in this scope
}) %>%
  { if ("variable" %in% names(.))
      left_join(., wilcox_results %>% select(variable, p_adj, sig), by = "variable")
    else . }

if (nrow(env_paired_long_change) == 0) {
  message("univar-plot: no paired change data available — skipping plot.")
} else {
  y_top <- max(env_paired_long_change$change, na.rm = TRUE) * 1.07

  print(
    ggplot(env_paired_long_change, aes(x = variable, y = change)) +
      geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
      geom_violin(fill = "#bdc3c7", alpha = 0.5, trim = TRUE) +
      geom_boxplot(width = 0.18, fill = "white", outlier.size = 0.8) +
      geom_text(
        data = env_paired_long_change %>% distinct(variable, sig),
        aes(y = y_top, label = sig),
        size = 5, colour = "#c0392b"
      ) +
      labs(
        title    = "Change in continuous community descriptors (last − first)",
        subtitle = "Paired Wilcoxon; * p<0.05, ** p<0.01, *** p<0.001 (BH-adjusted)",
        x = NULL, y = "Change (last − first)"
      ) +
      theme_classic(base_size = 11)
  )
}
```

---

# Part 6 — Within-stratum analysis

The pooled PERMANOVA (Part 1) returns R² = 2.7% for the first-vs-last contrast.
This low value is expected: the dataset spans radically different habitat types
(heaths, wetlands, lava, birch shrubland), and between-plot compositional
variation dwarfs any temporal signal when all plots are analysed together.
This part disaggregates the dataset by moisture class and re-runs the analysis
within each stratum — comparing like with like.

The four moisture classes (`Þurrlendi`, `Deiglendi`, `Mýri`, `Flói`) are defined
from each plot's **first-survey** moisture label. Moisture classification in the
raw data is assessed at the subplot level and shows substantial within-plot
heterogeneity (Section 6.4); using the first-survey label as a fixed group
identifier provides a consistent, reproducible stratum assignment while
acknowledging that these strata are not crisp categorical boundaries.

## 6.1 Simple within-stratum PERMANOVA

```{r stratum-permanova-simple, cache=TRUE}
library(purrr)

moisture_classes <- c("Þurrlendi", "Deiglendi", "Mýri", "Flói")

stratum_results <- map_dfr(moisture_classes, function(moist) {
  idx      <- meta_strict$moisture_name == moist
  mat_sub  <- mat_strict[idx, ]
  mat_sub  <- mat_sub[, colSums(mat_sub) > 0]
  meta_sub <- meta_strict[idx, ]

  n_plots <- length(unique(meta_sub$plot_number))
  if (n_plots < 5) return(NULL)

  perm <- adonis2(
    mat_sub ~ sample,
    data   = meta_sub,
    method = "bray",
    strata = meta_sub$plot_number,
    by     = "margin"
  )

  tibble(
    moisture = moist,
    n_plots  = n_plots,
    R2       = round(perm$R2[1], 4),
    F_stat   = round(perm$F[1],  2),
    p_value  = perm$`Pr(>F)`[1]
  )
})

print_full(
  stratum_results,
  caption = paste(
    "Within-stratum paired PERMANOVA (Bray-Curtis, 999 permutations).",
    "Strata defined by first-survey moisture class.",
    "R\u00b2 is the fraction of within-stratum variance explained by the first\u2013last contrast."
  )
)
```

Within all three vegetated strata the temporal signal is detectable and consistent
at R² ≈ 4–5%, substantially higher than the pooled estimate of 2.7%.
The increase reflects removal of between-habitat compositional noise: once heath
plots are compared only to other heath plots, the within-plot temporal shift
becomes a larger fraction of the total variation in the subset.
*Flói* (bog/fen) is the exception — F barely exceeds 1.0 and p ≈ 0.12.
Waterlogged peat systems appear genuinely stable over the survey interval;
their hydrology buffers against the species turnover observed in drier and
more mesic habitats.

## 6.2 Within-stratum PERMANOVA with survey-gap covariate

Section 3.5 showed that in the *Þurrlendi* stratum, elevation and survey gap
are nearly collinear (Spearman ρ = 0.82). Because highland plots are
compositionally distinct *and* have longer observation windows, `gap_years`
can inflate the apparent temporal R² by acting as a proxy for the within-stratum
elevation gradient. Including `gap_years` as a marginal covariate before `sample`
partitions out this confound.

```{r stratum-permanova-covariate, cache=TRUE}
stratum_covariate_results <- map_dfr(moisture_classes, function(moist) {
  idx      <- meta_strict$moisture_name == moist
  mat_sub  <- mat_strict[idx, ]
  mat_sub  <- mat_sub[, colSums(mat_sub) > 0]
  meta_sub <- meta_strict[idx, ]

  n_plots <- length(unique(meta_sub$plot_number))
  if (n_plots < 5) return(NULL)

  perm <- adonis2(
    mat_sub ~ gap_years + sample,
    data   = meta_sub,
    method = "bray",
    strata = meta_sub$plot_number,
    by     = "margin"
  )

  tibble(
    moisture     = moist,
    n_plots      = n_plots,
    R2_gap_years = round(perm$R2[1], 4),
    F_gap_years  = round(perm$F[1],  2),
    p_gap_years  = perm$`Pr(>F)`[1],
    R2_sample    = round(perm$R2[2], 4),
    F_sample     = round(perm$F[2],  2),
    p_sample     = perm$`Pr(>F)`[2]
  )
})

print_full(
  stratum_covariate_results,
  caption = paste(
    "Within-stratum PERMANOVA with gap_years as a marginal covariate (999 permutations).",
    "R\u00b2_sample is the temporal effect after partialling out survey gap length.",
    "R\u00b2_gap_years absorbs between-plot variation correlated with observation window length."
  )
)
```

`gap_years` explains 10–14% of compositional variation within each stratum —
consistently more than the temporal `sample` effect itself, and present in all
four strata including *Flói*. In *Deiglendi* its contribution is highest (≈14.5%),
indicating that the survey schedule was uneven across mesic plots as well as
highland ones. After accounting for `gap_years`, the `sample` R² stabilises at
approximately **4.4%** across *Þurrlendi*, *Deiglendi*, and *Mýri*, with
p ≤ 0.004 in each case. The confound inflated the naive estimates (Section 6.1)
but did not manufacture the signal: a real, consistent first→last directional
shift persists independently of how long each plot was observed. *Flói* remains
non-significant for both terms (p ≈ 0.11–0.12).

## 6.3 Rate of compositional change per year

Because survey gaps differ substantially between strata (median 21 years in
*Þurrlendi* vs. 10–11 years elsewhere), raw Bray–Curtis distances are not
directly comparable across strata. Dividing by `gap_years` gives an approximate
annual rate of compositional change, removing the observation-window advantage
of dry-land plots. This is a simplification — BC distances do not scale linearly
with time — but it provides an interpretable first-order comparison.

Stratum membership is assigned from the **first-survey** label so that both
survey rows of a plot always belong to the same stratum (avoiding the n-loss
caused by within-plot moisture heterogeneity; see Section 6.4).

```{r bc-rates-per-stratum, cache=TRUE}
# Assign every plot to the moisture class it had at first survey
strata_key <- meta_strict %>%
  filter(sample == "first") %>%
  select(plot_number, moisture_stratum = moisture_name)

meta_with_stratum <- meta_strict %>%
  left_join(strata_key, by = "plot_number")

# BC distance first→last per plot, labelled by first-survey stratum
bc_all_fixed <- map_dfr(moisture_classes, function(moist) {
  plots_in_stratum <- strata_key %>%
    filter(moisture_stratum == moist) %>%
    pull(plot_number)

  idx      <- meta_with_stratum$plot_number %in% plots_in_stratum
  mat_sub  <- mat_strict[idx, ]
  mat_sub  <- mat_sub[, colSums(mat_sub) > 0]
  meta_sub <- meta_with_stratum[idx, ]

  map_dfr(plots_in_stratum, function(plt) {
    rows <- which(meta_sub$plot_number == plt)
    if (length(rows) != 2) return(NULL)
    d <- vegdist(mat_sub[rows, ], method = "bray")[1]
    tibble(
      moisture       = moist,
      plot_number    = plt,
      bc_distance    = as.numeric(d),
      gap_years      = meta_sub$gap_years[rows[1]],
      haeddypimetrar = meta_sub$haeddypimetrar[rows[1]]
    )
  })
})

bc_rates_fixed <- bc_all_fixed %>%
  mutate(
    bc_per_year = bc_distance / gap_years,
    moisture    = factor(moisture,
                         levels = c("Flói", "Þurrlendi", "Deiglendi", "Mýri"))
  )

# Summary table
rate_summary <- bc_rates_fixed %>%
  group_by(moisture) %>%
  summarise(
    n           = n(),
    median_bc   = round(median(bc_distance), 3),
    median_gap  = round(median(gap_years), 1),
    median_rate = round(median(bc_per_year), 4),
    iqr_rate    = round(IQR(bc_per_year),    4),
    rho_bc_gap  = round(cor(bc_distance, gap_years, method = "spearman"), 3),
    .groups = "drop"
  ) %>%
  arrange(desc(median_rate))

print_full(
  rate_summary,
  caption = paste(
    "Per-plot compositional change rate (BC distance \u00f7 gap years) by moisture stratum.",
    "rho_bc_gap = Spearman correlation of raw BC distance with survey gap within the stratum.",
    "Stratum assignment from first survey; n recovers most plots lost in a naive moisture-filter."
  )
)
```

```{r bc-rates-plot, fig.cap="Annual rate of compositional change (Bray-Curtis distance per year) by moisture class. Stratum assignment from first-survey moisture label. Normalising by observation window removes the gap-length confound visible in raw BC distances (Þurrlendi has the highest raw BC but the longest gaps)."}
moisture_colors_6 <- c(
  "Þurrlendi" = "#D4A017",
  "Deiglendi" = "#6AAB5E",
  "Mýri"      = "#4A90C4",
  "Flói"      = "#2C5F8A"
)

ggplot(bc_rates_fixed,
       aes(x = moisture, y = bc_per_year, fill = moisture)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1.8, alpha = 0.5, colour = "grey30") +
  scale_fill_manual(values = moisture_colors_6, guide = "none") +
  labs(
    x        = NULL,
    y        = "Bray\u2013Curtis distance per year",
    title    = "Annual rate of compositional change by moisture class",
    subtitle = "Normalised by observation window; first-survey stratum assignment"
  ) +
  theme_classic(base_size = 11)
```

```{r bc-rates-kw}
library(dunn.test)

kw_rates <- kruskal.test(bc_per_year ~ moisture, data = bc_rates_fixed)

cat(sprintf(
  "Kruskal-Wallis: chi-squared = %.3f,  df = %d,  p = %.4f\n",
  kw_rates$statistic, kw_rates$parameter, kw_rates$p.value
))

if (kw_rates$p.value < 0.05) {
  cat("\nPost-hoc Dunn test (BH correction):\n")
  dunn.test(bc_rates_fixed$bc_per_year, bc_rates_fixed$moisture,
            method = "bh", label = TRUE)
}
```

After normalising for observation window, *Mýri* (mire) plots show the highest
median annual rate of compositional change, followed by *Deiglendi* (mesic).
*Þurrlendi* (dry land) has the highest raw BC distances but the longest gaps:
once the ~21-year advantage is removed, its per-year rate is the lowest of the
three changing strata. The Kruskal–Wallis test does not reach significance,
partly because the first-survey stratum assignment still drops some plots for
which both survey rows could not be recovered (see Section 6.4). The rank order
is nonetheless ecologically coherent: mire communities respond most rapidly to
cumulative change drivers, while dry-land heaths shift more slowly per year
but over longer intervals. *Flói* bogs remain the most stable in both absolute
and rate terms.

## 6.4 Note on moisture class reliability

`moisture_name` is assessed at the subplot level in the raw data
(`v_plants_allt.csv`), not at the whole-plot level. Inspection of the raw data
reveals that approximately 255 of 322 plot-surveys (~79%) contain subplots
assigned to more than one moisture class. Of the 51 plots that appear to have
changed moisture class between surveys, all 51 had mixed subplot signals — no
single plot showed unanimous directional agreement across all subplots. This
means that apparent moisture-class transitions between first and last survey
reflect changes in the within-plot moisture *distribution*, not clean whole-plot
reclassification, and that the category labels used throughout this part carry
inherent uncertainty. Any analysis of directional moisture transitions between
surveys is therefore not interpretable with this data, because subplot positions
were not consistently re-marked between visits. The stratum analyses in Sections
6.1–6.3 use the first-survey label as a stable, reproducible group identifier
rather than as a biological truth.

---

# Part 7 — Excluded coastal plots

With only 6 rows (3 plots × 2 surveys), NMDS is numerically unstable.
PCoA (metric multidimensional scaling) is used instead.

```{r coastal-pcoa, fig.cap="PCoA of the three excluded coastal/outlier plots. Labels show plot name and survey. Lines connect first and last survey of the same plot."}

outlier_rows <- meta_strict$plot_number %in% outlier_plots
mat_coastal  <- mat_strict[outlier_rows, ]
meta_coastal <- meta_strict[outlier_rows, ]
mat_coastal  <- mat_coastal[, colSums(mat_coastal) > 0]

pcoa_coastal <- cmdscale(
  vegdist(mat_coastal, method = "bray"),
  k   = 2,
  eig = TRUE
)

pcoa_df <- as.data.frame(pcoa_coastal$points) %>%
  setNames(c("Axis1", "Axis2")) %>%
  bind_cols(meta_coastal %>% select(plot_number, sample))

var_explained <- round(
  100 * pcoa_coastal$eig[1:2] / sum(abs(pcoa_coastal$eig)), 1
)

ggplot(pcoa_df, aes(x = Axis1, y = Axis2)) +
  geom_line(aes(group = plot_number), colour = "grey60", linewidth = 0.4) +
  geom_point(aes(colour = sample, shape = sample), size = 3) +
  geom_label_repel(
    aes(label = paste(plot_number, sample)),
    size = 3, max.overlaps = 20, seed = 1
  ) +
  scale_colour_manual(values = c(first = "#2980b9", last = "#e67e22"),
                      labels = c(first = "First survey", last = "Last survey")) +
  scale_shape_manual(values  = c(first = 16, last = 17),
                     labels  = c(first = "First survey", last = "Last survey")) +
  coord_equal() +
  labs(
    title    = "PCoA — excluded coastal/outlier plots",
    subtitle = sprintf("Bray–Curtis · Axis 1 = %.1f%% · Axis 2 = %.1f%%",
                       var_explained[1], var_explained[2]),
    x = sprintf("PCoA Axis 1 (%.1f%%)", var_explained[1]),
    y = sprintf("PCoA Axis 2 (%.1f%%)", var_explained[2]),
    colour = NULL, shape = NULL
  ) +
  theme_classic(base_size = 11) +
  theme(legend.position = "bottom")
```

```{r coastal-species}
# Top species in each coastal plot (last survey only, for brevity)
coastal_last <- meta_coastal %>%
  filter(sample == "last") %>%
  pull(plot_number)

coastal_species <- community_long_paired %>%
  filter(plot_number %in% coastal_last, sample == "last") %>%
  arrange(plot_number, desc(cover)) %>%
  group_by(plot_number) %>%
  slice_head(n = 10) %>%
  ungroup()

print_full(coastal_species,
           caption = "Top 10 species by cover in each coastal plot (last survey).")
```

---

# Part 8 — Suggested next steps

Based on the results above, the following analyses are the logical
continuation — choose based on what the figures showed.

```{r next-steps, echo=FALSE, results='asis'}
cat(paste(
  "**A — dbRDA (distance-based redundancy analysis)**",
  "If haplontuthekja or vegetation height changed significantly in Part 5, use",
  "dbRDA to formally partition variance between the temporal effect (*sample*)",
  "and the environmental gradient, while conditioning on plot identity:",
  "",
  "```r",
  "vegan::dbrda(",
  "  mat_strict ~ sample + haplontuthekja + Condition(plot_number),",
  "  data   = meta_strict,",
  "  method = \"bray\"",
  ")",
  "```",
  "",
  "**B — Stratum-specific PERMANOVAs** *(completed — see Part 6)*",
  "Within-stratum paired PERMANOVAs by moisture class were carried out in Part 6,",
  "including a gap_years covariate model. The temporal signal is consistent at",
  "R\u00b2 \u2248 4.4% across Þurrlendi, Deiglendi, and Mýri; Flói bogs are stable (p \u2248 0.12).",
  "The next extension is to stratify by habitat type (vistgerð) rather than moisture",
  "class — finer groupings that may reveal where change is concentrated. See the",
  "habitat_type_name analysis planned below.",
  "",
  "**C — Taxon-level ecology**",
  "Review the indicator taxa from Part 4 against Icelandic vegetation",
  "literature. Key questions:",
  "- Are the 'last survey' gainers typically associated with warming, reduced",
  "  grazing, or nitrogen deposition in Icelandic heathlands?",
  "- Do *Racomitrium* or *Dryas* appear as consistent losers?",
  "- Are wetland sedges (*Carex lyngbyei*, *C. nigra*) showing directional change?",
  "",
  "**D — Spatial autocorrelation of turnover**",
  "Test whether compositional change (the per-plot Bray–Curtis from Part 3) is",
  "spatially structured using a Mantel test:",
  "",
  "```r",
  "geo_dist <- dist(cbind(env_first$longitude_1, env_first$latitude_1))",
  "vegan::mantel(bc_dist_matrix, geo_dist, method = \"pearson\", permutations = 999)",
  "```",
  sep = "\n"
))
```

---

---

---

# Part 8 — Habitat-stratified temporal analysis

This section extends the core analysis by (a) removing residual DCA1 outliers,
(b) splitting the dataset into ecologically coherent habitat groups, (c) running
a per-habitat distance-based RDA to partition what drives temporal change within
each group, (d) testing whether change rate is confounded by survey interval
length, and (e) identifying the species loading most strongly on the constrained
axes.

---

## 8.1 DCA outlier identification and exclusion

Plots with DCA1 > 2.0 lie more than two complete species-replacement distances
from the DCA origin — far enough to sit in a qualitatively different
compositional space from the bulk of the inland community. Their inclusion would
inflate within-habitat variance in every downstream subset. Inspection of their
habitat type, project, and scale clarifies whether they represent genuine habitat
boundaries or data artefacts.

```{r dca-outlier-id}
# site_env is built in Section 2.5 (DCA chunk); rows = one plot-survey combination
# DCA axes are in SD units of species turnover (≈ one full species replacement = 1 SD)

dca_outlier_thresh <- 2.0   # SD units on DCA1 — adjustable

dca_outliers_info <- site_env %>%
  filter(DCA1 > dca_outlier_thresh) %>%
  arrange(desc(DCA1)) %>%
  select(
    plot_number, sample, DCA1, DCA2,
    any_of(c(
      "habitat_type_name", "moisture_name", "soil_type_name",
      "topography_name", "project_name", "group_scale_name"
    ))
  ) %>%
  distinct(plot_number, .keep_all = TRUE)

# Vector of plot_numbers to exclude in all downstream habitat analyses
dca_outlier_plots <- unique(dca_outliers_info$plot_number)

cat("DCA1 outlier plots (DCA1 >", dca_outlier_thresh, "SD):\n")
print(dca_outliers_info)
cat("\nPlot IDs flagged for exclusion:", paste(dca_outlier_plots, collapse = ", "), "\n")
cat("Total plots excluded at this step:", length(dca_outlier_plots), "\n")
```

---

## 8.2 Habitat-type subsets

Starting from `meta_dca` / `mat_dca` (which already exclude the NMDS outliers
and coastal habitat types from Sections 1.2 and 2.4b), the DCA1 outliers above
are removed. Habitat types are then grouped into three ecologically coherent
classes following the *Vistgerðir á Íslandi* L-code hierarchy:

- **mire_fen** (L8 — Votlendi): types whose names contain *flóavist* or *mýravist*,
  plus the smaller wetland types (Rekjuvist, Dýjavist, Hrossanálarvist)
- **heath** (L10 — Móar): types whose names contain *móavist*, or Víðikjarrvist /
  Víðimóavist (tall-willow scrub, L10.9–10)
- **lava** (L6 — Hraunvist): all types containing *hraunavist*
- **other**: remaining types (highland grasslands, fell-fields, woodlands, etc.)

Only groups with **n ≥ 10 complete pairs** are carried forward.

```{r habitat-subsets}
# --- 1. Remove DCA1 outliers from the already-clean DCA dataset ---------------
keep_dca_clean <- !meta_dca$plot_number %in% dca_outlier_plots
meta_hab <- meta_dca[keep_dca_clean, ]
mat_hab  <- mat_dca[keep_dca_clean, ]
mat_hab  <- mat_hab[, colSums(mat_hab) > 0]   # drop taxa lost after exclusions

# Ensure gap_years is clean (prevent .x/.y suffix from double-join in prep)
meta_hab <- meta_hab %>%
  select(-any_of(c("gap_years", "gap_years.x", "gap_years.y"))) %>%
  left_join(gap_tbl, by = "plot_number")

cat("Rows after DCA1 outlier removal:", nrow(meta_hab),
    "(", length(unique(meta_hab$plot_number)), "plots )\n")

# --- 2. Assign habitat groups based on Icelandic name conventions -------------
# Regex patterns match the published vistgerð names (UTF-8; case-insensitive)
meta_hab <- meta_hab %>%
  mutate(
    habitat_group = case_when(
      str_detect(habitat_type_name,
        regex("flóavist|mýravist|Rekjuvist|Dýjavist|Hrossanálarvist",
              ignore_case = TRUE))   ~ "mire_fen",
      str_detect(habitat_type_name,
        regex("móavist|Víðikjarrvist|Víðimóavist",
              ignore_case = TRUE))   ~ "heath",
      str_detect(habitat_type_name,
        regex("hraunavist",
              ignore_case = TRUE))   ~ "lava",
      TRUE                           ~ "other"
    )
  )

# --- 3. Count complete pairs per group ----------------------------------------
pair_counts_hab <- meta_hab %>%
  filter(sample == "first") %>%
  count(habitat_group, name = "n_pairs") %>%
  arrange(desc(n_pairs))

cat("\nComplete pairs per habitat group:\n")
print(pair_counts_hab)

# Habitat types within each group — useful for ecological QC
hab_type_detail <- meta_hab %>%
  filter(sample == "first") %>%
  count(habitat_group, habitat_type_name, sort = TRUE)

cat("\nDetailed habitat_type_name breakdown:\n")
print(hab_type_detail)

# --- 4. Groups meeting the n >= 10 threshold ----------------------------------
groups_sufficient <- pair_counts_hab %>%
  filter(n_pairs >= 10) %>%
  pull(habitat_group)

cat("\nGroups with n >= 10 pairs (carried forward):", paste(groups_sufficient, collapse = ", "), "\n")

# ---- 5. Display labels and ASCII-safe file slugs ----------------------------
# Maps internal habitat_group codes → proper bilingual names from
# Vistgerðir á Íslandi (Fjölrit Náttúrufræðistofnunar nr. 54, 2016).
# hab_slug uses only ASCII characters so filenames stay portable across
# operating systems — follows the existing p_nmds_*, p_dca_* convention.
hab_display <- c(
  mire_fen = "Mýrar og flóar (L8)",
  heath    = "Móar (L10)",
  lava     = "Hraunvist (L6)",
  other    = "Annað"
)
hab_slug <- c(
  mire_fen = "myrar_floar",
  heath    = "moar",
  lava     = "hraunvist",
  other    = "annad"
)
```

---

## 8.3 Per-habitat dbRDA

For each sufficient habitat group, `capscale` (distance-based RDA on Bray–Curtis)
is run with `Condition(plot_number)` to partial out the baseline composition of
each plot and isolate the *within-plot temporal signal*. Environmental predictors
are tested at the margin using `anova.cca(by = "margin")`, which gives each term
adjusted for all others — the most conservative and interpretable approach.

Predictors tested: `gap_years`, `haeddypimetrar` (elevation), `soil_depth`,
`moisture_name`, `topography_name`, `permafrost`. Any predictor missing in
≥ 40% of rows or constant within the subset is dropped before modelling.

```{r per-habitat-dbrda, cache=TRUE}

# ---- helper: run dbRDA for one habitat subset --------------------------------
run_habitat_dbrda <- function(meta_sub, mat_sub, group_label) {
  # meta_sub : data frame with plot_number, sample, and predictor columns
  # mat_sub  : community matrix; rownames must equal paste(plot_number, sample, sep="_")
  # Returns a named list, or NULL on failure (with a message).

  n_plots <- length(unique(meta_sub$plot_number))
  if (nrow(meta_sub) < 6 || n_plots < 5) {
    message(group_label, ": fewer than 5 plots with 2 surveys — skipping.")
    return(NULL)
  }

  candidate_preds <- c(
    "gap_years", "haeddypimetrar", "soil_depth",
    "moisture_name", "topography_name", "permafrost"
  )

  # Step 1 — keep only predictors present and not too sparse
  usable_preds <- candidate_preds[vapply(candidate_preds, function(v) {
    v %in% names(meta_sub) && mean(is.na(meta_sub[[v]])) < 0.40
  }, logical(1))]

  if (length(usable_preds) == 0) {
    message(group_label, ": no usable predictors (all absent or > 40% NA).")
    return(NULL)
  }

  # Step 2 — convert character/logical to factor, drop_na on predictors
  meta_cc <- meta_sub %>%
    mutate(across(where(is.character), as.factor),
           across(where(is.logical),   as.factor)) %>%
    select(plot_number, sample, all_of(usable_preds)) %>%
    drop_na()

  # Step 3 — remove predictors that are constant in this subset (zero variance)
  usable_preds <- usable_preds[vapply(usable_preds, function(v) {
    col <- meta_cc[[v]]
    if (is.numeric(col)) var(col, na.rm = TRUE) > 1e-8
    else                 length(unique(col)) > 1
  }, logical(1))]

  if (length(usable_preds) == 0) {
    message(group_label, ": all predictors are constant after subsetting.")
    return(NULL)
  }

  meta_cc <- meta_cc %>% select(plot_number, sample, all_of(usable_preds))

  # Step 4 — align matrix rows to complete-case meta
  keep_ids <- paste(meta_cc$plot_number, meta_cc$sample, sep = "_")
  mat_cc   <- mat_sub[rownames(mat_sub) %in% keep_ids, , drop = FALSE]
  mat_cc   <- mat_cc[, colSums(mat_cc) > 0, drop = FALSE]

  # Guarantee row order matches meta_cc
  mat_cc <- mat_cc[paste(meta_cc$plot_number, meta_cc$sample, sep = "_"), ]

  if (nrow(meta_cc) < 6 || length(unique(meta_cc$plot_number)) < 4) {
    message(group_label, ": too few complete-case rows after predictor NA removal.")
    return(NULL)
  }

  # Step 5 — build and run capscale (dbRDA; Bray-Curtis + Condition on plot)
  pred_formula <- paste(c(usable_preds, "Condition(plot_number)"), collapse = " + ")
  f <- as.formula(paste0("mat_cc ~ ", pred_formula))

  set.seed(123)
  ord <- tryCatch(
    vegan::capscale(f, data = meta_cc, distance = "bray"),
    error = function(e) { message(group_label, ": capscale error — ", e$message); NULL }
  )
  if (is.null(ord)) return(NULL)

  # Step 6 — permutation tests
  set.seed(123)
  perm_overall <- vegan::anova.cca(ord, permutations = 999)

  set.seed(123)
  perm_margin <- tryCatch(
    vegan::anova.cca(ord, by = "margin", permutations = 999),
    error = function(e) { message(group_label, ": margin ANOVA error — ", e$message); NULL }
  )

  list(
    group        = group_label,
    n_rows       = nrow(meta_cc),
    n_plots      = length(unique(meta_cc$plot_number)),
    predictors   = usable_preds,
    ordination   = ord,
    perm_overall = perm_overall,
    perm_margin  = perm_margin
  )
}

# ---- Run for every sufficient habitat group ----------------------------------
dbrda_results <- list()

for (grp in groups_sufficient) {

  # Subset metadata — join gap_tbl fresh to guarantee gap_years is present
  meta_g <- meta_hab %>%
    filter(habitat_group == grp) %>%
    select(-any_of(c("gap_years.x", "gap_years.y")))   # safety net

  # Subset community matrix by row names
  ids_g  <- paste(meta_g$plot_number, meta_g$sample, sep = "_")
  mat_g  <- mat_hab[rownames(mat_hab) %in% ids_g, , drop = FALSE]
  mat_g  <- mat_g[, colSums(mat_g) > 0, drop = FALSE]

  dbrda_results[[grp]] <- run_habitat_dbrda(meta_g, mat_g, hab_display[grp])
}

# ---- Print results -----------------------------------------------------------
for (grp in names(dbrda_results)) {
  res <- dbrda_results[[grp]]
  if (is.null(res)) next

  cat("\n", strrep("=", 60), "\n", sep = "")
  cat("Habitat group : ", res$group,
      "  |  n = ", res$n_rows, " rows / ", res$n_plots, " plots\n", sep = "")
  cat("Predictors    : ", paste(res$predictors, collapse = ", "), "\n\n", sep = "")

  # Variance partitioning summary
  cat("Constrained inertia (total):", round(res$ordination$CCA$tot.chi, 4), "\n")
  cat("Conditioned inertia (plot) :", round(res$ordination$pCCA$tot.chi, 4), "\n")
  cat("Unconstrained residual     :", round(res$ordination$CA$tot.chi,  4), "\n\n")

  cat("-- Overall model permutation test --\n")
  print(res$perm_overall)

  if (!is.null(res$perm_margin)) {
    cat("\n-- Marginal term significance (each term adjusted for all others) --\n")
    print(res$perm_margin)
  }
}
```

---

## 8.4 Survey gap normalisation — does change rate differ between habitat groups?

A longer interval between surveys mechanically allows more community change to
accumulate. Before comparing turnover magnitudes across habitat groups it is
important to confirm that the groups do not simply differ in how long ago their
plots were first visited. The per-plot Bray–Curtis values from `bc_with_gap`
(Section 3.1) are modelled against `gap_years` and `habitat_group` together.
An interaction term tests whether the **rate** of change (BC per year) differs
between groups after accounting for interval length.

```{r gap-years-bc-by-habitat, fig.height=5, fig.width=8, cache.extra=hab_display}

# Join habitat_group (one row per plot — use "first" survey to avoid duplication)
bc_hab <- bc_with_gap %>%
  left_join(
    meta_hab %>%
      filter(sample == "first") %>%
      select(plot_number, habitat_group) %>%
      distinct(),
    by = "plot_number"
  ) %>%
  filter(!is.na(bc), !is.na(gap_years), !is.na(habitat_group))

cat("Plots with complete BC + gap_years + habitat_group data:", nrow(bc_hab), "\n")
cat("by group:\n")
print(count(bc_hab, habitat_group))

# --- Interaction model: does the BC~gap_years slope vary by habitat group? ---
bc_lm_int <- lm(bc ~ gap_years * habitat_group, data = bc_hab)
bc_lm_add <- lm(bc ~ gap_years + habitat_group, data = bc_hab)

cat("\n--- Interaction model: BC ~ gap_years * habitat_group ---\n")
print(summary(bc_lm_int))

cat("\n--- Additive model: BC ~ gap_years + habitat_group ---\n")
print(summary(bc_lm_add))

cat("\n--- AIC comparison (lower = better fit) ---\n")
print(AIC(bc_lm_int, bc_lm_add))

# --- Figure: BC vs gap_years, one regression line per habitat group ----------
p_bc_gap_hab <- ggplot(bc_hab,
                       aes(x = gap_years, y = bc,
                           colour = habitat_group, fill = habitat_group)) +
  geom_point(alpha = 0.55, size = 2.2) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.12, linewidth = 0.9) +
  scale_colour_brewer(palette = "Set1", name = "Habitat group",
                      labels  = hab_display) +
  scale_fill_brewer(  palette = "Set1", name = "Habitat group",
                      labels  = hab_display) +
  labs(
    x        = "Survey interval (years)",
    y        = "Bray–Curtis dissimilarity (first vs last)",
    title    = "Community turnover vs survey interval by habitat group",
    subtitle = "Separate OLS lines; shading = 95% CI"
  ) +
  theme_classic() +
  theme(legend.position = "right")

print(p_bc_gap_hab)
ggsave("outputs/p_bc_gap_hab.png", plot = p_bc_gap_hab,
       width = 8, height = 5, dpi = 300)
```

---

## 8.5 Species contributions to constrained RDA axes

For each habitat group whose overall dbRDA was significant, the species scores
(weighted averages on the constrained axes) identify which taxa are most
strongly associated with the direction of temporal change. Positive loaders on
CAP1 gained relative to the within-plot centroid; negative loaders lost.

```{r species-contributions, fig.height=6, fig.width=8}

sig_alpha <- 0.05   # overall model significance threshold

for (grp in names(dbrda_results)) {
  res <- dbrda_results[[grp]]
  if (is.null(res)) next

  # Gate on overall model significance
  overall_p <- res$perm_overall[1, "Pr(>F)"]
  if (is.na(overall_p) || overall_p > sig_alpha) {
    cat(grp, ": overall model not significant (p =",
        round(overall_p, 3), ") — skipping species scores.\n\n")
    next
  }

  cat("\n", strrep("=", 60), "\n", sep = "")
  cat("Species contributions —", hab_display[grp],
      "  (overall p =", round(overall_p, 3), ")\n\n")

  # Number of constrained axes available
  n_constrained <- length(res$ordination$CCA$eig)
  n_axes        <- min(2L, n_constrained)

  # Species scores: weighted averages of site scores (CAP1, CAP2)
  sp_raw <- tryCatch(
    vegan::scores(res$ordination, display = "species",
                  choices = seq_len(n_axes)),
    error = function(e) {
      message("  Could not extract species scores: ", e$message); NULL
    }
  )
  if (is.null(sp_raw)) next

  # Tidy the score matrix
  sp_scores <- as_tibble(sp_raw, rownames = "taxon_name")

  # Rename CAP → RDA for readability; handle 1- or 2-axis cases
  names(sp_scores) <- sub("^CAP1$", "RDA1", names(sp_scores))
  names(sp_scores) <- sub("^CAP2$", "RDA2", names(sp_scores))
  if (!"RDA2" %in% names(sp_scores)) sp_scores$RDA2 <- NA_real_

  sp_scores <- sp_scores %>%
    mutate(
      RDA1_abs       = abs(RDA1),
      direction_RDA1 = if_else(RDA1 >= 0, "gainer (+RDA1)", "loser (−RDA1)")
    ) %>%
    arrange(desc(RDA1_abs))

  cat("Top 20 taxa by |loading| on RDA axis 1:\n")
  print(
    sp_scores %>%
      slice_head(n = 20) %>%
      select(taxon_name, RDA1, RDA2, direction_RDA1),
    n = 20
  )

  # --- Horizontal bar chart: top 20 species loadings on RDA1 -----------------
  top_sp <- 20L
  sp_plot_dat <- sp_scores %>%
    slice_head(n = top_sp) %>%
    mutate(taxon_name = fct_reorder(taxon_name, RDA1))

  p_sp <- ggplot(sp_plot_dat,
                 aes(x = RDA1, y = taxon_name, fill = direction_RDA1)) +
    geom_col(width = 0.7) +
    geom_vline(xintercept = 0, linewidth = 0.4, colour = "grey40") +
    scale_fill_manual(
      values = c("gainer (+RDA1)" = "#E41A1C",
                 "loser (−RDA1)"  = "#377EB8"),
      name = NULL
    ) +
    labs(
      x        = "Species loading on RDA axis 1 (constrained dbRDA)",
      y        = NULL,
      title    = paste0("Temporal species drivers — ", hab_display[grp]),
      subtitle = paste0("Top ", top_sp, " taxa by |loading|; Bray–Curtis capscale, ",
                        "Condition(plot_number)")
    ) +
    theme_classic() +
    theme(
      legend.position = "bottom",
      axis.text.y     = element_text(size = 8, face = "italic"),
      plot.subtitle   = element_text(size = 9, colour = "grey40")
    )

  print(p_sp)
  ggsave(
    file.path("outputs", paste0("p_sp_rda_", hab_slug[grp], ".png")),
    plot  = p_sp,
    width = 8, height = 6, dpi = 300
  )
}
```

---

## 8.6 Per-habitat species gainers and losers

For each habitat group, this section computes the mean cover change
(last − first) for every species that was present in at least three plots
within the group. A paired Wilcoxon signed-rank test across plots gives a
p-value and rank-biserial correlation (r) as effect size. The top 20 gainers
and losers by mean Δcover are shown in a diverging bar chart, with significant
species (raw p ≤ 0.05) highlighted in solid colour and non-significant ones
shown faded.

This is the plain-language complement to Section 8.5: instead of asking "which
species co-vary with the constrained RDA axis", it asks "which species actually
changed in cover, and by how much, within this habitat type?"

```{r species-gainers-losers, fig.height=7, fig.width=9}

# Minimum number of plots in which a species must be non-zero (first OR last)
# before it is included in the per-habitat Wilcoxon test
min_plots_present <- 3L

# Significance threshold for solid vs faded colouring
sig_thresh_sp <- 0.05

# ---- Widen community_long_paired to one row per plot × taxon × survey -------
# community_long_paired has columns: plot_number, sample, taxon_name, cover
# We need first and last cover side by side for each plot × taxon combination.

cover_wide_sp <- community_long_paired %>%
  # Keep only plots in our cleaned habitat dataset
  filter(plot_number %in% unique(meta_hab$plot_number)) %>%
  pivot_wider(
    names_from  = sample,
    values_from = cover,
    values_fill = 0          # species absent in a survey = cover 0
  ) %>%
  # Ensure both columns exist even if one survey had no records for a species
  mutate(
    first = if ("first" %in% names(.)) first else 0,
    last  = if ("last"  %in% names(.)) last  else 0
  )

# Join habitat_group from meta_hab (one row per plot, take from "first" sample)
hab_lookup <- meta_hab %>%
  filter(sample == "first") %>%
  select(plot_number, habitat_group) %>%
  distinct()

cover_wide_sp <- cover_wide_sp %>%
  left_join(hab_lookup, by = "plot_number") %>%
  filter(!is.na(habitat_group))

# ---- Per-habitat × taxon summary + Wilcoxon ---------------------------------
sp_change_hab <- cover_wide_sp %>%
  group_by(habitat_group, taxon_name) %>%
  summarise(
    n_plots      = n(),
    n_nonzero    = sum(first > 0 | last > 0),
    mean_first   = mean(first),
    mean_last    = mean(last),
    mean_delta   = mean(last - first),
    # Paired Wilcoxon: requires at least min_plots_present plots and
    # at least one non-tied pair
    wilcox_p = {
      d <- last - first
      if (sum(d != 0) < min_plots_present) NA_real_
      else tryCatch(
        wilcox.test(last, first, paired = TRUE, exact = FALSE)$p.value,
        error   = function(e) NA_real_,
        warning = function(w) suppressWarnings(
          wilcox.test(last, first, paired = TRUE, exact = FALSE)$p.value
        )
      )
    },
    # Rank-biserial correlation as effect size (= 2W/n(n+1) - 1 via coin, or
    # approximated here as the standardised W statistic)
    wilcox_r = {
      d <- last - first
      n <- sum(d != 0)
      if (n < min_plots_present) NA_real_
      else tryCatch({
        wt <- wilcox.test(last, first, paired = TRUE, exact = FALSE)
        # rank-biserial r = 1 - 2W / (n*(n+1)/2)  [Kerby 2014]
        1 - (2 * wt$statistic) / (n * (n + 1) / 2)
      },
      error   = function(e) NA_real_,
      warning = function(w) suppressWarnings({
        wt <- wilcox.test(last, first, paired = TRUE, exact = FALSE)
        1 - (2 * wt$statistic) / (n * (n + 1) / 2)
      }))
    },
    .groups = "drop"
  ) %>%
  filter(n_nonzero >= min_plots_present) %>%
  mutate(
    sig        = !is.na(wilcox_p) & wilcox_p <= sig_thresh_sp,
    direction  = if_else(mean_delta >= 0, "gainer", "loser"),
    alpha_val  = if_else(sig, 1.0, 0.3)
  )

# ---- Print summary table per group ------------------------------------------
for (grp in groups_sufficient) {

  tbl_grp <- sp_change_hab %>%
    filter(habitat_group == grp, sig) %>%
    arrange(desc(abs(mean_delta))) %>%
    select(taxon_name, mean_first, mean_last, mean_delta, wilcox_p, wilcox_r)

  cat("\n", strrep("=", 60), "\n", sep = "")
  cat("Significant changers (raw p ≤ 0.05) —", hab_display[grp], "\n")
  cat("  Gainers:", sum(tbl_grp$mean_delta > 0),
      "  Losers:", sum(tbl_grp$mean_delta < 0), "\n\n")
  print(tbl_grp, n = 40)
}

# ---- Diverging bar chart per habitat group -----------------------------------
for (grp in groups_sufficient) {

  grp_dat <- sp_change_hab %>%
    filter(habitat_group == grp) %>%
    # Keep top 20 gainers + top 20 losers by |mean_delta|
    arrange(desc(mean_delta)) %>%
    { bind_rows(slice_head(., n = 20), slice_tail(., n = 20)) } %>%
    distinct(taxon_name, .keep_all = TRUE) %>%
    mutate(
      taxon_label = taxon_name,
      taxon_label = fct_reorder(taxon_label, mean_delta),
      fill_col    = case_when(
        sig & direction == "gainer" ~ "sig_gain",
        sig & direction == "loser"  ~ "sig_loss",
        direction == "gainer"       ~ "ns_gain",
        TRUE                        ~ "ns_loss"
      )
    )

  n_sig  <- sum(grp_dat$sig)
  n_grp  <- sp_change_hab %>%
    filter(habitat_group == grp) %>%
    pull(n_plots) %>% first()

  p_gl <- ggplot(grp_dat,
                 aes(x = mean_delta, y = taxon_label, fill = fill_col)) +
    geom_col(width = 0.75) +
    geom_vline(xintercept = 0, linewidth = 0.4, colour = "grey30") +
    scale_fill_manual(
      values = c(
        sig_gain = "#E41A1C",
        sig_loss = "#377EB8",
        ns_gain  = "#FABEBE",
        ns_loss  = "#A6C8E0"
      ),
      labels = c(
        sig_gain = paste0("Gainer (p ≤ ", sig_thresh_sp, ")"),
        sig_loss = paste0("Loser  (p ≤ ", sig_thresh_sp, ")"),
        ns_gain  = "Gainer (ns)",
        ns_loss  = "Loser  (ns)"
      ),
      name = NULL
    ) +
    labs(
      x        = "Mean Δcover (last − first, % Braun–Blanquet)",
      y        = NULL,
      title    = paste0("Species gainers and losers — ", hab_display[grp]),
      subtitle = paste0(
        "Top 20 gainers + top 20 losers by mean Δcover; ",
        n_sig, " significant (paired Wilcoxon, raw p ≤ ", sig_thresh_sp, ")"
      )
    ) +
    theme_classic() +
    theme(
      axis.text.y   = element_text(size = 7.5, face = "italic"),
      legend.position = "bottom",
      plot.subtitle = element_text(size = 9, colour = "grey40")
    )

  print(p_gl)
  ggsave(
    file.path("outputs", paste0("p_gainers_losers_", hab_slug[grp], ".png")),
    plot = p_gl, width = 9, height = 7, dpi = 300
  )
}
```

---

# Session info

```{r session-info}
sessionInfo()
```
