---
title: "v_plants variable screening"
author: "Valtýr"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(lubridate)
```

```{r source-prep}
# Shared computational backbone used across reports
source("v_plants_prep.R")
```

**Note on drift control.** The objects created by `v_plants_prep.R` are the single source of truth.  
Chunks below are kept for transparency and diagnostics; where relevant they avoid overwriting backbone objects.

## Purpose

This note documents **which columns in `plants_export/v_plants_allt.csv` are excluded from modelling and ordination**, based on the `var_profile` screen in `analysis_precheck_checklist.R` (missingness, type, and role).

It is meant as an audit trail so future analyses of the vegetation dataset can clearly see which variables were ignored and why.

## Columns excluded without further discussion

These variables are either **effectively unmeasured (~all missing)** or **free-text notes / images** that are not suitable as structured covariates.

```{r excluded_columns_definition}
# Variables with ~complete missingness or effectively not measured
vars_exclude_missing <- c(
  "biomass",
  "precipitation",
  "stone_class",
  "temperature_january",
  "temperature_july",
  "temperature_year",
  "heildarþekja",
  "subplot_carbon",
  "subplot_nitrogen",
  "subplot_ph",
  "plant_community_other",
  "plant_community_3_code",
  "plant_community_3_id",
  "plant_community_3_name",
  "slope_aspect",
  "slope",
  "autt",
  "surface_type_code",
  "surface_type_id",
  "surface_type_name",
  "soil_temperature_2",
  "plant_community_cover"
)

# Free-text/comment/image fields not used as covariates
vars_exclude_text <- c(
  "comments",
  "subplot_comments",
  "images",
  "subplot_images",
  "project_descrtiption"
)

vars_exclude_strict <- c(vars_exclude_missing, vars_exclude_text)
vars_exclude_strict
```

You can use this vector when preparing analysis-ready tables, for example:

```{r usage_example, eval=FALSE}
plants_path <- file.path(getwd(), "plants_export", "v_plants_allt.csv")
v_plants <- readr::read_csv(plants_path, show_col_types = FALSE)

v_plants_use <- v_plants %>%
  select(-any_of(vars_exclude_strict))
```

## Reading and filtering `v_plants_allt`

This section mirrors the core loading and cleaning used in `analysis_precheck_checklist.R`, so downstream scripts can source a **single, consistent definition** of the analysis-ready vegetation data.

The `heiti_include` mechanism is a **hard include-list for plots**:

- set `use_heiti_include <- TRUE` to activate it;
- put the exact plot codes you want to keep into `heiti_include`, e.g.

```r
heiti_include <- c("KMM68", "KMM70", "KMM72")
```

Only those `plot_number` values are then retained in `v_plants_use`. When `use_heiti_include` is `FALSE`, all plots in the file are kept.

```{r read_and_filter_v_plants}
# Diagnostic-only: core data objects are sourced from v_plants_prep.R above.
tibble(
  object = c("v_plants_raw", "v_plants2", "v_plants_use"),
  n_rows = c(nrow(v_plants_raw), nrow(v_plants2), nrow(v_plants_use)),
  n_cols = c(ncol(v_plants_raw), ncol(v_plants2), ncol(v_plants_use))
)
```

## Categorize variables for analysis

This creates a reproducible variable catalog from the filtered dataset (`v_plants_use`) and assigns each variable to an analysis role.

```{r categorize_variables}
id_vars <- c(
  "id", "plot_id", "subplot_id", "group_type_subplot_id",
  "project_id", "project_number", "analyser_id", "tag_id"
)

time_pairing_vars <- c("field_date", "field_date_parsed", "year", "plot_number")

species_cover_core_vars <- c(
  "taxon_id", "taxon_name", "scale_value_code", "scale_value_value", "tag_name"
)

spatial_vars <- c("longitude_1", "latitude_1", "longitude_2", "latitude_2")

design_method_vars <- c(
  "plot_size_id", "plot_size_width", "plot_size_length",
  "subplot_size_id", "subplot_size_width", "subplot_size_length",
  "subplot_group_scale_id", "group_scale_name",
  "subplot_taxon_scale_id", "taxon_scale_name",
  "project_name", "analyser_name", "area"
)

var_catalog <- tibble(variable = names(v_plants_use)) %>%
  mutate(
    type = map_chr(variable, ~ class(v_plants_use[[.x]])[1]),
    prop_missing = map_dbl(variable, ~ mean(is.na(v_plants_use[[.x]]))),
    n_unique = map_dbl(variable, ~ n_distinct(v_plants_use[[.x]], na.rm = TRUE)),
    analysis_role = case_when(
      variable %in% id_vars ~ "identifier",
      variable %in% time_pairing_vars ~ "time_or_pairing",
      variable %in% species_cover_core_vars ~ "species_cover_core",
      variable %in% spatial_vars ~ "spatial",
      variable %in% design_method_vars ~ "design_or_metadata",
      type %in% c("numeric", "integer", "double") & prop_missing <= 0.40 ~ "env_numeric_candidate",
      type %in% c("character", "factor") & n_unique <= 20 & prop_missing <= 0.40 ~ "env_factor_candidate",
      type %in% c("logical") & prop_missing <= 0.40 ~ "env_factor_candidate",
      type %in% c("numeric", "integer", "double") & prop_missing > 0.40 ~ "deferred_sparse_numeric",
      type %in% c("character", "factor", "logical") & prop_missing > 0.40 ~ "deferred_sparse_categorical",
      TRUE ~ "review_manually"
    )
  ) %>%
  arrange(analysis_role, desc(prop_missing), variable)

var_catalog
```

```{r variable_groups_for_downstream_use}
vars_species_cover_core <- var_catalog %>%
  filter(analysis_role == "species_cover_core") %>%
  pull(variable)

vars_env_numeric <- var_catalog %>%
  filter(analysis_role == "env_numeric_candidate") %>%
  pull(variable)

vars_env_factor <- var_catalog %>%
  filter(analysis_role == "env_factor_candidate") %>%
  pull(variable)

vars_design_metadata <- var_catalog %>%
  filter(analysis_role == "design_or_metadata") %>%
  pull(variable)

vars_spatial <- var_catalog %>%
  filter(analysis_role == "spatial") %>%
  pull(variable)

vars_deferred_sparse <- var_catalog %>%
  filter(analysis_role %in% c("deferred_sparse_numeric", "deferred_sparse_categorical")) %>%
  pull(variable)

list(
  vars_species_cover_core = vars_species_cover_core,
  vars_env_numeric = vars_env_numeric,
  vars_env_factor = vars_env_factor,
  vars_design_metadata = vars_design_metadata,
  vars_spatial = vars_spatial,
  vars_deferred_sparse = vars_deferred_sparse
)
```

```{r summary_counts_by_role}
var_catalog %>%
  count(analysis_role, sort = TRUE)
```

## Paired-design essentials mirrored from precheck

The following pieces from `analysis_precheck_checklist.R` are required here to keep the analysis design consistent:

- define revisited plots (`n_years >= 2`) and `year_first` / `year_last`;
- assign `sample` as `first` / `last` from those years;
- harmonize cover values (`scale_value_value` with Braun-Blanquet fallback);
- aggregate duplicate taxa entries at `plot_number x sample x taxon_name`;
- enforce complete pairs (exactly two rows per plot) before matrix analyses.

```{r mirrored_paired_design_steps}
# Diagnostic-only: report paired-design objects produced by v_plants_prep.R
diag_counts <- list(
  n_v_plants2 = nrow(v_plants2),
  n_plot_years = nrow(plot_years),
  n_revisited_plots = nrow(revisited),
  n_rows_paired_long = nrow(paired_long),
  n_rows_paired_cover = nrow(paired_cover),
  n_rows_community_long = nrow(community_long),
  n_rows_complete_pairs = nrow(community_long_paired),
  n_plots_complete_pairs = dplyr::n_distinct(community_long_paired$plot_number)
)
diag_counts
```

## First analysis steps

Start analysis from `community_long_paired` and proceed in this order.

```{r first_steps_analysis}
# Diagnostic-only: use community_wide/meta/mat from v_plants_prep.R

# Step 2: basic matrix readiness checks
has_rows <- nrow(mat) > 0
two_rows_per_plot <- has_rows && all((meta %>% count(plot_number) %>% pull(n)) == 2)

readiness <- tibble(
  check = c(
    "matrix_rows",
    "matrix_taxa_columns",
    "all_plots_have_two_rows"
  ),
  value = c(
    nrow(mat),
    ncol(mat),
    as.integer(two_rows_per_plot)
  )
)
readiness

# Step 3: first-pass paired composition summaries
if (has_rows) {
  bray <- vegan::vegdist(mat, method = "bray")
  bray_summary <- summary(as.numeric(bray))
  bray_summary
} else {
  message("Skipping Bray-Curtis summary because matrix_rows == 0.")
}

env_candidates_ps %>%
  summarise(across(all_of(c(vars_env_numeric, vars_env_factor)), ~ mean(is.na(.x)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "prop_missing") %>%
  arrange(desc(prop_missing))
```

Interpretation rule for next decision point:

- keep variables in strict driver models if they have acceptable overall missingness and limited first-vs-last missingness shift;
- move sparse or strongly shifted variables to exploratory-only models.

```{r driver_variable_eligibility}
# Thresholds for strict driver set (edit as needed)
max_missing_strict <- 0.35
max_shift_pp_strict <- 10

candidate_vars <- intersect(c(vars_env_numeric, vars_env_factor), names(env_candidates_ps))

driver_eligibility <- purrr::map_dfr(candidate_vars, function(v) {
  x <- env_candidates_ps %>%
    mutate(is_miss = is.na(.data[[v]])) %>%
    group_by(sample) %>%
    summarise(prop_miss = mean(is_miss), .groups = "drop") %>%
    tidyr::pivot_wider(names_from = sample, values_from = prop_miss, values_fill = 0)

  miss_first <- x$first %||% NA_real_
  miss_last <- x$last %||% NA_real_
  delta_pp <- 100 * (miss_last - miss_first)
  prop_missing <- mean(is.na(env_candidates_ps[[v]]))

  tibble(
    variable = v,
    prop_missing = prop_missing,
    miss_first = miss_first,
    miss_last = miss_last,
    delta_pp = delta_pp,
    strict_ok = prop_missing <= max_missing_strict && abs(delta_pp) <= max_shift_pp_strict,
    exploratory_ok = prop_missing <= 0.60
  )
}) %>%
  mutate(
    use_class = case_when(
      strict_ok ~ "strict",
      exploratory_ok ~ "exploratory_only",
      TRUE ~ "exclude"
    )
  ) %>%
  arrange(use_class, desc(prop_missing), desc(abs(delta_pp)), variable)

driver_eligibility

vars_driver_strict <- driver_eligibility %>%
  filter(use_class == "strict") %>%
  pull(variable)

vars_driver_exploratory <- driver_eligibility %>%
  filter(use_class == "exploratory_only") %>%
  pull(variable)

vars_driver_exclude <- driver_eligibility %>%
  filter(use_class == "exclude") %>%
  pull(variable)

list(
  vars_driver_strict = vars_driver_strict,
  vars_driver_exploratory = vars_driver_exploratory,
  vars_driver_exclude = vars_driver_exclude
)
```

```{r strict_driver_set_curated}
# Curated strict driver set:
# - remove index-like fields (e.g., subplot_number)
# - prefer name-coded factors over duplicate *_id fields for interpretability
vars_driver_strict_proposed <- c(
  "háplöntuþekja",
  "soil_depth",
  "total_cover",
  "vegetation_height_1",
  "soil_type_name",
  "moisture_name",
  "topography_name",
  "permafrost"
)

# Keep only variables that exist in the current filtered data
vars_driver_strict_proposed <- intersect(vars_driver_strict_proposed, names(env_candidates_ps))

# Diagnostic-only object (avoid overwriting prep backbone objects)
env_strict_diag <- env_candidates_ps %>%
  select(plot_number, sample, all_of(vars_driver_strict_proposed)) %>%
  distinct()

# Diagnostics: level counts for categorical vars, SD for numeric vars
factor_checks <- env_strict_diag %>%
  summarise(across(
    where(~ is.character(.x) || is.factor(.x) || is.logical(.x)),
    ~ n_distinct(.x, na.rm = TRUE)
  )) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_levels")

numeric_checks <- env_strict_diag %>%
  summarise(across(where(is.numeric), ~ sd(.x, na.rm = TRUE))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "sd")

list(
  vars_driver_strict_final_from_prep = vars_driver_strict_final,
  vars_driver_strict_proposed = vars_driver_strict_proposed,
  factor_checks = factor_checks,
  numeric_checks = numeric_checks
)
```

```{r strict_complete_case_paired_readiness}
# Diagnostic-only: readiness for strict objects from v_plants_prep.R
strict_readiness <- tibble(
  n_plots = n_distinct(meta_strict$plot_number),
  n_rows = nrow(meta_strict),
  n_taxa = ncol(mat_strict),
  paired_ok = all((meta_strict %>% count(plot_number) %>% pull(n)) == 2)
)

strict_readiness
```

```{r strict_permanova_and_dispersion}
# Strict paired PERMANOVA baseline
perm_strict <- vegan::adonis2(
  mat_strict ~ sample,
  data = meta_strict,
  method = "bray",
  permutations = 999,
  strata = meta_strict$plot_number
)

# Dispersion check
bd_strict <- vegan::betadisper(
  vegan::vegdist(mat_strict, method = "bray"),
  group = meta_strict$sample
)
bd_test_strict <- vegan::permutest(bd_strict, permutations = 999)

perm_strict
bd_test_strict
```

```{r strict_envfit_screening}
# Driver screening with envfit on strict set (diagnostic table only).
# Uses nmds_strict and fit_strict objects created in v_plants_prep.R.

# Numeric vectors
fit_tbl_numeric <- tibble(
  variable = names(fit_strict$vectors$r),
  r2 = as.numeric(fit_strict$vectors$r),
  p_value = as.numeric(fit_strict$vectors$pvals)
) %>%
  arrange(p_value)

# Factor centroids (if present)
if (!is.null(fit_strict$factors) && length(fit_strict$factors$pvals) > 0) {
  fit_tbl_factor <- tibble(
    variable = names(fit_strict$factors$pvals),
    r2 = as.numeric(fit_strict$factors$r),
    p_value = as.numeric(fit_strict$factors$pvals)
  ) %>%
    arrange(p_value)
} else {
  fit_tbl_factor <- tibble(variable = character(), r2 = numeric(), p_value = numeric())
}

fit_tbl <- bind_rows(
  fit_tbl_numeric %>% mutate(kind = "numeric"),
  fit_tbl_factor %>% mutate(kind = "factor")
) %>%
  arrange(p_value)

fit_tbl
```

```{r strict_driver_model_paired_permanova}
# Next step: paired multivariable model
# Goal: test which strict drivers remain associated with composition
# after controlling for first/last sample and other retained drivers.

# 1) Start from envfit-supported candidates (p <= 0.05)
driver_candidates <- fit_tbl %>%
  filter(p_value <= 0.05, variable %in% vars_driver_strict_final) %>%
  pull(variable)

# 2) Basic collinearity pruning for numeric variables
num_candidates <- driver_candidates[driver_candidates %in% names(meta_strict)[sapply(meta_strict, is.numeric)]]
fac_candidates <- setdiff(driver_candidates, num_candidates)

if (length(num_candidates) >= 2) {
  cor_mat <- meta_strict %>%
    select(all_of(num_candidates)) %>%
    cor(use = "pairwise.complete.obs")

  high_pairs <- which(abs(cor_mat) > 0.80 & abs(cor_mat) < 1, arr.ind = TRUE)
  drop_numeric <- character(0)

  if (nrow(high_pairs) > 0) {
    # Drop the second variable in each high-correlation pair (simple deterministic rule)
    drop_numeric <- unique(colnames(cor_mat)[high_pairs[, "col"]])
  }
} else {
  cor_mat <- matrix(numeric(), nrow = 0, ncol = 0)
  drop_numeric <- character(0)
}

driver_final_diag <- c(setdiff(num_candidates, drop_numeric), fac_candidates)

# 3) Build paired PERMANOVA formula with sample + drivers
driver_formula <- as.formula(
  paste("mat_strict ~ sample +", paste(driver_final_diag, collapse = " + "))
)

perm_drivers_terms <- vegan::adonis2(
  formula = driver_formula,
  data = meta_strict,
  method = "bray",
  permutations = 999,
  strata = meta_strict$plot_number,
  by = "terms"
)

perm_drivers_margin <- vegan::adonis2(
  formula = driver_formula,
  data = meta_strict,
  method = "bray",
  permutations = 999,
  strata = meta_strict$plot_number,
  by = "margin"
)

list(
  driver_candidates = driver_candidates,
  dropped_numeric_due_to_high_correlation = drop_numeric,
  driver_final_diag = driver_final_diag,
  cor_mat_numeric = cor_mat,
  perm_drivers_terms = perm_drivers_terms,
  perm_drivers_margin = perm_drivers_margin
)
```


