---
title: "v_plants — Paired Resurvey Analysis"
author: "Valtýr"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: kable
    code_folding: hide
    fig_width: 9
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo        = TRUE,
  warning     = FALSE,
  message     = FALSE,
  cache       = TRUE,         # long-running chunks (NMDS, IndVal) cached to disk
  cache.path  = "cache/v_plants_",
  fig.align   = "center",
  fig.path    = "outputs/v_plants_analysis-",
  dev         = "png",
  dpi         = 300
)

# Helper: print a tibble fully if ≤ 400 rows, truncated otherwise
print_full <- function(x, caption = NULL) {
  n <- nrow(x)
  if (n <= 400) {
    knitr::kable(x, digits = 4, caption = caption)
  } else {
    knitr::kable(
      head(x, 400),
      digits  = 4,
      caption = paste0(caption, sprintf("  [showing first 400 of %d rows]", n))
    )
  }
}
```

```{r libraries}
library(tidyverse)
library(vegan)
library(broom)
library(ggrepel)
library(indicspecies)
library(knitr)

set.seed(123)
```

```{r bootstrap-required-objects}
# Knit runs in a clean session. Use shared prep script as single source of truth.
prep_script <- "v_plants_prep.R"
if (!file.exists(prep_script)) {
  stop("Required script not found: ", prep_script)
}
source(prep_script, local = .GlobalEnv)
```

---

> **Purpose.**  This document picks up directly after `v_plants_variable_screening.html`.
> All core objects are created by `v_plants_prep.R` (sourced above); the screening
> document is an audit trail that describes how they were selected and filtered.
>
> | Object | Description |
> |---|---|
> | `community_long_paired` | Long-format cover table (plot × sample × taxon) |
> | `community_wide_strict` | Wide community matrix with `plot_number` + `sample` columns |
> | `meta_strict` | Metadata aligned row-for-row with `community_wide_strict` |
> | `mat_strict` | Numeric species matrix (`r nrow(mat_strict)` rows × `r ncol(mat_strict)` cols) |
> | `nmds_strict` | `metaMDS` result on `mat_strict` (all plots, pre-outlier removal) |
> | `fit_strict` | `envfit` result on `nmds_strict` |
> | `vars_driver_strict_final`, `driver_final` | Variable vectors from screening |

---

# Part 1 — Preliminary NMDS (all plots)

## 1.1 Ordination plot

This first pass runs on the full `r length(unique(meta_strict$plot_number))`-plot dataset before any outlier removal.
It is included here as a diagnostic to show why the three coastal/dune plots
need special treatment.

```{r nmds-all-plot, fig.height=7,fig.cap=sprintf("NMDS of all %d revisited plots. Note the three outlier sites near NMDS1 = 17 that stretch the axis and compress the main community cloud.",length(unique(meta_strict$plot_number)))}

# Extract site scores
nmds_scores <- as.data.frame(scores(nmds_strict, display = "sites")) %>%
  bind_cols(meta_strict %>% select(plot_number, sample))

# Build envfit data including gap_years in addition to the strict drivers
env_all <- meta_strict %>%
  select(all_of(c(
    vars_driver_strict_final,        # robust core drivers
    "gap_years",                     # time gap effect
    #"habitat_type_name",             # broader habitat grouping
    "grýtniþekja"                    # exploratory rock cover
  ))) %>%
  mutate(
    across(where(is.character), as.factor),
    across(where(is.logical),   as.factor)
  )

fit_all <- vegan::envfit(nmds_strict, env_all, permutations = 999, na.rm=T)

# --- envfit arrows (continuous vectors) ---
sig_thresh <- 0.05

vec_df <- tibble(variable = character(), NMDS1 = numeric(), NMDS2 = numeric(),
                 r2 = numeric(), p_value = numeric(), kind = character())

if (!is.null(fit_all$vectors) && !is.null(fit_all$vectors$arrows)) {
  vec_raw  <- as.data.frame(fit_all$vectors$arrows) %>%
    rownames_to_column("variable")
  vec_axes <- setdiff(names(vec_raw), "variable")
  if (length(vec_axes) >= 2) {
    vec_df <- vec_raw %>%
      rename(NMDS1 = all_of(vec_axes[1]), NMDS2 = all_of(vec_axes[2])) %>%
      select(variable, NMDS1, NMDS2) %>%
      mutate(
        r2      = as.numeric(fit_all$vectors$r[variable]),
        p_value = as.numeric(fit_all$vectors$pvals[variable]),
        kind    = "numeric"
      )
  }
}

# --- envfit factor centroids ---
fac_df <- tibble(variable = character(), NMDS1 = numeric(), NMDS2 = numeric(),
                 r2 = numeric(), p_value = numeric(), kind = character())

if (!is.null(fit_all$factors) && !is.null(fit_all$factors$centroids)) {
  fac_raw  <- as.data.frame(fit_all$factors$centroids) %>%
    rownames_to_column("variable")
  fac_axes <- setdiff(names(fac_raw), "variable")
  if (length(fac_axes) >= 2) {
    fac_df <- fac_raw %>%
      rename(NMDS1 = all_of(fac_axes[1]), NMDS2 = all_of(fac_axes[2])) %>%
      select(variable, NMDS1, NMDS2) %>%
      mutate(r2 = NA_real_, p_value = NA_real_, kind = "factor")
  }
}

arrow_df_all <- bind_rows(vec_df, fac_df) %>%
  filter(is.na(p_value) | p_value <= sig_thresh)

ordi_radius_all <- max(abs(c(nmds_scores$NMDS1, nmds_scores$NMDS2))) * 0.7

arrow_df_all <- arrow_df_all %>%
  mutate(xend = NMDS1 * ordi_radius_all,
         yend = NMDS2 * ordi_radius_all)

ggplot(nmds_scores, aes(x = NMDS1, y = NMDS2)) +
  geom_line(aes(group = plot_number),
            colour = "grey70", linewidth = 0.3, alpha = 0.6) +
  geom_point(aes(colour = sample, shape = sample),
             size = 2.0, alpha = 0.85) +
  geom_segment(
    data = arrow_df_all %>% filter(kind == "numeric"),
    aes(x = 0, y = 0, xend = xend, yend = yend),
    arrow = arrow(length = unit(0.2, "cm")),
    colour = "#c0392b", linewidth = 0.7, inherit.aes = FALSE
  ) +
  geom_label_repel(
    data = arrow_df_all %>% filter(kind == "numeric"),
    aes(x = xend * 1.1, y = yend * 1.1, label = variable),
    size = 2.8, colour = "#c0392b",
    fill = alpha("white", 0.7), label.size = 0,
    max.overlaps = 20, seed = 42, inherit.aes = FALSE
  ) +
  scale_colour_manual(values = c(first = "#2980b9", last = "#e67e22"),
                      labels = c(first = "First survey", last = "Last survey")) +
  scale_shape_manual(values  = c(first = 16, last = 17),
                     labels  = c(first = "First survey", last = "Last survey")) +
  labs(
    title    = "NMDS of plant community composition (Bray–Curtis) — all plots",
    subtitle = sprintf("%d plots · stress = %.3f · paired PERMANOVA R² = 2.7%%, p = 0.001",
                       length(unique(nmds_scores$plot_number)), nmds_strict$stress),
    x = "NMDS1", y = "NMDS2", colour = NULL, shape = NULL,
    caption = "Arrows: envfit vectors (p ≤ 0.05). Lines connect first and last survey of the same plot."
  ) +
  theme_classic(base_size = 11) +
  theme(legend.position = "bottom")
```

## 1.2 Outlier identification

The three plots sitting far to the right are coastal/sandy communities.
Their species composition is so distinct from the rest of the Icelandic
vegetation dataset that NMDS places them ~17 units away from the main cloud,
stretching the axis and making the central cloud unreadable.

```{r outlier-table}
# Data‑driven outlier detection on the all‑plots NMDS
nmds_scores_all <- as.data.frame(scores(nmds_strict, display = "sites")) %>%
  bind_cols(meta_strict %>% select(plot_number, sample))

# 1) Distance from the NMDS centroid
centroid <- colMeans(nmds_scores_all[, c("NMDS1", "NMDS2")])
nmds_scores_all <- nmds_scores_all %>%
  mutate(dist_center = sqrt((NMDS1 - centroid["NMDS1"])^2 +
                            (NMDS2 - centroid["NMDS2"])^2))

# 2) Flag plots whose distance is in the top tail (e.g. top 3 plots)
outlier_plots <- nmds_scores_all %>%
  arrange(desc(dist_center)) %>%
  distinct(plot_number, .keep_all = TRUE) %>%
  slice_head(n = 3) %>%                # or use a threshold on dist_center
  pull(plot_number)

# Inspect which plots were flagged
outlier_plots

meta_outliers <- meta_strict %>%
  filter(plot_number %in% outlier_plots) %>%
  select(
    plot_number, sample,
    soil_depth, grýtniþekja,
    any_of(c("vegetation_height_mean", "soil_type_name")),
    moisture_name, topography_name
  )

print_full(
  meta_outliers,
  caption = "Metadata for automatically detected outlier plots (both surveys)."
)
```

```{r taxa-shift-outliers}
# Which taxa are responsible for the outlier separation?
# Compare mean relative abundance in outlier vs non-outlier rows.
outlier_rows     <- meta_strict$plot_number %in% outlier_plots
mat_rel          <- mat_strict / rowSums(mat_strict)

taxa_shift <- tibble(
  taxon               = colnames(mat_strict),
  mean_rel_outlier    = colMeans(mat_rel[outlier_rows,  , drop = FALSE]),
  mean_rel_non_outlier = colMeans(mat_rel[!outlier_rows, , drop = FALSE])
) %>%
  mutate(delta_rel = mean_rel_outlier - mean_rel_non_outlier) %>%
  arrange(desc(abs(delta_rel)))

print_full(taxa_shift %>% slice_head(n = 30),
           caption = "Top 30 taxa driving the outlier separation (|Δ relative abundance|).")
```

In other words, these plots are outliers in the ordination **because of their taxonomic composition**, not because of extreme values in the metadata. They are dominated by coastal/sandy specialists such as `Puccinellia coarctata`, `Leymus arenarius`, `Honckenya peploides`, `Cakile maritima` and `Mertensia maritima`, which are almost absent elsewhere in the dataset. This makes their Bray–Curtis distances to the rest of the Icelandic vegetation so large that NMDS places them far away along one axis, even though their plot-level covariates (soil depth, total cover, tall-plant cover, moisture, topography) are within otherwise reasonable ranges. For this reason they are best treated as a separate coastal subset rather than as part of the main inland gradient.

---

# Part 2 — Core NMDS (outliers excluded)

```{r build-core-matrices, cache=TRUE}
keep_rows <- !meta_strict$plot_number %in% outlier_plots
meta_core <- meta_strict[keep_rows, ]
mat_core  <- mat_strict[keep_rows, ]

# Drop taxa that only occurred in the removed plots
taxa_present <- colSums(mat_core) > 0
mat_core     <- mat_core[, taxa_present]
```

After removing **`r length(outlier_plots)` plots** (`r paste(outlier_plots, collapse = ", ")`),
the core matrix has **`r nrow(mat_core)` rows** and **`r ncol(mat_core)` taxa**.
The three excluded plots are analysed separately in Part 5.

## 2.1 Re-run NMDS

```{r nmds-core, cache=TRUE}
nmds_core <- metaMDS(
  mat_core,
  distance = "bray",
  k        = 2,
  trymax   = 200,
  trace    = FALSE
)
```

**NMDS stress (core, k = 2): `r round(nmds_core$stress, 4)`**

```{r stress-interpretation, echo=FALSE, results='asis'}
s <- nmds_core$stress
interp <- dplyr::case_when(
  s < 0.10 ~ "**Excellent** — ordination is a reliable low-dimensional representation.",
  s < 0.15 ~ "**Good** — ordination is suitable for publication.",
  s < 0.20 ~ "**Acceptable** — ordination is usable; mention stress in figure caption.",
  TRUE     ~ "**Poor** — consider k = 3 or interpret with caution."
)
cat(paste0("> Stress interpretation: ", interp, "\n"))

if (s > 0.20) {
  cat("\n> ⚠️  **Stress > 0.20.** Consider re-running with `k = 3`.",
      "A 3D ordination can be visualised as a pairs plot (`pairs(scores(nmds_core))`).\n")
}
```

### 2.1b Optional: 3D NMDS (diagnostic)

```{r nmds-core-3d, cache=TRUE, fig.height=6, fig.width=6, fig.cap="3D NMDS (k = 3): pairwise site scores for the core dataset."}
# 3D NMDS to check whether a higher-dimensional solution improves stress
nmds_core_3d <- metaMDS(
  mat_core,
  distance = "bray",
  k        = 3,
  trymax   = 200,
  trace    = FALSE
)

# Fit the same environmental set on the 3D solution
fit_core_3d <- envfit(nmds_core_3d, env_core, permutations = 999)

# Inspect stress (should be noticeably lower than the 2D value)
print(nmds_core_3d$stress)

# Visualise the 3D ordination as pairwise axis plots
pairs(
  scores(nmds_core_3d, display = "sites"),
  main = sprintf("3D NMDS (k = 3): stress = %.3f", nmds_core_3d$stress)
)
```

For interpretation, we use the **2D NMDS (`nmds_core`) with `fit_core` arrows**
as the main ordination figure. The arrows show which environmental gradients
best align with the dominant compositional axes.

The **3D NMDS (`nmds_core_3d` with `fit_core_3d`) is used as a diagnostic only**:
its lower stress confirms that the inland community structure is genuinely
three‑dimensional, but we visualise it via a pairs plot without envfit arrows.
Adding vectors directly in 3D would require multiple views or interactive
graphics and tends to obscure rather than clarify the main patterns, so all
arrow-based interpretation remains on the 2D biplot.


### Shepard plot

```{r shepard, fig.height=5, fig.width=5, fig.cap="Shepard diagram: ordination distances vs. dissimilarities. Points should follow the step function tightly."}
stressplot(nmds_core,
           main = sprintf("Shepard plot — stress %.4f", nmds_core$stress))
```

## 2.2 Re-fit environmental vectors

```{r envfit-core, cache=TRUE}
env_core <- meta_core %>%
  select(all_of(vars_driver_strict_final)) %>%
  mutate(
    across(where(is.character), as.factor),
    across(where(is.logical),   as.factor)
  )

fit_core <- envfit(nmds_core, env_core, permutations = 999)
```

```{r envfit-table}
fit_tbl_core <- bind_rows(
  tibble(
    variable = names(fit_core$vectors$r),
    r2       = as.numeric(fit_core$vectors$r),
    p_value  = as.numeric(fit_core$vectors$pvals),
    kind     = "numeric"
  ),
  if (!is.null(fit_core$factors)) tibble(
    variable = names(fit_core$factors$r),
    r2       = as.numeric(fit_core$factors$r),
    p_value  = as.numeric(fit_core$factors$pvals),
    kind     = "factor"
  )
) %>%
  arrange(p_value) %>%
  mutate(sig = case_when(p_value < 0.001 ~ "***",
                         p_value < 0.01  ~ "**",
                         p_value < 0.05  ~ "*",
                         TRUE            ~ "ns"))

print_full(fit_tbl_core,
           caption = "envfit results on core dataset (outliers removed). r² = goodness-of-fit.")
```

## 2.3 NMDS biplot

```{r nmds-core-plot, fig.height=7.5, fig.cap=sprintf("NMDS biplot of the core %d-plot dataset. Red arrows are significant envfit vectors (p ≤ 0.05); length is proportional to r². Grey lines connect first and last survey of the same plot.",length(unique(meta_core$plot_number)))}

# --- Build arrow data frame ---
sig_thresh <- 0.05

arrow_df_core <- as.data.frame(fit_core$vectors$arrows) %>%
  rownames_to_column("variable") %>%
  rename(ax1 = NMDS1, ax2 = NMDS2) %>%
  mutate(
    r2      = fit_core$vectors$r,
    p_value = fit_core$vectors$pvals
  ) %>%
  filter(p_value <= sig_thresh)

# Factor centroids
factor_df_core <- tryCatch({
  as.data.frame(scores(fit_core, display = "factors")) %>%
    rownames_to_column("label") %>%
    rename(fax1 = NMDS1, fax2 = NMDS2) %>%
    mutate(
      p_value = fit_core$factors$pvals[
        sapply(label, function(nm) {
          matched <- names(fit_core$factors$pvals)[startsWith(nm, names(fit_core$factors$pvals))]
          if (length(matched)) matched[1] else NA_character_
        })
      ]
    ) %>%
    filter(!is.na(p_value), p_value <= sig_thresh)
}, error = function(e) tibble())

# --- Site scores ---
site_scores <- as.data.frame(scores(nmds_core, display = "sites")) %>%
  bind_cols(meta_core %>% select(plot_number, sample))

ordi_radius_core <- max(abs(c(site_scores$NMDS1, site_scores$NMDS2))) * 0.70

arrow_df_core <- arrow_df_core %>%
  mutate(xend = ax1 * ordi_radius_core,
         yend = ax2 * ordi_radius_core)

# --- Plot ---
p_nmds_core <- ggplot(site_scores, aes(x = NMDS1, y = NMDS2)) +
  geom_line(aes(group = plot_number),
            colour = "grey60", linewidth = 0.30, alpha = 0.55) +
  geom_point(aes(colour = sample, shape = sample),
             size = 2.0, alpha = 0.85) +
  geom_segment(
    data = arrow_df_core,
    aes(x = 0, y = 0, xend = xend, yend = yend),
    arrow = arrow(length = unit(0.22, "cm"), type = "closed"),
    colour = "#c0392b", linewidth = 0.75, inherit.aes = FALSE
  ) +
  geom_label_repel(
    data        = arrow_df_core,
    aes(x = xend * 1.12, y = yend * 1.12, label = variable),
    size        = 3.0, colour = "#c0392b",
    fill        = alpha("white", 0.80), label.size = 0,
    max.overlaps = 20, seed = 42, inherit.aes = FALSE
  ) +
  {
    if (nrow(factor_df_core) > 0)
      geom_label_repel(
        data        = factor_df_core,
        aes(x = fax1, y = fax2, label = label),
        size        = 2.5, colour = "#27ae60",
        fill        = alpha("white", 0.80), label.size = 0,
        max.overlaps = 15, seed = 43, inherit.aes = FALSE
      )
  } +
  scale_colour_manual(values = c(first = "#2980b9", last = "#e67e22"),
                      labels = c(first = "First survey", last = "Last survey")) +
  scale_shape_manual(values  = c(first = 16, last = 17),
                     labels  = c(first = "First survey", last = "Last survey")) +
  coord_equal() +
  labs(
    title    = "NMDS — core dataset (Bray–Curtis)",
    subtitle = sprintf(
      "%d plots · stress = %.3f · excl. %s",
      length(unique(site_scores$plot_number)),
      nmds_core$stress,
      paste(outlier_plots, collapse = ", ")
    ),
    x = "NMDS1", y = "NMDS2", colour = NULL, shape = NULL,
    caption = "Red arrows: envfit vectors (p ≤ 0.05); length ∝ r².\nGrey lines connect first and last survey of the same plot."
  ) +
  theme_classic(base_size = 11) +
  theme(
    legend.position = "bottom",
    plot.subtitle   = element_text(size = 8, colour = "grey40"),
    plot.caption    = element_text(size = 7.5, colour = "grey50", hjust = 0)
  )

p_nmds_core
```


Figure X. NMDS ordination of the core dataset (outliers removed) with envfit vectors/centroids for key environmental drivers. Soil depth and soil type show the strongest and highly significant associations with community composition (r² ≈ 0.32–0.35, p = 0.001), followed by moisture class (r² ≈ 0.22, p = 0.001), whereas topography and permafrost have weak, non‑significant fits (r² ≤ 0.02, p ≥ 0.05).
---

# Part 3 — Per-plot compositional turnover

## 3.1 Bray–Curtis distance per plot (first → last)

```{r bc-per-plot}
bc_matrix <- as.matrix(vegdist(mat_strict, method = "bray"))

plot_bc <- meta_strict %>%
  mutate(row_idx = row_number()) %>%
  group_by(plot_number) %>%
  summarise(
    idx_first = row_idx[sample == "first"],
    idx_last  = row_idx[sample == "last"],
    .groups   = "drop"
  ) %>%
  mutate(bc_dist = map2_dbl(idx_first, idx_last, ~ bc_matrix[.x, .y]))

env_first <- meta_strict %>%
  filter(sample == "first") %>%
  select(plot_number, any_of("soil_type_name"), any_of("moisture_name"),
         soil_depth, grýtniþekja)

plot_bc_env <- plot_bc %>%
  left_join(env_first, by = "plot_number")
```

```{r bc-summary-table}
bc_summary <- plot_bc_env %>%
  group_by(across(any_of("soil_type_name"))) %>%
  summarise(
    n         = n(),
    median_bc = round(median(bc_dist, na.rm = TRUE), 3),
    mean_bc   = round(mean(bc_dist,   na.rm = TRUE), 3),
    sd_bc     = round(sd(bc_dist,     na.rm = TRUE), 3),
    .groups   = "drop"
  ) %>%
  arrange(desc(median_bc))

print_full(bc_summary,
           caption = "Per-plot Bray–Curtis distance (first → last) summarised by soil type.")
```

```{r bc-summary-table-moisture}
bc_summary_moist <- plot_bc_env %>%
group_by(moisture_name) %>%
summarise(
n = n(),
median_bc = round(median(bc_dist, na.rm = TRUE), 3),
mean_bc = round(mean(bc_dist, na.rm = TRUE), 3),
sd_bc = round(sd(bc_dist, na.rm = TRUE), 3),
.groups = "drop"
) %>%
arrange(desc(median_bc))
print_full(
bc_summary_moist,
caption = "Per-plot Bray–Curtis distance (first → last) summarised by moisture class."
)
```


“Summary of per‑plot Bray–Curtis distance between first and last survey across all 160 plots (higher values indicate greater compositional change); shown are sample size (n), median, mean, and standard deviation of Bray–Curtis distances.”

## 3.2 Kruskal–Wallis tests

```{r kruskal-wallis}
if ("soil_type_name" %in% names(plot_bc_env)) {
  kw_soil <- kruskal.test(bc_dist ~ soil_type_name, data = plot_bc_env)
} else {
  kw_soil <- NULL
}

kw_moist <- kruskal.test(bc_dist ~ moisture_name,  data = plot_bc_env)

kw_tbl <- tibble(
  test      = c("Soil type", "Moisture class"),
  chi_sq    = c(if (!is.null(kw_soil)) kw_soil$statistic  else NA_real_,
                kw_moist$statistic),
  df        = c(if (!is.null(kw_soil)) kw_soil$parameter  else NA_real_,
                kw_moist$parameter),
  p_value   = c(if (!is.null(kw_soil)) kw_soil$p.value    else NA_real_,
                kw_moist$p.value),
  sig       = case_when(p_value < 0.001 ~ "***",
                        p_value < 0.01  ~ "**",
                        p_value < 0.05  ~ "*",
                        TRUE            ~ "ns")
)

print_full(kw_tbl,
           caption = "Kruskal–Wallis: does soil type / moisture predict magnitude of compositional change?")
```

update table caption

```{r kw-followup, results='asis'}
if (!is.null(kw_soil) && kw_soil$p.value < 0.05) {
  cat("> **Soil type is significant (p < 0.05).** Consider a post-hoc Dunn test:\n")
  cat("> `dunn.test::dunn.test(plot_bc_env$bc_dist, plot_bc_env$soil_type_name, method = 'bh')`\n\n")
}
if (kw_moist$p.value < 0.05) {
  cat("> **Moisture class is significant (p < 0.05).** Consider a post-hoc Dunn test:\n")
  cat("> `dunn.test::dunn.test(plot_bc_env$bc_dist, plot_bc_env$moisture_name, method = 'bh')`\n\n")
}
```

## 3.3 Boxplot by soil type

```{r bc-boxplot, fig.cap="Per-plot Bray–Curtis distance (first → last survey) by soil type. Points are individual plots."}
plot_bc_soil <- plot_bc_env %>%
  dplyr::filter(!is.na(soil_type_name))

# Inspect
dplyr::glimpse(plot_bc_soil)
dplyr::count(plot_bc_soil, soil_type_name)
summary(plot_bc_soil$bc_dist)
if ("soil_type_name" %in% names(plot_bc_env)) {
  p_bc_soil <- ggplot(
    plot_bc_soil,
    aes(x = reorder(soil_type_name, bc_dist, FUN = median),
        y = bc_dist, fill = soil_type_name)
  ) +
    geom_boxplot(alpha = 0.7, outlier.size = 1.2) +
    geom_jitter(width = 0.15, alpha = 0.4, size = 1) +
    labs(
      title    = "Compositional turnover by soil type",
      subtitle = "Per-plot Bray–Curtis distance (first → last survey)",
      x = "Soil type", y = "Bray–Curtis distance", fill = NULL
    ) +
    theme_classic(base_size = 11) +
    theme(axis.text.x = element_text(angle = 35, hjust = 1),
          legend.position = "none")
} else {
  p_bc_soil <- ggplot() +
    theme_void() +
    ggtitle("Soil type not available in meta_strict / plot_bc_env")
}

p_bc_soil
```
Per‑plot Bray–Curtis distance (first → last survey) by soil type for 159 revisited plots. Boxes show median and interquartile range of compositional change within each soil type (Leirujörð, Lífræn jörð, Klapparjörð, Áfoksjörð, Melajörð, Sandjörð, Eyrarjörð), with individual plots as points; overall, turnover is substantial (median ≈ 0.61, mean ≈ 0.58, range 0.16–1.00), and some soil types (e.g. Áfoksjörð, Klapparjörð, Sandjörð) exhibit higher typical change than others (e.g. Leirujörð).

```{r pca-turnover-drivers, fig.height=6, fig.width=8}


bc_matrix <- as.matrix(vegdist(mat_strict, method = "bray"))
plot_bc <- meta_strict %>%
mutate(row_idx = dplyr::row_number()) %>%
dplyr::group_by(plot_number) %>%
dplyr::summarise(
idx_first = row_idx[sample == "first"],
idx_last = row_idx[sample == "last"],
.groups = "drop"
) %>%
dplyr::mutate(bc_dist = purrr::map2_dbl(idx_first, idx_last, ~ bc_matrix[.x, .y]))

env_first_pca <- meta_strict %>%
dplyr::filter(sample == "first") %>%
dplyr::select(
plot_number,
háplöntuþekja,
vegetation_height_mean,
total_cover,
soil_depth
)
pca_dat <- plot_bc %>%
dplyr::left_join(env_first_pca, by = "plot_number") %>%
dplyr::filter(stats::complete.cases(háplöntuþekja, vegetation_height_mean,
total_cover, soil_depth))

pca_env <- prcomp(
pca_dat %>%
dplyr::select(háplöntuþekja, vegetation_height_mean, total_cover, soil_depth),
center = TRUE,
scale. = TRUE
)
pca_scores <- as.data.frame(pca_env$x) %>%
dplyr::mutate(
plot_number = pca_dat$plot_number,
bc_dist = pca_dat$bc_dist
)
pca_loadings <- as.data.frame(pca_env$rotation) %>%
tibble::rownames_to_column("variable")

var_expl <- 100 * (pca_env$sdev^2) / sum(pca_env$sdev^2)
ggplot(pca_scores, aes(x = PC1, y = PC2, colour = bc_dist)) +
geom_point(size = 2, alpha = 0.85) +
scale_colour_viridis_c(option = "C", name = "Bray–Curtis\n(first → last)") +
geom_segment(
data = pca_loadings,
aes(x = 0, y = 0, xend = PC1 * 2.5, yend = PC2 * 2.5),
arrow = arrow(length = unit(0.2, "cm")),
inherit.aes = FALSE,
colour = "#c0392b",
linewidth = 0.7
) +
geom_label_repel(
data = pca_loadings,
aes(x = PC1 * 2.7, y = PC2 * 2.7, label = variable),
inherit.aes = FALSE,
size = 3,
colour = "#c0392b",
fill = alpha("white", 0.8),
label.size = 0,
max.overlaps = 15,
seed = 44
) +
labs(
title = "PCA of environmental drivers (first survey)",
subtitle = sprintf("PC1: %.1f%%, PC2: %.1f%% of variance",
var_expl[1], var_expl[2]),
x = "PC1",
y = "PC2"
) +
theme_classic(base_size = 11) +
theme(
legend.position = "right",
plot.subtitle = element_text(size = 8, colour = "grey40")
)

```
Principal components analysis of first‑survey environmental drivers (háplöntuþekja, vegetation_height_mean, total_cover, soil_depth) for the 159 core plots; points are plots positioned by PC1 and PC2, coloured by their Bray–Curtis compositional change (first → last), and red arrows show variable loadings (PC1 explaining 62.8% and PC2 16.9% of total variance).


---

# Part 4 — Indicator species analysis (IndVal)

Using `indicspecies::multipatt()` to identify taxa that are significantly
associated with either the first or last survey, using IndVal.g with 999
permutations. P-values are Benjamini–Hochberg FDR-corrected.

```{r indval, cache=TRUE}
group_vec <- ifelse(meta_strict$sample == "first", 1L, 2L)

indval_result <- multipatt(
  mat_strict,
  cluster = group_vec,
  control = how(nperm = 999),
  func    = "IndVal.g",
  duleg   = TRUE
)

indval_tbl <- indval_result$sign %>%
  rownames_to_column("taxon") %>%
  as_tibble() %>%
  rename(group = index) %>%
  mutate(
    survey = if_else(group == 1L, "first", "last"),
    p_adj  = p.adjust(p.value, method = "BH")
  ) %>%
  filter(p_adj <= 0.05) %>%
  arrange(p_adj)
```

**`r nrow(indval_tbl)` taxa** pass FDR ≤ 5%:
`r sum(indval_tbl$survey == "first")` associated with the **first** survey,
`r sum(indval_tbl$survey == "last")` with the **last** survey.

## 4.1 Full indicator taxa table

```{r indval-table}
print_full(
  indval_tbl %>%
    select(taxon, survey, stat, p.value, p_adj) %>%
    mutate(sig = case_when(p_adj < 0.001 ~ "***",
                           p_adj < 0.01  ~ "**",
                           p_adj < 0.05  ~ "*",
                           TRUE          ~ "ns")),
  caption = "Indicator taxa (IndVal.g, FDR ≤ 5%). 'first' = associated with first survey; 'last' = last survey."
)
```

Indicator species (IndVal.g) for first vs last survey based on the core inland dataset. Columns show the taxon name, its association with first (s.1) and last (s.2) surveys, the group it favors (1 = first, 2 = last), IndVal statistic (stat), raw p‑value (p.value), survey group label, and Benjamini–Hochberg FDR‑adjusted p‑value (p_adj); all taxa listed here are significant at FDR ≤ 5%.

## 4.2 Top 15 per survey — bar chart

```{r indval-plot, fig.height=7, fig.cap="Top 15 indicator taxa per survey by IndVal.g statistic (FDR ≤ 5%)."}
top_indval <- indval_tbl %>%
  group_by(survey) %>%
  slice_min(order_by = p_adj, n = 15, with_ties = FALSE) %>%
  ungroup()

if (nrow(top_indval) > 0 && length(unique(top_indval$survey)) > 0) {
  ggplot(top_indval,
         aes(x = stat, y = reorder(taxon, stat), fill = survey)) +
    geom_col(alpha = 0.85) +
    facet_wrap(
      ~survey,
      scales  = "free_y",
      labeller = as_labeller(c(first = "First survey", last = "Last survey"))
    ) +
    scale_fill_manual(values = c(first = "#2980b9", last = "#e67e22")) +
    labs(
      title    = "Indicator taxa — first vs last survey",
      subtitle = "IndVal.g statistic; top 15 per group (FDR ≤ 5%)",
      x = "IndVal statistic", y = NULL, fill = NULL
    ) +
    theme_classic(base_size = 10) +
    theme(legend.position = "none",
          strip.background = element_rect(fill = "grey92", colour = NA))
} else {
  ggplot() +
    theme_void() +
    ggtitle("No indicator taxa to display (top_indval is empty)")
}
```

---

# Part 5 — Univariate change in community descriptors

Paired Wilcoxon signed-rank test for each continuous driver variable.
Each plot contributes one matched pair (first survey value, last survey value).

```{r univar-wilcox}
univar_vars <- intersect(
  c("háplöntuþekja", "vegetation_height_mean", "total_cover", "soil_depth"),
  names(meta_strict)
)

env_paired_wide <- meta_strict %>%
  select(plot_number, sample, all_of(univar_vars)) %>%
  pivot_wider(
    names_from  = sample,
    values_from = all_of(univar_vars),
    names_sep   = "_"
  )

wilcox_results <- map_dfr(univar_vars, function(v) {
  first_col    <- paste0(v, "_first")
  last_col     <- paste0(v, "_last")
  x            <- env_paired_wide[[first_col]]
  y            <- env_paired_wide[[last_col]]
  complete_idx <- complete.cases(x, y)
  n_pairs      <- sum(complete_idx)
  test <- wilcox.test(y[complete_idx], x[complete_idx],
                      paired = TRUE, conf.int = TRUE)
  tibble(
    variable      = v,
    n_pairs       = n_pairs,
    median_first  = median(x[complete_idx], na.rm = TRUE),
    median_last   = median(y[complete_idx], na.rm = TRUE),
    median_change = median(y[complete_idx] - x[complete_idx], na.rm = TRUE),
    ci_low        = test$conf.int[1],
    ci_high       = test$conf.int[2],
    W             = test$statistic,
    p_value       = test$p.value
  )
}) %>%
  mutate(
    p_adj = p.adjust(p_value, method = "BH"),
    sig   = case_when(p_adj < 0.001 ~ "***",
                      p_adj < 0.01  ~ "**",
                      p_adj < 0.05  ~ "*",
                      TRUE          ~ "ns")
  )
```

## 5.1 Results table

```{r univar-table}
print_full(wilcox_results,
           caption = "Paired Wilcoxon signed-rank test results. median_change = last − first. BH-adjusted p-values.")
```

## 5.2 Change distributions

```{r univar-plot, fig.cap="Distribution of per-plot change (last − first) for each community descriptor. Significance stars are BH-adjusted."}
env_paired_long_change <- map_dfr(univar_vars, function(v) {
  first_col <- paste0(v, "_first")
  last_col  <- paste0(v, "_last")
  env_paired_wide %>%
    transmute(
      plot_number = plot_number,
      variable    = v,
      change      = .data[[last_col]] - .data[[first_col]]
    )
}) %>%
  filter(!is.na(change)) %>%
  left_join(wilcox_results %>% select(variable, p_adj, sig), by = "variable")

y_top <- max(env_paired_long_change$change, na.rm = TRUE) * 1.07

ggplot(env_paired_long_change, aes(x = variable, y = change)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_violin(fill = "#bdc3c7", alpha = 0.5, trim = TRUE) +
  geom_boxplot(width = 0.18, fill = "white", outlier.size = 0.8) +
  geom_text(
    data = env_paired_long_change %>% distinct(variable, sig),
    aes(y = y_top, label = sig),
    size = 5, colour = "#c0392b"
  ) +
  labs(
    title    = "Change in continuous community descriptors (last − first)",
    subtitle = "Paired Wilcoxon; * p<0.05, ** p<0.01, *** p<0.001 (BH-adjusted)",
    x = NULL, y = "Change (last − first)"
  ) +
  theme_classic(base_size = 11)
```

---

# Part 6 — Excluded coastal plots

With only 6 rows (3 plots × 2 surveys), NMDS is numerically unstable.
PCoA (metric multidimensional scaling) is used instead.

```{r coastal-pcoa, fig.cap="PCoA of the three excluded coastal/outlier plots. Labels show plot name and survey. Lines connect first and last survey of the same plot."}

outlier_rows <- meta_strict$plot_number %in% outlier_plots
mat_coastal  <- mat_strict[outlier_rows, ]
meta_coastal <- meta_strict[outlier_rows, ]
mat_coastal  <- mat_coastal[, colSums(mat_coastal) > 0]

pcoa_coastal <- cmdscale(
  vegdist(mat_coastal, method = "bray"),
  k   = 2,
  eig = TRUE
)

pcoa_df <- as.data.frame(pcoa_coastal$points) %>%
  setNames(c("Axis1", "Axis2")) %>%
  bind_cols(meta_coastal %>% select(plot_number, sample))

var_explained <- round(
  100 * pcoa_coastal$eig[1:2] / sum(abs(pcoa_coastal$eig)), 1
)

ggplot(pcoa_df, aes(x = Axis1, y = Axis2)) +
  geom_line(aes(group = plot_number), colour = "grey60", linewidth = 0.4) +
  geom_point(aes(colour = sample, shape = sample), size = 3) +
  geom_label_repel(
    aes(label = paste(plot_number, sample)),
    size = 3, max.overlaps = 20, seed = 1
  ) +
  scale_colour_manual(values = c(first = "#2980b9", last = "#e67e22"),
                      labels = c(first = "First survey", last = "Last survey")) +
  scale_shape_manual(values  = c(first = 16, last = 17),
                     labels  = c(first = "First survey", last = "Last survey")) +
  coord_equal() +
  labs(
    title    = "PCoA — excluded coastal/outlier plots",
    subtitle = sprintf("Bray–Curtis · Axis 1 = %.1f%% · Axis 2 = %.1f%%",
                       var_explained[1], var_explained[2]),
    x = sprintf("PCoA Axis 1 (%.1f%%)", var_explained[1]),
    y = sprintf("PCoA Axis 2 (%.1f%%)", var_explained[2]),
    colour = NULL, shape = NULL
  ) +
  theme_classic(base_size = 11) +
  theme(legend.position = "bottom")
```

```{r coastal-species}
# Top species in each coastal plot (last survey only, for brevity)
coastal_last <- meta_coastal %>%
  filter(sample == "last") %>%
  pull(plot_number)

coastal_species <- community_long_paired %>%
  filter(plot_number %in% coastal_last, sample == "last") %>%
  arrange(plot_number, desc(cover)) %>%
  group_by(plot_number) %>%
  slice_head(n = 10) %>%
  ungroup()

print_full(coastal_species,
           caption = "Top 10 species by cover in each coastal plot (last survey).")
```

---

# Part 7 — Suggested next steps

Based on the results above, the following analyses are the logical
continuation — choose based on what the figures showed.

```{r next-steps, echo=FALSE, results='asis'}
cat(paste(
  "**A — dbRDA (distance-based redundancy analysis)**",
  "If háplöntuþekja or vegetation height changed significantly in Part 5, use",
  "dbRDA to formally partition variance between the temporal effect (*sample*)",
  "and the environmental gradient, while conditioning on plot identity:",
  "",
  "```r",
  "vegan::dbrda(",
  "  mat_strict ~ sample + háplöntuþekja + Condition(plot_number),",
  "  data   = meta_strict,",
  "  method = \"bray\"",
  ")",
  "```",
  "",
  "**B — Stratum-specific PERMANOVAs**",
  "If soil type or moisture was significant in the Kruskal–Wallis (Part 3),",
  "split the dataset by those classes and re-run paired PERMANOVA within each",
  "stratum — this tests whether the temporal signal is consistent across",
  "habitat types or concentrated in one.",
  "",
  "**C — Taxon-level ecology**",
  "Review the indicator taxa from Part 4 against Icelandic vegetation",
  "literature. Key questions:",
  "- Are the 'last survey' gainers typically associated with warming, reduced",
  "  grazing, or nitrogen deposition in Icelandic heathlands?",
  "- Do *Racomitrium* or *Dryas* appear as consistent losers?",
  "- Are wetland sedges (*Carex lyngbyei*, *C. nigra*) showing directional change?",
  "",
  "**D — Spatial autocorrelation of turnover**",
  "Test whether compositional change (the per-plot Bray–Curtis from Part 3) is",
  "spatially structured using a Mantel test:",
  "",
  "```r",
  "geo_dist <- dist(cbind(env_first$longitude_1, env_first$latitude_1))",
  "vegan::mantel(bc_dist_matrix, geo_dist, method = \"pearson\", permutations = 999)",
  "```",
  sep = "\n"
))
```

---

# Session info

```{r session-info}
sessionInfo()
```
