---
title: "v_plants variable screening"
author: "Valtýr"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(lubridate)
```

```{r source-prep}
# Shared computational backbone used across reports
source("v_plants_prep.R")
```

**Note on drift control.** The objects created by `v_plants_prep.R` are the single source of truth.  
Chunks below are for transparency and diagnostics; they do not overwrite backbone objects.

## Purpose

This note documents **how variable exclusion and environmental variable selection work** in the vegetation pipeline, as implemented in `v_plants_prep.R`. It serves as an audit trail so future analyses can see which variables are excluded, which are used as env candidates, and how the strict driver set is formed.

## Columns excluded (from `v_plants_prep.R`)

Exclusion lists are defined in `v_plants_prep.R`. They are either **effectively unmeasured (~all missing)** or **free-text / images** not used as structured covariates. Displayed here for reference only.

```{r excluded_columns_display}
# From v_plants_prep.R: vars_exclude_missing, vars_exclude_text, vars_exclude_strict
list(
  vars_exclude_missing = vars_exclude_missing,
  vars_exclude_text = vars_exclude_text,
  vars_exclude_strict = vars_exclude_strict
)
```

## Reading and filtering (from `v_plants_prep.R`)

Data flow in prep:

1. **Raw** `v_plants_raw`: from `plants_export/v_plants_allt.csv`; optionally restricted by **`heiti_include`** (plot allow-list).
2. **Taxa exclusion**: rows with `taxon_name` listed in `v_plants_taxa_exclude_flags.csv` (column 2 where column 1 is flagged) are dropped.
3. **Subplot filter**: main pipeline uses only subplots `1`–`8`; rows with `subplot_number` in `0`, `"Utan reita"`, `"utan reita"` are dropped for community/ordination (retained only for diversity-with-outside-subplot views).
4. **Derived**: `vegetation_height_mean` = mean of `vegetation_height_1`–`4`; `field_date_parsed`, `year`.
5. **Excluded columns**: `vars_exclude_strict` dropped from `v_plants_use`.

```{r read_and_filter_v_plants}
# Diagnostic: object sizes from v_plants_prep.R
tibble(
  object = c("v_plants_raw", "v_plants2", "v_plants_use"),
  n_rows = c(nrow(v_plants_raw), nrow(v_plants2), nrow(v_plants_use)),
  n_cols = c(ncol(v_plants_raw), ncol(v_plants2), ncol(v_plants_use))
)
```

## Variable catalog and env selection (from `v_plants_prep.R`)

**Variable catalog**: `var_catalog` is built in prep from `v_plants_use` (type, prop_missing, n_unique, analysis_role). Shown here for reference.

**Environmental variable selection**: In prep, which columns are used as env candidates is controlled by **`v_plants_envvar_selection.csv`**: column 2 = variable name, column 1 = flag (e.g. `x`) to include. Prep sets `vars_env_from_sheet` from that file. The four raw height columns, if flagged, are replaced by a single **`vegetation_height_mean`** (mean of `vegetation_height_1`–`4`). Then:

- `vars_env_numeric` = variables in `vars_env_from_sheet` that are numeric in `var_catalog`
- `vars_env_factor` = variables in `vars_env_from_sheet` that are character/factor/logical

Prep also adds **time gap** and **altitude** (when available) to the env data:

- **`gap_years`**: time between first and last survey (from `plot_years`); joined into `env_candidates_ps` and `meta_strict`.
- **`haeddypimetrar`**: elevation in metres from `snid_with_elevation.gpkg` (when `sf` and the gpkg exist); joined into `env_candidates_ps`, `meta_strict`, and `bc_enriched`. Many plots may have `NA` if not in the gpkg.

So env candidates are **sheet-driven** plus these two covariates.

```{r categorize_variables}
# From v_plants_prep.R: var_catalog (no re-definition)
var_catalog %>%
  arrange(analysis_role, desc(prop_missing), variable)
```

```{r env_selection_from_prep}
# From v_plants_prep.R: sheet-driven env candidates (do not overwrite)
list(
  vars_env_from_sheet_n = length(vars_env_from_sheet),
  vars_env_from_sheet = vars_env_from_sheet,
  vars_env_numeric = vars_env_numeric,
  vars_env_factor = vars_env_factor
)
```

```{r variable_groups_for_downstream_use}
# Display only: variable groups from prep (catalog roles still available for reference)
var_catalog %>%
  count(analysis_role, sort = TRUE)
```

## Paired-design essentials (from `v_plants_prep.R`)

Prep defines revisited plots (`n_years >= 2`), `year_first` / `year_last`, `sample` = first/last, harmonized cover, and complete pairs (two rows per plot) for `community_wide`, `meta`, `mat`.

```{r mirrored_paired_design_steps}
# Diagnostic-only: report paired-design objects produced by v_plants_prep.R
diag_counts <- list(
  n_v_plants2 = nrow(v_plants2),
  n_plot_years = nrow(plot_years),
  n_revisited_plots = nrow(revisited),
  n_rows_paired_long = nrow(paired_long),
  n_rows_paired_cover = nrow(paired_cover),
  n_rows_community_long = nrow(community_long),
  n_rows_complete_pairs = nrow(community_long_paired),
  n_plots_complete_pairs = dplyr::n_distinct(community_long_paired$plot_number)
)
diag_counts
```

## First analysis steps

Diagnostics using `community_wide`, `meta`, `mat`, and `env_candidates_ps` from `v_plants_prep.R` (env columns = sheet-driven `vars_env_numeric` / `vars_env_factor`).

```{r first_steps_analysis}
# Matrix readiness and env missingness (prep objects)
has_rows <- nrow(mat) > 0
two_rows_per_plot <- has_rows && all((meta %>% count(plot_number) %>% pull(n)) == 2)

readiness <- tibble(
  check = c(
    "matrix_rows",
    "matrix_taxa_columns",
    "all_plots_have_two_rows"
  ),
  value = c(
    nrow(mat),
    ncol(mat),
    as.integer(two_rows_per_plot)
  )
)
readiness

# First-pass paired composition and env missingness
if (has_rows) {
  bray <- vegan::vegdist(mat, method = "bray")
  bray_summary <- summary(as.numeric(bray))
  bray_summary
} else {
  message("Skipping Bray-Curtis summary because matrix_rows == 0.")
}

env_candidates_ps %>%
  summarise(across(all_of(c(vars_env_numeric, vars_env_factor)), ~ mean(is.na(.x)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "prop_missing") %>%
  arrange(desc(prop_missing))
```

The table below is a **diagnostic** screen (missingness and first-vs-last shift). The strict driver set actually used in ordination and PERMANOVA is **`vars_driver_strict_final`** from `v_plants_prep.R` (curated list intersected with `env_candidates_ps`).

```{r driver_variable_eligibility}
# Thresholds for strict driver set (edit as needed)
max_missing_strict <- 0.35
max_shift_pp_strict <- 10

candidate_vars <- intersect(c(vars_env_numeric, vars_env_factor), names(env_candidates_ps))

driver_eligibility <- purrr::map_dfr(candidate_vars, function(v) {
  x <- env_candidates_ps %>%
    mutate(is_miss = is.na(.data[[v]])) %>%
    group_by(sample) %>%
    summarise(prop_miss = mean(is_miss), .groups = "drop") %>%
    tidyr::pivot_wider(names_from = sample, values_from = prop_miss, values_fill = 0)

  miss_first <- x$first %||% NA_real_
  miss_last <- x$last %||% NA_real_
  delta_pp <- 100 * (miss_last - miss_first)
  prop_missing <- mean(is.na(env_candidates_ps[[v]]))

  tibble(
    variable = v,
    prop_missing = prop_missing,
    miss_first = miss_first,
    miss_last = miss_last,
    delta_pp = delta_pp,
    strict_ok = prop_missing <= max_missing_strict && abs(delta_pp) <= max_shift_pp_strict,
    exploratory_ok = prop_missing <= 0.60
  )
}) %>%
  mutate(
    use_class = case_when(
      strict_ok ~ "strict",
      exploratory_ok ~ "exploratory_only",
      TRUE ~ "exclude"
    )
  ) %>%
  arrange(use_class, desc(prop_missing), desc(abs(delta_pp)), variable)

driver_eligibility

vars_driver_strict <- driver_eligibility %>%
  filter(use_class == "strict") %>%
  pull(variable)

vars_driver_exploratory <- driver_eligibility %>%
  filter(use_class == "exploratory_only") %>%
  pull(variable)

vars_driver_exclude <- driver_eligibility %>%
  filter(use_class == "exclude") %>%
  pull(variable)

list(
  vars_driver_strict = vars_driver_strict,
  vars_driver_exploratory = vars_driver_exploratory,
  vars_driver_exclude = vars_driver_exclude
)
```

```{r strict_driver_set_curated}
# Curated strict driver set (aligned with v_plants_prep.R):
# - use vegetation_height_mean (derived in prep), not raw height columns
# - prefer name-coded factors over *_id for interpretability
vars_driver_strict_proposed <- c(
  "háplöntuþekja",
  "soil_depth",
  "total_cover",
  "vegetation_height_mean",
  "soil_type_name",
  "moisture_name",
  "topography_name",
  "permafrost"
)

# Keep only variables that exist in the current filtered data
vars_driver_strict_proposed <- intersect(vars_driver_strict_proposed, names(env_candidates_ps))

# Diagnostic-only object (avoid overwriting prep backbone objects)
env_strict_diag <- env_candidates_ps %>%
  select(plot_number, sample, all_of(vars_driver_strict_proposed)) %>%
  distinct()

# Diagnostics: level counts for categorical vars, SD for numeric vars (by name to avoid empty results)
driver_vars <- setdiff(vars_driver_strict_proposed, c("plot_number", "sample"))
num_vars <- driver_vars[driver_vars %in% names(env_strict_diag) & purrr::map_lgl(env_strict_diag[driver_vars], is.numeric)]
fac_vars <- driver_vars[driver_vars %in% names(env_strict_diag) & purrr::map_lgl(env_strict_diag[driver_vars], ~ is.character(.x) || is.factor(.x) || is.logical(.x))]

factor_checks <- if (length(fac_vars) > 0) {
  env_strict_diag %>%
    summarise(across(all_of(fac_vars), ~ n_distinct(.x, na.rm = TRUE))) %>%
    pivot_longer(everything(), names_to = "variable", values_to = "n_levels")
} else {
  tibble(variable = character(), n_levels = integer())
}

numeric_checks <- if (length(num_vars) > 0) {
  env_strict_diag %>%
    summarise(across(all_of(num_vars), ~ sd(.x, na.rm = TRUE))) %>%
    pivot_longer(everything(), names_to = "variable", values_to = "sd")
} else {
  tibble(variable = character(), sd = numeric())
}

list(
  vars_driver_strict_final_from_prep = vars_driver_strict_final,
  vars_driver_strict_proposed = vars_driver_strict_proposed,
  factor_checks = factor_checks,
  numeric_checks = numeric_checks
)
```

```{r strict_complete_case_paired_readiness}
# Diagnostic-only: readiness for strict objects from v_plants_prep.R
strict_readiness <- tibble(
  n_plots = n_distinct(meta_strict$plot_number),
  n_rows = nrow(meta_strict),
  n_taxa = ncol(mat_strict),
  paired_ok = all((meta_strict %>% count(plot_number) %>% pull(n)) == 2)
)

strict_readiness
```

```{r strict_permanova_and_dispersion}
# From v_plants_prep.R: perm_strict, bd_strict, bd_test_strict (no re-run)
perm_strict
bd_test_strict
```

```{r strict_envfit_screening}
# Driver screening with envfit on strict set (diagnostic table only).
# Uses nmds_strict and fit_strict objects created in v_plants_prep.R.

# Numeric vectors
fit_tbl_numeric <- tibble(
  variable = names(fit_strict$vectors$r),
  r2 = as.numeric(fit_strict$vectors$r),
  p_value = as.numeric(fit_strict$vectors$pvals)
) %>%
  arrange(p_value)

# Factor centroids (if present)
if (!is.null(fit_strict$factors) && length(fit_strict$factors$pvals) > 0) {
  fit_tbl_factor <- tibble(
    variable = names(fit_strict$factors$pvals),
    r2 = as.numeric(fit_strict$factors$r),
    p_value = as.numeric(fit_strict$factors$pvals)
  ) %>%
    arrange(p_value)
} else {
  fit_tbl_factor <- tibble(variable = character(), r2 = numeric(), p_value = numeric())
}

fit_tbl <- bind_rows(
  fit_tbl_numeric %>% mutate(kind = "numeric"),
  fit_tbl_factor %>% mutate(kind = "factor")
) %>%
  arrange(p_value)

fit_tbl
```

```{r strict_driver_model_paired_permanova}
# Next step: paired multivariable model
# Goal: test which strict drivers remain associated with composition
# after controlling for first/last sample and other retained drivers.

# 1) Start from envfit-supported candidates (p <= 0.05)
driver_candidates <- fit_tbl %>%
  filter(p_value <= 0.05, variable %in% vars_driver_strict_final) %>%
  pull(variable)

# 2) Basic collinearity pruning for numeric variables
num_candidates <- driver_candidates[driver_candidates %in% names(meta_strict)[sapply(meta_strict, is.numeric)]]
fac_candidates <- setdiff(driver_candidates, num_candidates)

if (length(num_candidates) >= 2) {
  cor_mat <- meta_strict %>%
    select(all_of(num_candidates)) %>%
    cor(use = "pairwise.complete.obs")

  high_pairs <- which(abs(cor_mat) > 0.80 & abs(cor_mat) < 1, arr.ind = TRUE)
  drop_numeric <- character(0)

  if (nrow(high_pairs) > 0) {
    # Drop the second variable in each high-correlation pair (simple deterministic rule)
    drop_numeric <- unique(colnames(cor_mat)[high_pairs[, "col"]])
  }
} else {
  cor_mat <- matrix(numeric(), nrow = 0, ncol = 0)
  drop_numeric <- character(0)
}

driver_final_diag <- c(setdiff(num_candidates, drop_numeric), fac_candidates)

# 3) Build paired PERMANOVA formula with sample + drivers
driver_formula <- as.formula(
  paste("mat_strict ~ sample +", paste(driver_final_diag, collapse = " + "))
)

perm_drivers_terms <- vegan::adonis2(
  formula = driver_formula,
  data = meta_strict,
  method = "bray",
  permutations = 999,
  strata = meta_strict$plot_number,
  by = "terms"
)

perm_drivers_margin <- vegan::adonis2(
  formula = driver_formula,
  data = meta_strict,
  method = "bray",
  permutations = 999,
  strata = meta_strict$plot_number,
  by = "margin"
)

list(
  driver_candidates = driver_candidates,
  dropped_numeric_due_to_high_correlation = drop_numeric,
  driver_final_diag = driver_final_diag,
  cor_mat_numeric = cor_mat,
  perm_drivers_terms = perm_drivers_terms,
  perm_drivers_margin = perm_drivers_margin
)
```


