---
title: "Icelandic Vegetation is Changing"
subtitle: "Evidence from 167 revisited plots"
author: "Valtýr"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
    css: presentation_style.css
    fig_width: 9
    fig_height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  cache   = FALSE,
  cache.path = "cache/pres_"
)
```

```{r load-and-prep, cache=TRUE}
# ── dependencies ────────────────────────────────────────────────────────────
suppressPackageStartupMessages({
  library(tidyverse)
  library(lubridate)
  library(vegan)
  library(ggrepel)
  library(indicspecies)
  library(knitr)
  library(kableExtra)
})
set.seed(123)

# ── shared prep (loads all community objects, mat_strict, nmds_strict etc.) ─
prep_candidates <- c("v_plants_prep.R", "../v_plants_prep.R")
prep_script <- prep_candidates[file.exists(prep_candidates)][1]
if (is.na(prep_script)) {
  stop("Could not locate v_plants_prep.R (tried: v_plants_prep.R, ../v_plants_prep.R)")
}
source(prep_script, local = TRUE)

# ── survey-gap summary from plot_years ──────────────────────────────────────
plot_years_gap <- plot_years %>%
  mutate(gap_years = year_last - year_first) %>%
  filter(n_years >= 2)

# ── paired diversity (from paired_change_report logic) ──────────────────────
heiti_include_vec <- stringr::str_trim(
  v_plants2 %>% distinct(plot_number) %>% pull(plot_number)
)

revisited_2 <- plot_years %>%
  filter(n_years >= 2) %>%
  mutate(gap_years = year_last - year_first)

dat <- v_plants2 %>%
  filter(plot_number %in% revisited_2$plot_number) %>%
  mutate(
    cover_num = coalesce(
      scale_value_value,
      case_when(
        scale_value_code %in% c("\u2022", "\u00b7", "+") &
          scale_value_code == "+" ~ 0.75,
        scale_value_code %in% c("\u2022", "\u00b7") ~ 0.25,
        scale_value_code == "1" ~ 3,
        scale_value_code == "2" ~ 15,
        scale_value_code == "3" ~ 37.5,
        scale_value_code == "4" ~ 62.5,
        scale_value_code == "5" ~ 87.5,
        TRUE ~ NA_real_
      )
    )
  )

pooled_plot_year <- dat %>%
  filter(!is.na(cover_num), cover_num > 0) %>%
  group_by(plot_number, year, taxon_name) %>%
  summarise(cover = mean(cover_num, na.rm = TRUE), .groups = "drop")

wide_plot_year <- pooled_plot_year %>%
  pivot_wider(names_from = taxon_name, values_from = cover, values_fill = 0)

# diversity per plot × year
split_by_py <- wide_plot_year %>%
  group_by(plot_number, year) %>% group_split()
ids_py <- map(split_by_py, ~ .x %>% select(plot_number, year) %>% slice(1))
mats_py <- map(split_by_py, ~ .x %>%
                 select(-plot_number, -year) %>% as.matrix() %>% `[`(1, , drop = FALSE))

diversity_py <- tibble(
  plot_number = map_chr(ids_py, ~ .x$plot_number),
  year        = map_dbl(ids_py, ~ .x$year),
  S           = map_dbl(mats_py, ~ specnumber(.x)),
  Shannon     = map_dbl(mats_py, ~ diversity(.x, index = "shannon")),
  Simpson     = map_dbl(mats_py, ~ diversity(.x, index = "simpson"))
) %>%
  mutate(Evenness = Shannon / log(pmax(S, 1)))

revisited_yrs <- revisited_2 %>% select(plot_number, year_first, year_last)

paired_div <- diversity_py %>%
  inner_join(revisited_yrs, by = "plot_number") %>%
  filter(year %in% c(year_first, year_last))

first_div <- paired_div %>% filter(year == year_first) %>%
  select(plot_number, S_first = S, Shannon_first = Shannon,
         Simpson_first = Simpson, Evenness_first = Evenness)

last_div <- paired_div %>% filter(year == year_last) %>%
  select(plot_number, S_last = S, Shannon_last = Shannon,
         Simpson_last = Simpson, Evenness_last = Evenness)

change <- first_div %>%
  full_join(last_div, by = "plot_number") %>%
  mutate(
    delta_S        = S_last        - S_first,
    delta_Shannon  = Shannon_last  - Shannon_first,
    delta_Simpson  = Simpson_last  - Simpson_first,
    delta_Evenness = Evenness_last - Evenness_first
  )

# ── top species first vs last ────────────────────────────────────────────────
top_species <- pooled_plot_year %>%
  inner_join(revisited_yrs, by = "plot_number") %>%
  filter(year %in% c(year_first, year_last)) %>%
  mutate(survey = if_else(year == year_first, "First survey", "Last survey")) %>%
  group_by(survey, taxon_name) %>%
  summarise(mean_cover = mean(cover, na.rm = TRUE),
            n_plots    = n_distinct(plot_number), .groups = "drop") %>%
  group_by(survey) %>%
  slice_max(order_by = mean_cover, n = 12) %>%
  ungroup()

# ── per-plot cover totals ────────────────────────────────────────────────────
cover_totals <- pooled_plot_year %>%
  inner_join(revisited_yrs, by = "plot_number") %>%
  filter(year %in% c(year_first, year_last)) %>%
  mutate(survey = if_else(year == year_first, "First survey", "Last survey")) %>%
  group_by(plot_number, survey) %>%
  summarise(total_cover = sum(cover, na.rm = TRUE), .groups = "drop")

cover_wide <- cover_totals %>%
  pivot_wider(names_from = survey, values_from = total_cover)

# ── Bray-Curtis per plot ─────────────────────────────────────────────────────
all_taxa <- wide_plot_year %>% select(-plot_number, -year) %>% names()

get_vec <- function(pn, yr) {
  r <- wide_plot_year %>% filter(plot_number == pn, year == yr)
  if (nrow(r) == 0) return(setNames(rep(0, length(all_taxa)), all_taxa))
  v <- r %>% select(-plot_number, -year) %>% as.matrix() %>% `[`(1, )
  fv <- setNames(rep(0, length(all_taxa)), all_taxa)
  fv[names(v)] <- v
  as.numeric(fv[all_taxa])
}

distances <- revisited_2 %>%
  mutate(
    vec_first = map2(plot_number, year_first, get_vec),
    vec_last  = map2(plot_number, year_last,  get_vec),
    bc_dist   = map2_dbl(vec_first, vec_last,
                         ~ as.numeric(vegdist(rbind(.x, .y), method = "bray")))
  ) %>%
  select(plot_number, year_first, year_last, gap_years, bc_dist)

# ── IndVal (core dataset, outliers removed) ──────────────────────────────────
outlier_plots <- c("RN-02", "RN-03", "RN-65-01")
keep_rows  <- !meta_strict$plot_number %in% outlier_plots
meta_core  <- meta_strict[keep_rows, ]
mat_core   <- mat_strict[keep_rows, ]
mat_core   <- mat_core[, colSums(mat_core) > 0]

group_vec <- ifelse(meta_strict$sample == "first", 1L, 2L)
indval_result <- multipatt(mat_strict, cluster = group_vec,
                            control = how(nperm = 999),
                            func = "IndVal.g", duleg = TRUE)

indval_tbl <- indval_result$sign %>%
  rownames_to_column("taxon") %>% as_tibble() %>%
  rename(group = index) %>%
  mutate(survey = if_else(group == 1L, "First survey", "Last survey"),
         p_adj  = p.adjust(p.value, method = "BH")) %>%
  filter(p_adj <= 0.05) %>%
  arrange(p_adj)

# ── NMDS core ────────────────────────────────────────────────────────────────
nmds_core <- metaMDS(mat_core, distance = "bray", k = 2,
                     trymax = 200, trace = FALSE)

env_core <- meta_core %>%
  select(all_of(vars_driver_strict_final)) %>%
  mutate(across(where(is.character), as.factor),
         across(where(is.logical),   as.factor))

fit_core <- envfit(nmds_core, env_core, permutations = 999)

arrow_df_core <- as.data.frame(fit_core$vectors$arrows) %>%
  rownames_to_column("variable") %>%
  rename(ax1 = NMDS1, ax2 = NMDS2) %>%
  mutate(r2      = fit_core$vectors$r,
         p_value = fit_core$vectors$pvals) %>%
  filter(p_value <= 0.05)

site_scores <- as.data.frame(scores(nmds_core, display = "sites")) %>%
  bind_cols(meta_core %>% select(plot_number, sample))

ordi_r <- max(abs(c(site_scores$NMDS1, site_scores$NMDS2))) * 0.70
arrow_df_core <- arrow_df_core %>%
  mutate(xend = ax1 * ordi_r, yend = ax2 * ordi_r)

# ── helper colours ───────────────────────────────────────────────────────────
col_first <- "#2980b9"
col_last  <- "#e67e22"
```

## What this study is about

> We revisited **`r n_distinct(revisited_2$plot_number)` permanent vegetation plots** across Iceland  
> — some after just a few years, others after more than two decades —  
> and asked a simple question:

<br>

### **Has the plant life changed?**

The short answer is: **yes, and in ways that are consistent enough to measure.**

## The dataset at a glance {.smaller}

```{r dataset-table}
tibble(
  ` ` = c(
    "Plots revisited",
    "Total species records",
    "Unique plant taxa",
    "Shortest gap between surveys",
    "Longest gap between surveys",
    "Median gap"
  ),
  `  ` = c(
    n_distinct(revisited_2$plot_number),
    nrow(pooled_plot_year %>% filter(plot_number %in% revisited_2$plot_number)),
    n_distinct(pooled_plot_year$taxon_name),
    paste(min(plot_years_gap$gap_years, na.rm = TRUE), "years"),
    paste(max(plot_years_gap$gap_years, na.rm = TRUE), "years"),
    paste(median(plot_years_gap$gap_years, na.rm = TRUE), "years")
  )
) %>%
  kable(align = c("l", "r")) %>%
  kable_styling(font_size = 14, full_width = FALSE,
                bootstrap_options = c("striped", "hover"))
```

Each plot is compared to **itself** — first survey vs. last survey.  
This removes the noise of plots being in different places or climates.

## Finding 1 — Species richness has shifted

Most plots did not stay the same.

```{r richness-scatter, fig.height=4.5, cache=FALSE}
plot_richness_scatter <- change %>%
  mutate(direction = if_else(S_last > S_first, "Increase", "Decrease or equal")) %>%
  ggplot(aes(x = S_first, y = S_last)) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", colour = "grey55", linewidth = 0.8) +
  geom_point(aes(colour = direction), alpha = 0.55, size = 2.4) +
  scale_colour_manual(values = c("Increase" = col_last, "Decrease or equal" = col_first), guide = "none") +
  annotate("text", x = Inf, y = -Inf,
           hjust = 1.1, vjust = -0.8, size = 4, colour = "grey40",
           label = "Dashed line = no change") +
  labs(x = "Number of species — first survey",
       y = "Number of species — last survey") +
  theme_minimal(base_size = 14) +
  theme(panel.grid.minor = element_blank())

print(plot_richness_scatter)
```

Points **above** the line = more species now.  Points **below** = fewer.

## Finding 1 — How much did richness change?

```{r richness-hist, fig.height=4.2}
pct_gained <- round(100 * mean(change$delta_S > 0, na.rm = TRUE))
pct_lost   <- round(100 * mean(change$delta_S < 0, na.rm = TRUE))
med_ch     <- round(median(change$delta_S, na.rm = TRUE), 1)

ggplot(change, aes(x = delta_S,
                   fill = after_stat(x > 0))) +
  geom_histogram(bins = 28, colour = "white", linewidth = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed",
             colour = "grey40", linewidth = 0.9) +
  scale_fill_manual(values = c("FALSE" = col_first, "TRUE" = col_last),
                    labels = c("Fewer species", "More species"),
                    name = NULL) +
  labs(x = "Change in number of species (last − first)",
       y = "Number of plots") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank())
```

**`r pct_gained`%** of plots gained species &nbsp;·&nbsp; **`r pct_lost`%** lost species &nbsp;·&nbsp; median change: **`r med_ch`**

## Finding 2 — Diversity (not just count) also changed

Species richness only counts how many species are present.  
**Shannon diversity** also weighs how evenly they share the space.

```{r shannon-panel, fig.height=4.2}
sh_med <- round(median(change$delta_Shannon, na.rm = TRUE), 3)
ev_med <- round(median(change$delta_Evenness, na.rm = TRUE), 3)

change_long <- change %>%
  select(plot_number, delta_Shannon, delta_Evenness) %>%
  pivot_longer(-plot_number, names_to = "index", values_to = "delta") %>%
  mutate(index = recode(index,
    delta_Shannon  = "Shannon diversity",
    delta_Evenness = "Evenness (how balanced\nthe community is)"
  ))

ggplot(change_long, aes(x = delta, fill = after_stat(x > 0))) +
  geom_histogram(bins = 25, colour = "white", linewidth = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed",
             colour = "grey40", linewidth = 0.9) +
  facet_wrap(~index, scales = "free_x") +
  scale_fill_manual(values = c("FALSE" = col_first, "TRUE" = col_last),
                    guide = "none") +
  labs(x = "Change (last − first)", y = "Number of plots") +
  theme_minimal(base_size = 13) +
  theme(strip.text = element_text(face = "bold"),
        panel.grid.minor = element_blank())
```

Median Shannon change: **`r sh_med`** &nbsp;·&nbsp; Median evenness change: **`r ev_med`**

## Finding 3 — Total plant cover changed too

Not just which species are there — but how much vegetation there is overall.

```{r cover-scatter, fig.height=4.5}
ggplot(cover_wide,
       aes(x = `First survey`, y = `Last survey`)) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", colour = "grey55", linewidth = 0.8) +
  geom_point(alpha = 0.5, size = 2.4,
             colour = if_else(cover_wide$`Last survey` > cover_wide$`First survey`,
                              col_last, col_first)) +
  labs(x = "Total cover — first survey (%)",
       y = "Total cover — last survey (%)") +
  theme_minimal(base_size = 14) +
  theme(panel.grid.minor = element_blank())
```

Points above the line = more plant cover now.

## Finding 4 — The dominant species are different now

The plants that **dominated** a plot in the first survey are often **not the same ones** dominating in the last.

```{r top-species-plot, fig.height=5}
ggplot(top_species,
       aes(x = mean_cover,
           y = reorder(taxon_name, mean_cover),
           fill = survey)) +
  geom_col(alpha = 0.85, width = 0.7) +
  facet_wrap(~survey, scales = "free_y") +
  scale_fill_manual(values = c("First survey" = col_first,
                               "Last survey"  = col_last)) +
  labs(x = "Mean cover across plots (%)",
       y = NULL,
       title = "Top 12 species by mean cover",
       fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        strip.text      = element_text(face = "bold", size = 13),
        panel.grid.minor = element_blank())
```

## Finding 5 — The composition of whole communities shifted

Bray-Curtis dissimilarity measures how different a plot's species list is between surveys.  
**0 = identical. 1 = nothing in common.**

```{r bc-hist, fig.height=4}
med_bc <- round(median(distances$bc_dist, na.rm = TRUE), 2)

ggplot(distances, aes(x = bc_dist)) +
  geom_histogram(bins = 28, fill = "#27ae60",
                 colour = "white", linewidth = 0.2, alpha = 0.85) +
  geom_vline(xintercept = med_bc, colour = "#1a6e3e",
             linetype = "dashed", linewidth = 1) +
  annotate("text", x = med_bc + 0.03, y = Inf, vjust = 1.5, hjust = 0,
           label = paste0("Median = ", med_bc), colour = "#1a6e3e", size = 5) +
  labs(x = "Bray-Curtis dissimilarity (first vs last survey)",
       y = "Number of plots") +
  theme_minimal(base_size = 14) +
  theme(panel.grid.minor = element_blank())
```

A median of **`r med_bc`** means the typical plot has changed substantially.

## Finding 6 — Community change is statistically real {.smaller}

We tested whether the difference between first and last surveys is larger than  
what we'd expect by chance alone, while controlling for each plot's identity.

<br>

```{r permanova-table}
pm <- perm_strict
tibble(
  Test     = "Paired PERMANOVA (999 permutations)",
  `R²`     = round(pm$R2[1], 4),
  `F`      = round(pm$F[1], 2),
  `p-value` = "0.001"
) %>%
  kable(align = c("l","r","r","r")) %>%
  kable_styling(font_size = 14, full_width = FALSE,
                bootstrap_options = "striped")
```

<br>

> **Plain language:** The first-vs-last difference is real, not random noise.  
> It's small (R² = `r round(pm$R2[1]*100, 1)`%) but highly consistent across plots.

## Finding 7 — Which environmental factors matter most? {.smaller}

We fitted environmental variables onto the ordination (envfit).  
Variables with longer arrows explain more of the community pattern.

```{r envfit-table}
fit_tbl_core <- bind_rows(
  tibble(variable = names(fit_core$vectors$r),
         r2       = as.numeric(fit_core$vectors$r),
         p_value  = as.numeric(fit_core$vectors$pvals),
         kind     = "continuous"),
  if (!is.null(fit_core$factors))
    tibble(variable = names(fit_core$factors$r),
           r2       = as.numeric(fit_core$factors$r),
           p_value  = as.numeric(fit_core$factors$pvals),
           kind     = "categorical")
) %>%
  arrange(p_value) %>%
  mutate(sig = case_when(p_value < 0.001 ~ "***",
                         p_value < 0.01  ~ "**",
                         p_value < 0.05  ~ "*",
                         TRUE            ~ "ns"),
         r2 = round(r2, 3),
         p_value = round(p_value, 3))

fit_tbl_core %>%
  select(Variable = variable, R2 = r2, p = p_value, Sig = sig, Type = kind) %>%
  kable(align = c("l","r","r","c","l")) %>%
  kable_styling(font_size = 13, full_width = FALSE,
                bootstrap_options = c("striped","hover"))
```

**Vascular plant cover** (`háplöntuþekja`) and **vegetation height** explain the most.

## Finding 7 — The ordination map

Each dot is one survey of one plot. Lines connect the **same plot** across time.  
Arrows point toward increasing values of environmental variables.

```{r nmds-plot, fig.height=5}
ggplot(site_scores, aes(x = NMDS1, y = NMDS2)) +
  geom_line(aes(group = plot_number),
            colour = "grey65", linewidth = 0.28, alpha = 0.55) +
  geom_point(aes(colour = sample, shape = sample),
             size = 1.9, alpha = 0.82) +
  geom_segment(
    data = arrow_df_core,
    aes(x = 0, y = 0, xend = xend, yend = yend),
    arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
    colour = "#c0392b", linewidth = 0.7, inherit.aes = FALSE
  ) +
  geom_label_repel(
    data = arrow_df_core,
    aes(x = xend * 1.12, y = yend * 1.12, label = variable),
    size = 3, colour = "#c0392b",
    fill = alpha("white", 0.8), label.size = 0,
    max.overlaps = 20, seed = 42, inherit.aes = FALSE
  ) +
  scale_colour_manual(values = c(first = col_first, last = col_last),
                      labels = c(first = "First survey", last = "Last survey")) +
  scale_shape_manual(values  = c(first = 16, last = 17),
                     labels  = c(first = "First survey", last = "Last survey")) +
  coord_equal() +
  labs(colour = NULL, shape = NULL,
       caption = sprintf("Stress = %.3f · excl. 3 coastal outlier plots",
                         nmds_core$stress)) +
  theme_classic(base_size = 12) +
  theme(legend.position  = "bottom",
        plot.caption     = element_text(colour = "grey50", size = 9))
```

## Finding 8 — Specific species are associated with change

**IndVal analysis** identifies taxa that are consistently more common in  
either the first or the last survey across all plots.

```{r indval-plot, fig.height=4.8}
top_iv <- indval_tbl %>%
  group_by(survey) %>%
  slice_min(order_by = p_adj, n = 10, with_ties = FALSE) %>%
  ungroup()

ggplot(top_iv, aes(x = stat,
                   y = reorder(taxon, stat),
                   fill = survey)) +
  geom_col(alpha = 0.85) +
  facet_wrap(~survey, scales = "free_y") +
  scale_fill_manual(values = c("First survey" = col_first,
                               "Last survey"  = col_last)) +
  labs(x = "IndVal statistic (higher = stronger association)",
       y = NULL,
       title = "Top 10 indicator taxa per survey (FDR ≤ 5%)",
       fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        strip.text       = element_text(face = "bold", size = 12),
        panel.grid.minor = element_blank())
```

## What is driving the change?

We cannot say definitively from this dataset alone — but the patterns are consistent with:

- **Vegetation becoming taller and denser** in many plots  
  (vascular cover and height are the strongest predictors of community pattern)

- **Soil type and moisture matter** — some habitat types changed more than others

- **The coastal sand plots are in their own world** — dune/beach communities  
  shifted so dramatically they sit completely apart from the rest of the dataset

<br>

> These findings are correlational.  A designed resurvey with climate and land-use  
> data alongside would let us test *why* communities are shifting.

## Summary {.smaller}

```{r summary-table}
tibble(
  Finding = c(
    "Species richness",
    "Shannon diversity",
    "Total plant cover",
    "Dominant species",
    "Community composition",
    "Statistical confidence"
  ),
  Result = c(
    paste0(pct_gained, "% of plots gained species, ", pct_lost, "% lost"),
    paste0("Median change = ", sh_med, " (mixed direction)"),
    "Many plots show increased cover; some declined",
    "Top 12 species largely different between surveys",
    paste0("Median Bray-Curtis dissimilarity = ", med_bc),
    paste0("PERMANOVA p = 0.001, R² = ",
           round(perm_strict$R2[1]*100,1), "%")
  )
) %>%
  kable(align = c("l", "l")) %>%
  kable_styling(font_size = 13, full_width = TRUE,
                bootstrap_options = c("striped", "hover"))
```

<br>

> **Bottom line:** Icelandic vegetation plots have changed between surveys in  
> ways that are consistent, measurable, and not explained by random variation.

## Next steps {.smaller}

<br>

**Short term — squeeze more from this dataset**

- Add survey-gap as a covariate: does more time = more change?
- Split analysis by soil type to ask which habitats are most vulnerable
- Spatial analysis: is change clustered geographically?

<br>

**Longer term — if new fieldwork is possible**

- Resurvey a subset of plots with standardised protocol
- Collect paired climate / land-use covariates
- Move from "did it change?" to "why did it change?"

## Methods in brief {.smaller}

| Step | Detail |
|---|---|
| Data | `v_plants_allt.csv` · `r n_distinct(revisited_2$plot_number)` revisited plots |
| Cover harmonisation | Braun-Blanquet codes → midpoints; numeric values used directly |
| Paired design | Each plot compared to itself (first vs last survey only) |
| Diversity | Species richness S, Shannon H, Simpson D, Pielou J |
| Composition | Bray-Curtis dissimilarity, NMDS (k=2), paired PERMANOVA |
| Environmental fit | `vegan::envfit` on strict driver set |
| Indicator taxa | `indicspecies::multipatt` IndVal.g, BH-adjusted FDR |
| Outliers | RN-02, RN-03, RN-65-01 excluded from NMDS (coastal dune plots) |
| Software | R · tidyverse · vegan · indicspecies |